### 2022-07-04 notes/variant_origin.tex
reorg notes

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\mat}[1]{{\bf #1}}
\newcommand{\vecb}[1]{{\bf #1}}
\newcommand{\abet}{\mathcal{A}}
\newcommand{\eqp}{p}
\newcommand{\LH}{\mathcal{L}}

\begin{document}
\title{When did a variant start spreading?}
\author{Richard Neher}
\date{\today}
\maketitle

Once a variant is at high frequency, it spreads mostly deterministically, but its early, often unobserved, spread is very stochastic.
Lets consider a discrete generation model and a certain observation probability $\epsilon$.
If we observed $m_1, m_2, \ldots, m_T$ sequences of a variant in successive generations, we'd like to infer the true prevalence $n_t$.
If the outbreak started with $n_\tau=1$ cases at seeding time $t=\tau$, we can write the likelihood of observing the case series as
\begin{equation}
    \begin{split}
        P(\{m_t\} | \tau) & = \prod_{t=\tau+1}^T\sum_{n_t=1}^\infty p(m_t|n_t,\epsilon) g_{n_t,n_{t-1}} \\
    & =  \sum_{n_T} p(m_{k}|n_{T},\epsilon) \sum_{n_{T-1}} g_{n_{T},n_{T-1}}p(m_{T-1}|n_{T-1},\epsilon)\cdots p(m_{t}|n_{t},\epsilon)\sum_{n_{t-1}}  g_{n_{t},n_{t-1}}\cdots\sum_{n_{\tau + 1}=1}^\infty g_{n_{\tau+2},n_{\tau+1}} p(m_{\tau+1}|n_{\tau+1},\epsilon) g_{n_{\tau+1},1}
\end{split}
\end{equation}
In addition, we can sum over possible seeding times $\tau$ where evidently $\tau$ can not be later than the first sampled case $t_0$.
\begin{equation}
    P(\{m_t\}) =\sum_{\tau}^{t_0} p(\tau)P(\{m_t\} | \tau)
\end{equation}

For any given $t$, we can write $P(\{m_t\} | \tau)$ as
\begin{equation}
    P(\{m_t\} | \tau) = \sum_{n=1}^\infty R(t)_{n} Q(t,\tau)_n  = \sum_{\tau}^{t_0} p(\tau) R(\tau)_1
\end{equation}
where $R(t)_n$ is given by
\begin{equation}
    R(t)_n = \sum_{n_T} p(m_{T}|n_{T},\epsilon) \sum_{n_{T-1}} g_{n_{T},n_{T-1}}p(m_{T-1}|n_{T-1},\epsilon)\cdots p(m_{t}|n,\epsilon)
\end{equation}
and
\begin{equation}
    Q(t, \tau)_n = \sum_{n_{t-1}}  g_{n,n_{t-1}}\cdots\sum_{n_{\tau + 1}=1}^\infty g_{n_{\tau+2},n_{\tau+1}} p(m_{\tau+1}|n_{\tau+1},\epsilon) g_{n_{\tau+1},1}
\end{equation}

These quantities can be calculated recursively:
\begin{equation}
    R(t-1)_n = \sum_m R(t)_m g_{m,n} \quad \mathrm{with} \quad R(T)_n = p(m_{T}|n,\epsilon)
\end{equation}
\begin{equation}
    Q(t+1,\tau)_n = \sum_m g_{n,m}Q(t,\tau)_m  \quad \mathrm{with} \quad Q(\tau,\tau)_n = \delta_{n,1}
\end{equation}

We can also calculate the distribution of $n$ at a specific time $t^*$
\begin{equation}
    P(n_{t^*}) = \frac{1}{P(\{m_t\})}\sum_{\tau}^{t_0} R(t^*)_n p(\tau) Q(t^*,\tau)_n
\end{equation}

The transmission matrix $g_{nm}$ needs to account for spread with rate $\rho$ and transmission noise.
It is common to parameterize the transmission by a Beta-binomial distribution with mean $m\rho$ and dispersion parameter $k$.
\begin{equation}
    g_{nm} = {n + mr - 1 \choose n} (1-p)^{m k} p^n
\end{equation}
where $p=k/(k+m\rho)$ to match the mean $\rho m$.

\subsection*{Multiple clones}
If an additional daughter variant emerged from the founder variant, that puts additional constraints on the trajectory of the founder.
The trajectory of the daughter is fed by mutations from the founder clone.
A slightly tricky part is what the mutation rate should be.
For any particular mutation, it is small (somewhere at 0.0005/year), while genome wide it is more like 15/year or 1-2 per month.
When selecting variants for further inclusion in the multi-genotype model, we are effectively picking those that are at the highest frequency and thus pick one or several arbitrary sites.
The relevant mutation rate is thus $\mu L/k$ where $k$ is the number of genotypes considered.

The likelihood of the sampling a given trajectory of daughter cases is $R(t)_1$ as calculated above.
The probability of seeding at least one variant in generation $t$ is $1-e^{-n_t\mu L}$ where $n_t$ is the number of parent genotypes at time $t$.
We again need to sum over different seeding times

\begin{equation}
    P(\{m_t\}, \{c_t\}) = \sum_{\tau_0}^{t_0}\sum_{\tau_d=\tau_0+1}^{t_d} R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
\end{equation}
where $\tau_d$ is the seeding time of the daughter clone, $\tau_0$ is the seeding of the parent clone, and $\{m_t\}, \{c_t\}$ are the parent and daughter cases count trajectories.
The probability of being seeded in generation $\tau_0$ is then
\begin{equation}
    P(\tau_0) \sim \sum_{\tau_d=\tau_0+1}^{t_d} R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
\end{equation}

\end{document}

### 2022-07-04 scripts/fit_outbreak.py
add seeding time distributions

@@ -137,9 +137,11 @@ if __name__=="__main__":
 
     logR_daughter = back_trace(data_daughter)
     logR = back_trace(data)
+    logQ = fwd_trace(data)
 
     daughter_seeding = {t:lR[lR[:,0]==1,1][0] for t,lR in logR_daughter.items() if lR[0,0]<=1}
     logR_wdaughter = {}
+    logQ_wdaughter = {}
     for seeding_time, lR_daughter in daughter_seeding.items():
         lR_wd = {}
         for t,lR in logR.items():
@@ -151,6 +153,20 @@ if __name__=="__main__":
                                           nmin=row.cases, nmax=3*(row.cases+3)/eps,
                                           daughters=(t+1==seeding_time), daughter_rate=0.02)
         logR_wdaughter[seeding_time] = lR_wd
+
+        lQ_wd = {}
+        for t,lQ in logQ.items():
+            if t<seeding_time:
+                lQ_wd[t] = lQ
+            else:
+                cases = row.cases if t else 0
+                next_row = data.loc[t]
+                row = data.loc[t-1] if t else next_row
+                prev = lQ_wd.get(t-1, None)
+                lQ_wd[t] = fwd_propagate(cases, prev, row.rho, row.k, row.eps, row.mu,
+                                          nmin=next_row.cases, nmax=3*(next_row.cases+3)/eps,
+                                          daughters=(t==seeding_time), daughter_rate=0.02)
+        logQ_wdaughter[seeding_time] = lQ_wd
 
     print(logR[0][0,1])
     for t in logR_wdaughter:
@@ -165,7 +181,6 @@ if __name__=="__main__":
     plt.xscale('log')
     plt.legend()
 
-    logQ = fwd_trace(data)
     plt.figure()
     for t in logQ:
         plt.plot(logQ[t][:,0]+0.5, logQ[t][:,1] - logQ[t][:,1].max(), label=f'cases={data.at[t,"cases"]}')
@@ -189,10 +204,9 @@ if __name__=="__main__":
     peak_val = -np.inf
     for i in time_points:
         tmp = {}
-        tm1 = data.iloc[i-1].name
         t = data.iloc[i].name
         for seeding_time, lR_daughter in daughter_seeding.items():
-            tmp[seeding_time] = marginal_distribution(logQ[tm1], logR_wdaughter[seeding_time][t], data.loc[t])
[…]

### 2022-07-06 notes/variant_origin.tex
multi-variant scribbles

@@
 \begin{equation}
-    R(t-1)_n = \sum_m R(t)_m g_{m,n} \quad \mathrm{with} \quad R(T)_n = p(m_{T}|n,\epsilon)
+    R(t-1)_n = \sum_k R(t)_k g_{k,n} p(m_{t-1}|n,\epsilon) \quad \mathrm{with} \quad R(T)_n = p(m_{T}|n,\epsilon)
 \end{equation}
 \begin{equation}
-    Q(t+1,\tau)_n = \sum_m g_{n,m}Q(t,\tau)_m  \quad \mathrm{with} \quad Q(\tau,\tau)_n = \delta_{n,1}
+    Q(t+1,\tau)_n = \sum_k g_{n,k} p(m_t|k,\epsilon) Q(t,\tau)_k  \quad \mathrm{with} \quad Q(\tau,\tau)_n = \delta_{n,1}
 \end{equation}
@@
 \begin{equation}
-    P(\tau_0) \sim \sum_{\tau_d=\tau_0+1}^{t_d} R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
+    P(\tau_0) \sim \sum_{\tau_d=\tau_0+1}^{t_d} \sum_n R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
+\end{equation}
+
+\subsection*{Mutation categories}
+Instead of tracking clones, we might want to track genotypes with 0,1,2,... mutations.
+In this case, mutation happen independently and there is a constant flux of cases from lower to higher mutation classes.
+The last "catch-all" class grows with a rate that is $\mu L$ higher than the rest.
+
+In this case, we'd start with the most mutated class and would need to extend the propagator to include the mutational influx.
+The trajectory of mutant class $k$ is conditional on the mutant class $k-1$.
+\begin{equation}
+    \begin{split}
+        P(\{m^0_t\}, \{m^1_t\}, \{m^2_t\} \ldots | \tau) & = \prod_{t=\tau+1}^T\sum_{n^0_t=1}\sum_{n^1_t=0}^\infty\sum_{n^2_t=0}^\infty p(m^0_t|n^0_t,\epsilon)g_{n^0_t,n^0_{t-1}} p(m^1_t|n^1_t,\epsilon) g_{n^1_t,n^1_{t-1}, n^0_{t-1}}p(m^2_t|n^2_t,\epsilon)g_{n^2_t,n^2_{t-1}, n^1_{t-1}} \\
+         & = \sum_{n^0_\tau=1}\sum_{n^1_t=0}^\infty\sum_{n^2_t=0}^\infty p(m^0_\tau|n^0_\tau,\epsilon) p(m^1_\tau|n^1_\tau,\epsilon) p(m^2_\tau|n^2_\tau,\epsilon) \\
+          &   \ldots \\
+          & \sum_{n^0_{T-1}=1}  p(m^0_{T-1}|n^0_{T-1},\epsilon) g_{n^0_{T-1},n^0_{T-2}} \sum_{n^1_{T-1}=0}^\infty p(m^1_{T-1}|n^1_{T-1},\epsilon) g_{n^1_{T-1},n^1_{T-2}, n^0_{T-2}}  \sum_{n^2_{T-1}=0}^\infty p(m^2_{T-1}|n^2_{T-1},\epsilon) g_{n^2_{T-1},n^2_{T-2}, n^1_{T-2}} \\
+          &   \sum_{n^0_T=1}p(m^0_T|n^0_T,\epsilon)  g_{n^0_T,n^0_{T-1}} \sum_{n^1_T=0}^\infty p(m^1_T|n^1_T,\epsilon)g_{n^1_T,n^1_{T-1}, n^0_{T-1}}\sum_{n^2_T=0}^\infty p(m^2_T|n^2_T,\epsilon)  g_{n^2_T,n^2_{T-1}, n^1_{T-1}}
+    \end{split}
+\end{equation}
+The quantity $g_{n^2_t,n^2_{t-1}, n^1_{t-1}}$ describes how many double mutated cases are present in generation $t$ if there where $n^2_{t-1}$ double mutated and $n^1_{t-1}$ single mutated individuals previously.
+The mean number of progeny is then $\rho e^{-\mu L} (n^2_{t-1} + \mu L n^1_{t-1})$.
+This problem can again be solved recursively by defining $R^k(t)_{n^k,n^{k-1}}$ of mutant category $k$ to be
+\begin{equation}
+    R(t-1)_{n^k,n^{k-1}} = \sum_{l^k} R(t)_{l^k,l^{k-1}} g_{k,n} p(m_{t-1}|n,\epsilon) \quad \mathrm{with} \quad R(T)_n = p(m_{T}|n,\epsilon)
 \end{equation}

### 2022-07-06 scripts/fit_outbreak.py
generate clone count data

@@ -9,7 +9,7 @@ LogTINY = np.log(TINY)
 birth_rate = 1e-6
 logCutoff = 50
 
-def log_propagator(n, m, rho, k, mu):
+def log_propagator(n, m, rho, k):
     '''
     nvec = np.arange(100)
     k,rho = 0.1, 1.1
@@ -17,107 +17,107 @@ def log_propagator(n, m, rho, k, mu):
     avg = np.sum(pn*nvec) # should be rho*5
     '''
     p = k/(k+rho)
-    if m:
-        return nbinom.logpmf(n,m*k,p)
-    else:
-        tmp = LogTINY*np.ones_like(n)
-        tmp[n==0] = -mu
-        tmp[n==1] = np.log(mu)
-        return tmp
-
-def log_propagator_mvec(n, m, rho, k, mu):
-    p = k/(k+rho)
-    res = LogTINY*np.ones_like(m)
-    res[m>0] = nbinom.logpmf(n,m[m>0]*k,p)
-    if n==0:
-        res[m==0] = -mu
-    elif n==1:
-        res[m==0] = np.log(mu)
-
-    return res
+    return nbinom.logpmf(n,m*k,p)
 
 def propagator(n, m, rho, k):
     return np.exp(log_propagator(n,m,rho,k))
 
 
-def log_sampling_prob(n_samples, n_cases, eps):
+def log_sampling_prob(n_samples, eps, n_cases):
     mean_density = eps*n_cases
     return poisson.logpmf(n_samples, mean_density)
 
 
-def back_propagate(samples_t, logR_tp1, rho, k, eps, mu, nmin=0, nmax=1000, daughters=False, daughter_rate=0):
-    nvals = np.arange(nmin, nmax)
+def back_propagate_R(sampling_t, logR_tp1, rho, k, nmin=1, nmax=1000):
+    nvals = np.arange(max(1,int(nmin)), int(nmax))
     if logR_tp1 is None:
-        logR_t = np.array([nvals, log_sampling_prob(samples_t, nvals, eps)]).T
+        logR_t = np.array([nvals, np.sum([log_sampling_prob(s_t[0], s_t[1], nvals) for s_t in sampling_t], axis=0)])
     else:
-        peak = logR_tp1[:,1].max()
-        logR_t = np.array([(n_t, log_sampling_prob(samples_t,n_t, eps)
-                       + peak + np.log(np.sum(np.exp(logR_tp1[:,1] - peak + log_propagator(logR_tp1[:,0], n_t, rho, k, mu)))))
[…]

### 2022-07-06 scripts/get_genotype_counts.py
add scripts to generate genotype counts

@@ -1,0 +1,58 @@
+import pandas as pd
+import argparse,json
+from datetime import datetime
+import numpy as np
+from root_to_tip import filter_and_transform
+
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+
+    parser.add_argument('--metadata', nargs='+', type=str, required=True, help="input data")
+    parser.add_argument('--clade-gts', type=str, required=True, help="input data")
+    parser.add_argument('--clade', type=str, required=True, help="input data")
+    parser.add_argument('--min-date', type=float, help="input data")
+    parser.add_argument('--max-date', type=float, help="input data")
+    parser.add_argument('--bin-size', type=float, help="input data")
+    parser.add_argument('--output-json', type=str, help="rate file")
+    args = parser.parse_args()
+
+    with open(args.clade_gts) as fh:
+        clade_gt = json.load(fh)[args.clade]
+
+    d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
+    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date, completeness=0)
+    filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())
+    print("clade", args.clade, "done filtering")
+    bins = np.arange(np.min(filtered_data.day),np.max(filtered_data.day), args.bin_size)
+    all_sequences = np.histogram(filtered_data.day, bins=bins)[0]
+    cumulative_sum = np.zeros_like(bins[:-1])
+
+    mutations = {}
+    nmax = 5
+    for n in range(nmax):
+        ind = filtered_data["divergence"]==n
+        mutations[n] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
+        cumulative_sum += mutations[n]
+    mutations[f'{nmax}+'] = all_sequences - cumulative_sum
+
+    print("clade", args.clade, "done mutations")
+    intra_geno = filtered_data["intra_substitutions_str"].value_counts()
+    genotypes = {}
+    for x,i in intra_geno.items():
+        if i<10 or i<len(filtered_data)/10000:
+            break
+        nmuts = len(x.split(',')) if x else 0
+        ind = filtered_data["intra_substitutions_str"]==x
[…]

### 2022-07-06 scripts/plot_genotype_counts.py
add plot scritp

@@ -1,0 +1,41 @@
+import argparse,json
+from datetime import datetime
+import matplotlib as mpl
+mpl.rcParams['axes.formatter.useoffset'] = False
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+
+    parser.add_argument('--counts', type=str, required=True, help="input data")
+    parser.add_argument('--output-plot', type=str, help="figure file")
+    args = parser.parse_args()
+
+    with open(args.counts) as fh:
+        counts = json.load(fh)
+
+
+    dates = [datetime.fromordinal(x) for x in counts['bins'][:-1]]
+    fig, axs = plt.subplots(2,1, figsize = (8,15), sharex=True)
+
+    axs[0].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
+    for m in counts['mutations']:
+        axs[0].plot(dates, counts['mutations'][m], '-o', label=f'{m} mutations')
+
+    axs[0].set_yscale('log')
+    axs[0].legend()
+
+    axs[1].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
+    for m in counts['genotypes']:
+        axs[1].plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")
+
+    axs[1].set_yscale('log')
+    axs[1].legend()
+
+    plt.savefig(args.output_plot)

### 2022-07-07 notes/variant_origin.tex
generate clone count data
add ideas to mutation categories, fix small typos

@@
 \newcommand{\LH}{\mathcal{L}}
-
 \begin{document}
@@
 \maketitle
-
 Once a variant is at high frequency, it spreads mostly deterministically, but its early, often unobserved, spread is very stochastic.
@@
 \begin{equation}
-    P(\{m_t\}) =\sum_{\tau}^{t_0} p(\tau)P(\{m_t\} | \tau)
+    P(\{m_t\}) =\sum_{\tau}^{t_0} p(\tau)P(\{m_t\} | \tau) = \sum_{\tau}^{t_0} p(\tau) R(\tau)_1
 \end{equation}
-
 For any given $t$, we can write $P(\{m_t\} | \tau)$ as
 \begin{equation}
-    P(\{m_t\} | \tau) = \sum_{n=1}^\infty R(t)_{n} Q(t,\tau)_n  = \sum_{\tau}^{t_0} p(\tau) R(\tau)_1
+    P(\{m_t\} | \tau) = \sum_{n=1}^\infty R(t)_{n} Q(t,\tau)_n = R(\tau)_1
 \end{equation}
@@
 \end{equation}
-
 These quantities can be calculated recursively:
@@
 \end{equation}
-
 We can also calculate the distribution of $n$ at a specific time $t^*$
@@
 \end{equation}
-
 The transmission matrix $g_{nm}$ needs to account for spread with rate $\rho$ and transmission noise.
@@
 where $p=k/(k+m\rho)$ to match the mean $\rho m$.
-
 \subsection*{Multiple clones}
@@
 The relevant mutation rate is thus $\mu L/k$ where $k$ is the number of genotypes considered.
-
 The likelihood of the sampling a given trajectory of daughter cases is $R(t)_1$ as calculated above.
@@
 We again need to sum over different seeding times
-
 \begin{equation}
-    P(\{m_t\}, \{c_t\}) = \sum_{\tau_0}^{t_0}\sum_{\tau_d=\tau_0+1}^{t_d} R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
+    P(\{m_t\}, \{c_t\}) = \sum_{\tau_0}^{t_0}\sum_{\tau_d=\tau_0+1}^{t_d} \sum_n  R(\tau_d)_n (1-e^{-n\mu L}) Q(\tau_d,\tau)_n R_d(\tau_d)
 \end{equation}
@@
 The last "catch-all" class grows with a rate that is $\mu L$ higher than the rest.
-
 In this case, we'd start with the most mutated class and would need to extend the propagator to include the mutational influx.
@@
 
+\subsection*{Mutation categories - other approach}
+Instead of tracking clones, we might want to track genotypes with 0,1,2,... mutations.
+In this case, mutation happen independently and there is a constant flux of cases from lower to higher mutation classes.
+The last "catch-all" class grows with a rate that is $\mu L$ higher than the rest.
+In this case, we'd start with the most mutated class and would need to extend the propagator to include the mutational influx.
+The trajectory of mutant class $k$ is conditional on the mutant class $k-1$.
+
+Define the $\textbf{m}_t = (m_0, ..., m_k)_t$ vector as the sampled number of genotypes with 0,1,..., k mutations at time $t$, then  $\textbf{n}_t = (n_0, ..., n_k)_t$ is the hidden true population size at each of these time points.
+Now the probability of the ${\textbf{m}_t}_t$ vector with seeding time of the first genotype at time $\tau$ (here the hidden vector is $\textbf{n}_\tau = (1, 0, ..., 0)$) can be expressed as
+
+\begin{equation}
+        P(\{\textbf{m}_t\}_t| \tau) & = \prod_{t=\tau+1}^T\sum_{\textbf{n}_t=(1,0,... ,0)}^\infty p(\textbf{m}_t|\textbf{n}_t,\epsilon) g_{\textbf{n}_t,\textbf{n}_{t-1}} \\
+\end{equation}
+
+Now the sampling of each $m_{i,t}$ is independent, i.e. 
+\begin{equation}
+    p(\textbf{m}_t|\textbf{n}_t,\epsilon) = \prod_0^k p(m_{k,t}|n_{k,t},\epsilon)
+\end{equation}
+
+The transition matrix is slightly more complex:
+\begin{equation}
+g_{\textbf{n}_t,\textbf{n}_{t-1}} = g_{n_{0,t},n_{0,t-1}} \prod_1^k g'_{n_{i,t},n_{i,t-1}, n_{i-1,t-1}}
+\end{equation}
+
+This is due to the fact that populations of genotypes with 1 or more mutations are not only dependent on the size of that population in the previous time step but also on the size of the population with 1 less mutation (we assume that the likelihood of more than one mutation in a generation step is negligibly small).
+
+\begin{equation}
+g'_{n_{i,t},n_{i,t-1}, n_{i-1,t-1}} = g_{n_{i,t},n_{i,t-1}} + \mu L n_{i-1, t-1}
+\end{equation}
+
+The mean number of progeny with two mutations  at time $t$ is then $n_{2,t} = \rho e^{-\mu L} (n_{2,t-1} + \mu L n_{1, t-1})$.
+
+However, this also means that the probability is not comprised of linearly independent sub-probabilities for each genotype's observed samples.
+Thus, in order to calculate the maximum likelihood seeding time $R$ and $Q$ must be calculated by iterating over a k dimensional subspace.
+
+\begin{equation}
+    R(t-1)_{\textbf{n}} = \sum_{\textbf{n}'} R(t)_{\textbf{n}'} g_{\textbf{n}',\textbf{n}} p(\textbf{m}_{t-1}|\textbf{n},\epsilon) \quad \mathrm{with} \quad R(T)_\textbf{n} = p(\textbf{m}_{T}|\textbf{n},\epsilon)
+\end{equation}
+
+\begin{equation}
+    Q(t+1,\tau)_{\textbf{n}} = \sum_{\textbf{n}'} g_{n,\textbf{n}'} p(\textbf{m}_t|\textbf{n}',\epsilon) Q(t,\tau)_{\textbf{n}}'  \quad \mathrm{with} \quad Q(\tau,\tau)_{\textbf{n}} = \delta_{{\textbf{n}},(1,0,...
+0)}
+\end{equation}
+
 \end{document}

### 2022-07-23 scripts/fit_origin.py
add composite variants

@@ -1,0 +1,49 @@
+import argparse,json
+import numpy as np
+from datetime import datetime
+import matplotlib as mpl
+from fit_outbreak import optimize_single_trajectory
+mpl.rcParams['axes.formatter.useoffset'] = False
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+
+    parser.add_argument('--counts', type=str, required=True, help="input data")
+    parser.add_argument('--output-plot', type=str, help="figure file")
+    args = parser.parse_args()
+
+    with open(args.counts) as fh:
+        counts = json.load(fh)
+
+    obs_dates = [datetime.fromordinal(x) for x in counts['bins']]
+
+    fig = plt.figure()
+    cutoff = 6
+    eps=0.02
+    print("fit all samples")
+    fit = optimize_single_trajectory(counts['bins'][:cutoff], counts['all_samples'][:cutoff], sampling=eps)
+    dates = [datetime.fromordinal(x) for x in fit['timepoints']]
+    peak = max(fit['logLH'])
+    plt.plot(dates, np.exp(fit['logLH'] - peak))
+
+    print("fit founder")
+    fit2 = optimize_single_trajectory(counts['bins'][:cutoff], counts['mutations']['0'][:cutoff], sampling=eps)
+    dates = [datetime.fromordinal(x) for x in fit2['timepoints']]
+    peak = max(fit2['logLH'])
+    plt.plot(dates, np.exp(fit2['logLH'] - peak))
+
+    print("fit one mutation class")
+    fit3 = optimize_single_trajectory(counts['bins'][:cutoff], counts['mutations']['1'][:cutoff], sampling=eps)
+    dates = [datetime.fromordinal(x) for x in fit3['timepoints']]
+    peak = max(fit3['logLH'])
+    plt.plot(dates, np.exp(fit3['logLH'] - peak))
+
+
+    fig.autofmt_xdate()

### 2022-07-23 scripts/fit_outbreak.py

@@ -1,8 +1,10 @@
 import numpy as np
+import pandas as pd
 from scipy.stats import nbinom, poisson
 from scipy.optimize import minimize
 import matplotlib.pyplot as plt
 from collections import defaultdict
+from scipy.optimize import minimize
 
 TINY = 1e-100
 LogTINY = np.log(TINY)
@@ -64,7 +66,7 @@ def back_trace(data):
     logR = {}
     for i, (t, row) in enumerate(sorted_data.iterrows()):
         prev = logR[sorted_data.iloc[i-1].name] if i else None
-        logR[t] = back_propagate_R([(row.cases, row.eps)], prev, row.rho, row.k, nmin=row.cases, nmax=3*(row.cases+3)/eps)
+        logR[t] = back_propagate_R([(row.cases, row.eps)], prev, row.rho, row.k, nmin=row.cases, nmax=3*(row.cases+3)/row.eps)
 
     return logR
 
@@ -76,18 +78,23 @@ def fwd_trace(data, tau):
     for i, (t, row) in enumerate(sorted_data.iterrows()):
         prev = logQ[sorted_data.iloc[i-1].name] if i else None
         prev_cases = sorted_data.iloc[i-1].cases if i else None
-        logQ[t] = fwd_propagate_Q([(prev_cases, row.eps)], prev, row.rho, row.k, nmin=row.cases, nmax=3*(row.cases+3)/eps)
+        logQ[t] = fwd_propagate_Q([(prev_cases, row.eps)], prev, row.rho, row.k, nmin=row.cases, nmax=3*(row.cases+3)/row.eps)
 
     return logQ
 
 def negLogLH(data):
     logR = back_trace(data)
-    peak = np.max([x[1,0] for x in logR.values() if x[0,0]==1])
-    res = peak + np.log(np.sum([np.exp(x[1,0] - peak) for x in logR.values() if x[0,0]==1]))
+    pre_cases = []
+    for t in data.index:
+        if data.loc[t,"cases"]>0:
+            break
+        else:
+            pre_cases.append(t)
+    peak = np.max([logR[t][1,0] for t in pre_cases])
+    res = peak + np.log(np.sum([np.exp(logR[t][1,0] - peak) for t in pre_cases]))
     return -res
 
 def cost(x, data, fields):
-    print(x)
     for y, n in zip(x,fields):
         data[n] = y**2
     #data['eps'] = x[0]
[…]

### 2022-07-23 scripts/plots.py

@@ -1,0 +1,47 @@
+
+
+    plt.figure()
+    for x,i in intra_subs_dis.items():
+        if i>10:
+            ind = d_complete.divergence==x
+            plt.plot(sorted(d_complete.loc[ind,"numdate"]), np.linspace(0,1,i), label=f"n={x}")
+    plt.legend()
+
+    plt.figure()
+    for x,i in intra_aaSubs_dis.items():
+        if i>10:
+            ind = d_complete.aaDivergence==x
+            plt.plot(sorted(d_complete.loc[ind,"numdate"]), np.linspace(0,1,i), label=f"n={x}")
+    plt.legend()
+
+    plt.figure()
+    for x,i in intra_subs_dis.items():
+        if i>10:
+            ind = d_complete.divergence==x
+            plt.plot(sorted(d_complete.loc[ind,"numdate"]), np.arange(i), label=f"n={x}")
+    plt.ylim(1,1000)
+    plt.yscale('log')
+    plt.legend()
+
+    plt.figure()
+    for x,i in intra_aaSubs_dis.items():
+        if i>10:
+            ind = d_complete.aaDivergence==x
+            plt.plot(sorted(d_complete.loc[ind,"numdate"]), np.arange(i), label=f"n={x}")
+    plt.ylim(1,1000)
+    plt.yscale('log')
+    plt.legend()
+
+    plt.figure()
+    for x,i in intra_geno.items():
+        if i>100:
+            ind = d_complete["intra_substitutions_str"]==x
+            print(sorted(d_complete.loc[ind,"numdate"])[:3])
+            plt.plot(sorted(d_complete.loc[ind,"numdate"]), np.arange(i)+1, label=f"n={x}")
+    plt.ylim(0.5,1000)
+    plt.yscale('log')
+    plt.legend()
+
+
+    plt.figure()
+    plt.scatter(d.numdate, d.divergence)

### 2022-07-23 scripts/root_to_tip.py

@@ -1,6 +1,6 @@
 import pandas as pd
 import argparse,json
-from treetime.utils import numeric_date
+from treetime.utils import numeric_date, datestring_from_numeric
 from datetime import datetime
 import numpy as np
 from scipy.stats import linregress, scoreatpercentile
@@ -54,6 +54,34 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, completeness
 
     return d
 
+def weighted_regression(x,y,w):
+    '''
+    This function determine slope and intercept by minimizing
+    sum_i w_i*(y_i - f(x_i))^2 with f(x) = slope*x + intercept
+
+    sum_i w_i*(y_i - f(x_i)) = 0 => sum_i w_i y_i = intercept * sum_i w_i + slope sum_i w_i x_i
+    sum_i w_i*(y_i - f(x_i)) x_i = 0 => sum_i w_i y_i x_i = intercept * sum_i w_i x_i + slope sum_i w_i x_i^2
+    '''
+    wa=np.array(w)
+    xa=np.array(x)
+    ya=np.array(y)
+    wx = np.sum(xa*wa)
+    wy = np.sum(ya*wa)
+    wxy = np.sum(xa*ya*wa)
+    wxx = np.sum(xa**2*wa)
+    wsum = np.sum(wa)
+    wmean = np.mean(wa)
+
+    slope = (wy*wx - wxy*wsum)/(wx**2 - wxx*wsum)
+    intercept = (wy - slope*wx)/wsum
+
+    # not correct
+    # hessianinv = np.linalg.inv(np.array([[wxx, wx], [wx, wsum]])/wsum)
+    # stderrs = wmean*np.sqrt(hessianinv.diagonal())
+
+    return {"slope": slope, "intercept":intercept} #, 'slope_err':stderrs[0], 'intercept_err':stderrs[1]}
+
+
 def regression_by_week(d, field, min_count=5):
     val = d.loc[:,[field,'CW']].groupby('CW').mean()
     std = d.loc[:,[field,'CW']].groupby('CW').std()
@@ -68,6 +96,11 @@ def regression_by_week(d, field, min_count=5):
             "mean":[x for x in val.loc[ind, field]],
             "stderr":[x for x in std.loc[ind, field]],
             "count":[x for x in count.loc[ind, field]]}
+
+def make_date_ticks(ax):
+    ax.set_xlabel('')
[…]

### 2022-07-23 scripts/split_by_variant.py

@@ -1,29 +1,38 @@
 import pandas as pd
-import sys
+import numpy as np
+import argparse
 
 columns = ['strain', 'virus', 'gisaid_epi_isl', 'genbank_accession',
-'sra_accession', 'date', 'region', 'country', 'division', 'location',
-'length', 'host', 'age', 'sex', 'Nextstrain_clade', 'pango_lineage',
-'date_submitted', 'sampling_strategy',
+'date', 'region', 'country', 'length', 'host', 'Nextstrain_clade', 'pango_lineage',
 'Nextclade_pango', 'missing_data', 'divergence', 'nonACGTN',
 'rare_mutations', 'reversion_mutations', 'potential_contaminants',
-'QC_missing_data', 'QC_mixed_sites', 'QC_rare_mutations',
-'QC_snp_clusters', 'QC_frame_shifts', 'QC_stop_codons',
 'QC_overall_score', 'QC_overall_status', 'frame_shifts', 'deletions',
 'insertions', 'substitutions', 'aaSubstitutions', 'clock_deviation']
 
 
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
 
-metadata = sys.argv[1]
-variants = sys.argv[2:]
+    parser.add_argument('--metadata', type=str, required=True, help="input data")
+    parser.add_argument('--variant-labels', nargs='+', type=str, required=True, help="input data")
+    parser.add_argument('--variants', nargs='+', type=str, required=True, help="input data")
+    args = parser.parse_args()
 
-files = [f"subsets/{v}.tsv" for v in variants]
+    files = [f"subsets/{v}.tsv" for v in args.variant_labels]
 
-d = pd.read_csv(metadata, sep='\t', usecols=columns).fillna('')
-d["Nextstrain_clade"] = d.Nextstrain_clade.apply(lambda x:x.split()[0] if x else '')
+    d = pd.read_csv(args.metadata, sep='\t', usecols=columns).fillna('')
+    d["Nextstrain_clade"] = d.Nextstrain_clade.apply(lambda x:x.split()[0] if x else '')
 
-for v, name in zip(variants, files):
-    subset = d.loc[d.Nextstrain_clade==v]
-    print(name, v, len(subset))
-    subset.to_csv(name, sep='\t')
+    for v, name in zip(args.variants, files):
+        if len(v.split(','))>1:
[…]

### 2022-07-23 notes/variant_origin.tex
Idea for Q computation in O(N^3k)

@@
 Define the $\textbf{m}_t = (m_0, ..., m_k)_t$ vector as the sampled number of genotypes with 0,1,..., k mutations at time $t$, then  $\textbf{n}_t = (n_0, ..., n_k)_t$ is the hidden true population size at each of these time points.
-Now the probability of the ${\textbf{m}_t}_t$ vector with seeding time of the first genotype at time $\tau$ (here the hidden vector is $\textbf{n}_\tau = (1, 0, ..., 0)$) can be expressed as
+Now the probability of the ${\textbf{m}}_t$ vector with seeding time of the first genotype at time $\tau$ (here the hidden vector is $\textbf{n}_\tau = (1, 0, ..., 0)$) can be expressed as
 
@@
 
-This is due to the fact that populations of genotypes with 1 or more mutations are not only dependent on the size of that population in the previous time step but also on the size of the population with 1 less mutation (we assume that the likelihood of more than one mutation in a generation step is negligibly small).
+This is due to the fact that population sizes of genotypes with 1 or more mutations are not only dependent on the size of that population in the previous time step but also on the size of the population with 1 less mutation, i.e the likelihood a mutation happened in the last time step (we assume that the likelihood of more than one mutation in a generation step is negligibly small).
+Thus, we can describe each transition probability by the previously defined transition matrix and the mutation probability.
 
 \begin{equation}
-g'_{n_{i,t},n_{i,t-1}, n_{i-1,t-1}} = g_{n_{i,t},n_{i,t-1}} + \mu L n_{i-1, t-1}
+g'_{n_{i,t},n_{i,t-1}, n_{i-1,t-1}} = \sum_{\epsilon=0}^{n_{i,t}} g_{n_{i,t} - \epsilon,n_{i,t-1}} \cdot  \chi_{\epsilon, n_{i-1, t-1}}
 \end{equation}
@@
 
-However, this also means that the probability is not comprised of linearly independent sub-probabilities for each genotype's observed samples.
-Thus, in order to calculate the maximum likelihood seeding time $R$ and $Q$ must be calculated by iterating over a k dimensional subspace.
+However, this also means that the probability is not comprised of linearly independent sub-probabilities for each genotype's observed samples and the maximum likelihood seeding time $R$ and $Q$ are expressed by iterating over a k dimensional subspace.
 
@@
 
+For many dimensions this becomes too complex to easily compute, therefore we utilize the fact that the transition matrix g is only maximally dependent on 2 sub-spaces. Note that $Q$ is equivalent to 
+\begin{equation}
+\begin{split}
+    Q(t,\tau)_{\textbf{n}} &= P(\{m_i\}_{1 \leq i<t} | \textbf{n}_t = \textbf{n}, \tau) \\
+    & = P(\{m_{0,i}\}_{1 \leq i<t} | n_{0,t} = n^0, \tau) P(\{m_{1,i}\}_{1 \leq i<t} | \{m_{0,i}\}_{1 \leq i<t}, n_{0,t} = n^0, n_{1,t} = n^1, \tau) \cdot ...\\ \cdot & P(\{m_{k,i}\}_{1 \leq i<t} | \{m_{k-1,i}\}_{1 \leq i<t}, n_{k-1, t} = n^{k-1}, n_{k,t} = n^k, \tau)
+    \end{split}
+\end{equation}
+
+Now each of these terms can be calculated in $\mathcal{O}(N^3)$, starting with genotype 0 and moving up to genotype $k$.
+\begin{equation}
+\begin{split}
+    P(\{m_{0,i}\}_{1 \leq i<t} | n_{0,t} = n^0) 
+    & = \sum_{n'} g_{n^0, n'} P(m_{0, t-1} | n_{0, t-1} = n', \epsilon) P(\{m_{0,i}\}_{1 \leq i<(t-1)} | n_{0,t-1} = n') \\
+    &= \sum_{n'} x_{0, t-1, n^0}(n') \\
+    \\
+    P(\{m_{1,i}\}_{1 \leq i<t} | \{m_{0,i}\}_{1 \leq i<t}, n_{1,t} = n^1, n_{0,t} = n^0) 
+    & = \sum_{n_1'} P(m_{1, t-1} | n_{1, t-1} = n_1', \epsilon) \cdot \\
+        & \sum_{n_0'}  g_{n^0, n_0'} P(m_{0, t-1} | n_{0, t-1} = n_0', \epsilon) P(\{m_{0,i}\}_{1 \leq i<(t-1)} | n_{0,t-1} = n_0') \cdot \\
+        & g_{n^1, n_0', n_1'} P(\{m_{1,i}\}_{1 \leq i<(t-1)} | \{m_{0,i}\}_{1 \leq i<(t-1)}, n_{1,t-1} = n_1', n_{0,t-1} = n_0')  \\
+    &= \sum_{n_1'} P(m_{1, t-1} | n_{1, t-1} = n_1', \epsilon) \sum_{n_0'} x_{0, t-1, n^0}(n_0') y_{0, 1, t-1, n^1}(n_0', n_1') \\
+    &= \sum_{n_1'} P(m_{1, t-1} | n_{1, t-1} = n_1', \epsilon) x_{0, 1, t-1, n^0, n^1}(n_1') \\
+    \\
+    &... \\
+    \\
+    P(\{m_{k,i}\}_{1 \leq i<t} | \{m_{k-1,i}\}_{1 \leq i<t} n_{k,t} = n^k, n_{k-1,t} = n^{k-1}) 
+    &= \sum_{n_k'} P(m_{k, t-1} | n_{k, t-1} = n_k', \epsilon) \cdot \\
+    & \sum_{n_{k-1}'} y_{k-1, k, t-1, n^k}(n_{k-1}', n_k') \sum_{n^{k-2}} x_{k-2, k-1, t-1, n^{k-2}, n^{k-1}}(n_{k-1}') \\
+    &= \sum_{n_k'} P(m_{k, t-1} | n_{k, t-1} = n_k', \epsilon)   x_{k-1, k, t-1, n^{k-1}, n^k}(n_k') \\
+    \\
+\end{split}
+\end{equation}
+
+We can perform the same operations to obtain $R$.
+\begin{equation}
+\begin{split}
+    R(t)_{\textbf{n}} &= P(\{m_i\}_{t \leq i \leq T} | \textbf{n}_t = \textbf{n}) \\
+    & = P(\{m_{0,i}\}_{t \leq i \leq T} | n_{0,t} = n^0) P(\{m_{1,i}\}_{t \leq i \leq T} | \{m_{0,i}\}_{t \leq i \leq T}, n_{0,t} = n^0, n_{1,t} = n^1) \cdot ...\\ \cdot & P(\{m_{k,i}\}_{t \leq i \leq T} | \{m_{k-1,i}\}_{t \leq i \leq T}, n_{k-1, t} = n^{k-1}, n_{k,t} = n^k)
+    \end{split}
+\end{equation}
+\begin{equation}
+    \begin{split}
+P(\{m_{0,i}\}_{t \leq i \leq T} &| n_{0,t} = n^0) = P(m_{0,t} | n_{0,t} = n^0, \epsilon) \sum_{n'} g_{n', n^0}  P(\{m_{0,i}\}_{t+1 \leq i \leq T} | n_{0,t+1} = n')\\
+&= P(m_{0,t} | n_{0,t} = n^0, \epsilon) \sum_{n'} x_{0, n^0}(n')
+\\
+P(\{m_{1,i}\}_{t \leq i \leq T} &| \{m_{0,i}\}_{t \leq i \leq T}, n_{0:1,t} = (n^0, n^1)) = P(m_{1,t} | n_{1,t} = n^1, \epsilon) P(\{m_{1,i}\}_{t+1 \leq i \leq T} | \{m_{0,i}\}_{t \leq i \leq T}, n_{0,t} = n^0, n_{1,t} = n^1) \\
+&= P(m_{1,t} | n_{1,t} = n^1, \epsilon) \cdot\\
+& \sum_{n_1'} \sum_{n_0'} P(\{m_{1,i}\}_{t+1 \leq i \leq T} | \{m_{0,i}\}_{t \leq i \leq T}, n_{0,t} = n^0, n_{1,t} = n^1, n_{1, t+1} = n_1', n_{0, t+1} = n_0')\\
+&= P(m_{1,t} | n_{1,t} = n^1, \epsilon) \sum_{n_1'} \sum_{n_0'} g_{n_1', n^0, n^1} x_{0, n^0}(n_0') \cdot \\
+&P(\{m_{1,i}\}_{t+1 \leq i \leq T} | \{m_{0,i}\}_{t+1 \leq i \leq T}, n_{1, t+1} = n_1', n_{0, t+1} = n_0')
+\\
+&= P(m_{1,t} | n_{1,t} = n^1, \epsilon) \sum_{n_1'} x_{0, 1, n^0, n^1}(n_1')
+    \\
+    &... \\
+    \\
+P(\{m_{k,i}\}_{t \leq i \leq T} &| \{m_{k-1,i}\}_{t \leq i \leq T}, n_{k-1:k,t}= (n^{k-1}, n^k)) 
+= P(m_{k,t} | n_{k,t} = n^k, \epsilon) \sum_{n_k'} \sum_{n_{k-1}'} g_{n_{k-1}', n^{k-1}, n^k} \cdot \\
+& \sum_{n_{k-2}'} x_{k-2, k-1, n_{k-2}', n^{k-1}}(n_{k-1}')
+P(\{m_{k,i}\}_{t+1 \leq i \leq T} | \{m_{k-1,i}\}_{t+1 \leq i \leq T}, n_{k, t+1} = n_k', n_{k-1, t+1} = n_{k-1}')
+\\
+&= P(m_{k,t} | n_{k,t} = n^k, \epsilon) \sum_{n_k'} x_{k-1, k, n^{k-1}, n^k}(n_k')
+    \end{split}
+\end{equation}
 \end{document}

### 2022-07-23 scripts/root_to_tip.py
add weighted regression

@@ -87,9 +87,13 @@ def regression_by_week(d, field, min_count=5):
     std = d.loc[:,[field,'CW']].groupby('CW').std()
     count = d.loc[:,[field,'CW']].groupby('CW').count()
     ind = count[field]>min_count
-    reg = linregress(val.index[ind], val.loc[ind, field])
-    slope = reg.slope * 365 / 7
-    intercept = reg.intercept - 2020*slope
+    #reg = linregress(val.index[ind], val.loc[ind, field])
+    # slope = reg.slope * 365 / 7
+    # intercept = reg.intercept - 2020*slope
+    reg = weighted_regression(val.index[ind], val.loc[ind, field], np.array((count[ind]-min_count)**0.25).squeeze())
+    slope = reg["slope"] * 365 / 7
+    intercept = reg["intercept"] - 2020*slope
+
     return {"slope":slope, "intercept":intercept,
             "origin": -intercept/slope,
             "date":[week_since2020_to_numdate(x) for x in val.index[ind]],

### 2022-07-25 scripts/clone_growth.py
fix scripts to all for inclusive clades

@@ -25,7 +25,7 @@ if __name__=="__main__":
     args = parser.parse_args()
 
     with open(args.clade_gts) as fh:
-        clade_gt = json.load(fh)[args.clade]
+        clade_gt = json.load(fh)[args.clade[:3]]
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.min_date + 0.3, completeness=0)

### 2022-07-25 scripts/get_genotype_counts.py

@@ -21,7 +21,7 @@ if __name__=="__main__":
     args = parser.parse_args()
 
     with open(args.clade_gts) as fh:
-        clade_gt = json.load(fh)[args.clade]
+        clade_gt = json.load(fh)[args.clade[:3]]
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date, completeness=0)

### 2022-07-25 manuscript/variant_rates.tex
add mutations to plots

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

\newcommand{\bq}{\begin{equation}}
\newcommand{\eq}{\end{equation}}
\newcommand{\bn}{\begin{eqnarray}}
\newcommand{\en}{\end{eqnarray}}
\newcommand{\Richard}[1]{{\color{red}Richard: #1}}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\mat}[1]{{\bf #1}}
\newcommand{\vecb}[1]{{\bf #1}}
\newcommand{\abet}{\mathcal{A}}
\newcommand{\eqp}{p}
\newcommand{\LH}{\mathcal{L}}

\bibliography{bib}

\begin{document}
\title{Evolutionary rates of SARS-CoV-2}
\author{Richard A.~Neher}
\date{\today}
\maketitle

Since its emergence in late 2019, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large unobserved jumps in sequence space giving rise to new variants \citep{alpha,beta,gamma,delta,omicron}.
\begin{itemize}
    \item hypothesis of the jumps
    \item link to other examples of decaying rate of evolution H1N1pdm, HIV within host
    \item
\end{itemize}

Traditionally, evolutionary rates and divergence times are estimates using phylogenetic frameworks for heterochronously sampled sequences.
The volume and heterogeneity of SARS-CoV-2 data, however, mean that one has to dramatically down-sample the data and remove many low quality sequences that would otherwise fatally distort the analysis.

To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and robust procedures to analyze the evolutionary patterns.
We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain \citep{hadfield_nextstrain_2018} clades and analyze each clade separately.
These clades represent well defined groups of sequences with little non-vertical evolution within them.
Furthermore, we define a ``founder'' genotype for each clade and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with reference sequence.
For this reduced set of sequences, we determine the mutations they carry on top of the founder genotype.
The latter step is done for nucleotide changes as well as for amino acid changes.

Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
To further exclude outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the variance by 3.

After removing these outliers, we now bin the data by calendar week and determine the mean and standard deviation in each bin.
Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each bin is weighted with the fourth root of the number of sequences in the bins.
The exact functional form of this weighing does not have a big influence on the results, but a strongly sublinear weighing helps to counter the large variation in sequencing effort and the natural imbalance due to the fact that few sequences are available early.

Due to shared ancestry, a divergence (or root-to-tip) regression against time is not a suitable method to determine the evolutionary rate.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from a basal polytomy.
Along each of these lineages, mutation accumulation is independent.
\begin{figure*}
    \includegraphics[width=\textwidth]{figures/20I_rtt.png}
    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.}
\end{figure*}

\end{document}

### 2022-07-25 scripts/clone_growth.py

@@ -19,13 +19,13 @@ if __name__=="__main__":
     parser.add_argument('--metadata', type=str, nargs='+', required=True, help="input data")
     parser.add_argument('--clade-gts', type=str, required=True, help="input data")
     parser.add_argument('--clade', type=str, required=True, help="input data")
+    parser.add_argument('--sub-clades', type=str, required=True, help="input data")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--output-plot', type=str, help="plot file")
     #parser.add_argument('--output-json', type=str, help="rate file")
     args = parser.parse_args()
 
-    with open(args.clade_gts) as fh:
-        clade_gt = json.load(fh)[args.clade[:3]]
+    clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.min_date + 0.3, completeness=0)

### 2022-07-25 scripts/get_genotype_counts.py

@@ -2,8 +2,8 @@ import pandas as pd
 import argparse,json
 from datetime import datetime
 import numpy as np
-from root_to_tip import filter_and_transform
-
+from root_to_tip import filter_and_transform, get_clade_gts
+from collections import defaultdict
 
 if __name__=="__main__":
     parser = argparse.ArgumentParser(
@@ -14,32 +14,49 @@ if __name__=="__main__":
     parser.add_argument('--metadata', nargs='+', type=str, required=True, help="input data")
     parser.add_argument('--clade-gts', type=str, required=True, help="input data")
     parser.add_argument('--clade', type=str, required=True, help="input data")
+    parser.add_argument('--sub-clades', type=str, required=True, help="input data")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--max-date', type=float, help="input data")
     parser.add_argument('--bin-size', type=float, help="input data")
     parser.add_argument('--output-json', type=str, help="rate file")
     args = parser.parse_args()
 
-    with open(args.clade_gts) as fh:
-        clade_gt = json.load(fh)[args.clade[:3]]
+    clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date, completeness=0)
     filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())
+
     print("clade", args.clade, "done filtering")
     bins = np.arange(np.min(filtered_data.day),np.max(filtered_data.day), args.bin_size)
     all_sequences = np.histogram(filtered_data.day, bins=bins)[0]
     cumulative_sum = np.zeros_like(bins[:-1])
 
-    mutations = {}
+    mutation_number = {}
     nmax = 5
     for n in range(nmax):
         ind = filtered_data["divergence"]==n
-        mutations[n] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
-        cumulative_sum += mutations[n]
-    mutations[f'{nmax}+'] = all_sequences - cumulative_sum
+        mutation_number[n] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
+        cumulative_sum += mutation_number[n]
[…]

### 2022-07-25 scripts/plot_genotype_counts.py

@@ -22,20 +22,28 @@ if __name__=="__main__":
 
 
     dates = [datetime.fromordinal(x) for x in counts['bins'][:-1]]
-    fig, axs = plt.subplots(2,1, figsize = (8,15), sharex=True)
+    fig, axs = plt.subplots(3,1, figsize = (8,20), sharex=True)
 
     axs[0].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
-    for m in counts['mutations']:
-        axs[0].plot(dates, counts['mutations'][m], '-o', label=f'{m} mutations')
+    for m in counts['mutation_number']:
+        axs[0].plot(dates, counts['mutation_number'][m], '-o', label=f'{m} mutations')
 
     axs[0].set_yscale('log')
     axs[0].legend()
 
     axs[1].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
-    for m in counts['genotypes']:
-        axs[1].plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")
+    for m in counts['mutations']:
+        axs[1].plot(dates, counts['mutations'][m], '-o', label=f'{m}')
 
     axs[1].set_yscale('log')
     axs[1].legend()
+
+    axs[2].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
+    for m in counts['genotypes']:
+        axs[2].plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")
+
+    axs[2].set_yscale('log')
+    axs[2].legend()
+    fig.autofmt_xdate()
 
     plt.savefig(args.output_plot)

### 2022-07-25 scripts/root_to_tip.py

@@ -7,6 +7,7 @@ from scipy.stats import linregress, scoreatpercentile
 import matplotlib.pyplot as plt
 import matplotlib as mpl
 import seaborn as sns
+from collections import defaultdict
 mpl.rcParams['axes.formatter.useoffset'] = False
 
 date_reference = datetime(2020,1,1).toordinal()
@@ -42,10 +43,12 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, completeness
     d["intra_substitutions_str"] = d.intra_substitutions.apply(lambda x: ','.join(x))
     # define "with-in clade substitutions"
     d["intra_aaSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y] if x else [])
+    d["intra_SpikeSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y and y[0]=='S'] if x else [])
 
     # within clade divergence
     d["divergence"] =   d.intra_substitutions.apply(lambda x:   len(x))
     d["aaDivergence"] = d.intra_aaSubstitutions.apply(lambda x: len(x))
+    d["spikeDivergence"] = d.intra_SpikeSubstitutions.apply(lambda x: len(x))
     d["synDivergence"] = d["divergence"] - d["aaDivergence"]
 
     # filter
@@ -105,6 +108,19 @@ def make_date_ticks(ax):
     ax.set_xlabel('')
     ax.set_xticklabels([datestring_from_numeric(x) for x in ax.get_xticks()], rotation=30, horizontalalignment='right')
 
+def get_clade_gts(all_gts, subclade_str):
+    with open(all_gts) as fh:
+        clade_gts = json.load(fh)
+
+    subclades = subclade_str.split(',')
+    if len(subclades)>1:
+        clade_gt = {'nuc':{}, 'aa':{}}
+        clade_gt['nuc'] = set.intersection(*[set(clade_gts[x]['nuc']) for x in subclades])
+        clade_gt['aa'] = set.intersection(*[set(clade_gts[x]['aa']) for x in subclades])
+    else:
+        clade_gt = clade_gts[subclade_str]
+
+    return clade_gt
 
 if __name__=="__main__":
     parser = argparse.ArgumentParser(
@@ -115,14 +131,14 @@ if __name__=="__main__":
     parser.add_argument('--metadata', type=str, required=True, help="input data")
[…]

### 2022-07-26 scripts/clone_growth.py
fix missing import

@@ -8,7 +8,7 @@ mpl.rcParams['axes.formatter.useoffset'] = False
 
 import matplotlib.pyplot as plt
 import seaborn as sns
-from root_to_tip import filter_and_transform
+from root_to_tip import filter_and_transform, get_clade_gts
 
 if __name__=="__main__":
     parser = argparse.ArgumentParser(

### 2022-07-29 manuscript/variant_rates.tex
add bib etc

@@
-\documentclass[aps,rmp, onecolumn]{revtex4}
+\documentclass[aps,rmp, twocolumn]{revtex4}
 %\documentclass[a4paper,10pt]{scrartcl}
@@
 \usepackage{color}
+\usepackage{booktabs}
 %\usepackage{cite}
@@
 
-\bibliography{bib}
-
 \begin{document}
@@
 
-Since its emergence in late 2019, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large unobserved jumps in sequence space giving rise to new variants \citep{alpha,beta,gamma,delta,omicron}.
-\begin{itemize}
-    \item hypothesis of the jumps
-    \item link to other examples of decaying rate of evolution H1N1pdm, HIV within host
-    \item
-\end{itemize}
+Since its emergence in late 2019, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large unobserved jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
+Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
+Variants of Concern or Interest were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
+The branches leading to these variants are characterized by many amino acid changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.
+This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells in common among many RNA viruses and for example well studied in influenza virus A evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
+
+In contrast to influenza virus evolution, the evolution of SARS-CoV-2 has been characterized by unusually large jumps in sequence space.
+New variants with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, most dramatically with the emergence of Omicron in later 2021 \citep{viana_rapid_2022}.
+The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are frequently observed in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
+Onward transmission from such chronic infections has also been documented \citep{gonzalez-reiche_intrahost_2022}.
+However, to date there is no direct evidence for the emergence of any variant and the degree to which chronic infections might have contributed is a matter of debate.
+The case for chronic infection being an important contributor is strongest for the variants Alpha and Omicron
+\citep{hill_origins_2022}.
+
+Here, we investigate the rate at which the SARS-CoV-2 genome changes within variants and compare this rate to the rate at which the variants themselves accumulate differences relative to reference sequence.
+
+\section*{Results}
 
@@
 However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from a basal polytomy.
-Along each of these lineages, mutation accumulation is independent.
-\begin{figure*}
+Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate.
+Fig.~\ref{fig:rate_alpha} shows the increasing intra-clade divergence for clade 20I corresponding to VoC Alpha.
+
+\begin{figure*}[tb]
     \includegraphics[width=\textwidth]{figures/20I_rtt.png}
@@
     The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter.
-Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.}
+Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
+    \label{fig:rate_alpha}}
 \end{figure*}
 
+For almost all major Nextstrain clades, the average divergence increases linearly in time (similar patterns as Fig.~\ref{fig:rate_alpha}) allowing us to estimate clade specific evolutionary rates for amino acid and silent changes.
+These rates are summarized in Fig.~\ref{fig:rate_summary}.
+
+\begin{figure*}
+    \includegraphics[width=\textwidth]{figures/rate_summary.png}
+    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
+In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
+    The standard deviation expected based on Poisson statistics is indicated as shared area.
+    Panel D summarizes the individual rate estimate (dots) and compares them to the estimate inter-clade rates (gray lines).
+    \label{fig:rate_summary} }
+\end{figure*}
+
+Rates of synonymous change are very consistent across variants and also agree with the overall rate of synonymous changes of about 5-8 changes per genome per year.
+Rates of non-synonymous changes are much more variable.
+Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
+Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year.
+In contrast, the int-clade non-synonymous rate exceeds 25 changes per year.
+
+\begin{table*}
+\begin{tabular}{l|rrrrrr}
+    \hline
+    {\bf clade} &  overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
+    \hline
+      19A &     19.21 &   11.85 &     7.36 &     0 &       0 &        0 \\
+      19B &     24.40 &   14.50 &     9.90 &     2 &       1 &        1 \\
+      20A &     20.79 &   11.44 &     9.35 &     4 &       2 &        2 \\
+      20B &     14.64 &    8.42 &     6.22 &     7 &       4 &        3 \\
+      20C &     19.22 &   11.64 &     7.57 &     6 &       4 &        2 \\
+     20A+ &     18.48 &   10.26 &     8.22 &     4 &       2 &        2 \\
+      20E &     13.77 &    6.04 &     7.73 &    11 &       4 &        7 \\
+      20H &     11.16 &    5.22 &     5.94 &    19 &      15 &        4 \\
+      20I &     11.65 &    5.77 &     5.88 &    28 &      18 &       10 \\
+      20J &     10.93 &    2.66 &     8.28 &    33 &      21 &       12 \\
+      21I &     16.53 &    8.88 &     7.65 &    28 &      23 &        5 \\
+      21J &     16.20 &    8.19 &     8.00 &    32 &      27 &        5 \\
+      21K &     12.74 &    5.41 &     7.34 &    54 &      45 &        9 \\
+      21L &      8.66 &    4.67 &     3.99 &    66 &      49 &       17 \\
+      22A &      8.46 &    2.92 &     5.55 &    69 &      51 &       18 \\
+      22B &     13.78 &    4.13 &     9.65 &    64 &      49 &       15 \\
+      \hline
+    \end{tabular}
+\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.} \label{tab:rates}}
+\end{table*}
+
+\begin{figure}
+    \includegraphics[width=0.5\textwidth]{figures/rate_progression.pdf}
+    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades.
+Non-synonymous rate estimates are highest for clades 19A - 20C.
+    \label{fig:rate_progression} }
+\end{figure}
+
+\subsection*{Diversification within variants}
+
+\bibliography{bib}
 \end{document}

### 2022-07-29 scripts/combine_fits.py

@@ -18,6 +18,7 @@ if __name__=="__main__":
 
     parser.add_argument('--rate-table', type=str, required=True, help="input data")
     parser.add_argument('--output-plot', type=str, help="plot file")
+    parser.add_argument('--output-plot-rates', type=str, help="plot file")
     args = parser.parse_args()
 
     rates = pd.read_csv(args.rate_table, sep='\t', index_col='clade')
@@ -51,13 +52,25 @@ if __name__=="__main__":
     for i, (mut_type, rate) in enumerate(inter_clade_rates.items()):
         axs[-1,-1].plot([i-0.4, i+0.4], [rate, rate], lw=3, c='k', alpha=0.5)
         clade_rates = rates[f"{mut_type}_rate"]
-        axs[-1,-1].scatter(i - 0.35 + np.random.random(size=len(clade_rates))*0.7, clade_rates, c=rates[f"{mut_type}_origin"])
+        axs[-1,-1].scatter(i - 0.35 + np.random.random(size=len(clade_rates))*0.7, clade_rates, c=[f"C{ci%10}" for ci in range(len(clade_rates))])
     axs[-1,-1].set_ylabel("substitutions per year")
     axs[-1,-1].set_xticks([0,1,2], ['nuc', 'aa', 'syn'])
     axs[-1,-1].set_ylim(0)
 
-
     if args.output_plot:
         plt.savefig(args.output_plot)
+    else:
+        plt.show()
+
+    plt.figure()
+    plt.plot(rates["nuc_origin"], rates["aa_rate"], 'o', label='amino acid rate')
+    plt.plot(rates["nuc_origin"], rates["syn_rate"], 'o', label='synonymous rate')
+    plt.plot(rates["nuc_origin"], np.ones_like(rates['nuc_origin'])*inter_clade_rates["aa"], label='inter-clade amino acid rate', c=f"C{0}", lw=3)
+    plt.plot(rates["nuc_origin"], np.ones_like(rates['nuc_origin'])*inter_clade_rates["syn"], label='inter-clade synonymous rate', c=f"C{1}", lw=3)
+    plt.ylabel('rate estimate [1/y]')
+    plt.legend()
+
+    if args.output_plot_rates:
+        plt.savefig(args.output_plot_rates)
     else:
         plt.show()

### 2022-07-30 manuscript/variant_rates.tex
additions to ms

@@
 Here, we investigate the rate at which the SARS-CoV-2 genome changes within variants and compare this rate to the rate at which the variants themselves accumulate differences relative to reference sequence.
+This comparison reveals a clear dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
+The rate of synonymous changes between and within variants agrees, while the non-synonymous rate within variants is far below the between variant rate.
+Early variants display more rapid non-synonymous evolution that later variants.
 
@@
 Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year.
-In contrast, the int-clade non-synonymous rate exceeds 25 changes per year.
+In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.
 
@@
 
+The apparent time dependence of non-synonymous rate suggests that at the beginning a large fraction of amino acid changes were beneficial to viral replication, not just the D614G mutation in the spike protein \citep{korber_tracking_2020}.
+This applies both to the changes within clades that were ultimately displaced by more successful variants, and inter-variant changes where the fraction of adaptive changes is even higher.
+Adaptive evolution complicates phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology.
+This assumption is true for neutral mutations that occur along every lineage with the same rate.
+Adaptive and deleterious mutations, however, affect the ability of the virus to spread.
+Deleterious mutations are relatively straightforward to account for: Strongly deleterious mutations don't spread and lead to an evolutionary rate that is smaller than the mutation rates.
+Weakly deleterious mutations can spread, but lineages that carry them tend to be short-lived.
+A spectrum of mutations with different deleterious effects leads to time dependent effective evolutionary rates \citep{wertheim_purifying_2011}.
+The intra-clade rates in Fig.~\ref{fig:rate_progression} are all estimated over a similar time frame and the time dependent rate phenomenon should not apply.
+
+Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
+Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
+Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
+Epistatic interactions can further steepen the dependence on population size and stochasticity.
+
 \subsection*{Diversification within variants}
+The analysis of evolutionary rates above only considers the mean divergence within rates.
 

### 2022-07-30 scripts/clone_growth.py
allow root swap

@@ -28,7 +28,7 @@ if __name__=="__main__":
     clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
-    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.min_date + 0.3, completeness=0)
+    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.min_date + 0.3, completeness=0, swap_root=args.clade_gts=='19B+')
     #filtered_data=filtered_data.loc[filtered_data.country!='China']
 
     intra_subs_dis = filtered_data["divergence"].value_counts().sort_index()

### 2022-07-30 scripts/get_genotype_counts.py

@@ -24,7 +24,7 @@ if __name__=="__main__":
     clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
-    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date, completeness=0)
+    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date, completeness=0, swap_root=args.clade_gts=='19B+')
     filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())
 
     print("clade", args.clade, "done filtering")

### 2022-07-30 scripts/root_to_tip.py

@@ -20,7 +20,7 @@ def week_since2020_to_date(d):
 def week_since2020_to_numdate(d):
     return numeric_date(week_since2020_to_date(d))
 
-def filter_and_transform(d, clade_gt, min_date=None, max_date=None, completeness=None):
+def filter_and_transform(d, clade_gt, min_date=None, max_date=None, completeness=None, swap_root=False):
     # filter for incomplete data
     d = d.loc[d.date.apply(lambda x:len(x)==10 and 'X' not in x)]
     d['datetime'] = d.date.apply(lambda x: datetime.strptime(x, '%Y-%m-%d'))
@@ -38,18 +38,30 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, completeness
     d["missing_subs"] = d.clade_substitutions.apply(lambda x: len(clade_gt['nuc'])-len(x))
 
     # define "with-in clade substitutions"
-    d["intra_substitutions"] = d.substitutions.apply(lambda x:     [y for y in x.split(',') if all([y not in clade_gt['nuc'], int(y[1:-1])>150,  int(y[1:-1])<29753])] if x else [])
-    # make a hashable string representation
-    d["intra_substitutions_str"] = d.intra_substitutions.apply(lambda x: ','.join(x))
+    d["intra_substitutions"] = d.substitutions.apply(lambda x:     [y for y in x.split(',')
+                                                     if all([y not in clade_gt['nuc'], int(y[1:-1])>150,  int(y[1:-1])<29753])] if x else [])
     # define "with-in clade substitutions"
     d["intra_aaSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y] if x else [])
     d["intra_SpikeSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y and y[0]=='S'] if x else [])
 
+    if swap_root:
+        muts = [("C8782T","T8782C"), ("T28144C","C28144T")]
+        def swap(mutations, pair):
+            return [y for y in mutations if y!=pair[0]] if pair[0] in mutations else mutations + [pair[1]]
+        for m in muts:
+            d["intra_substitutions"] = d.intra_substitutions.apply(lambda x: swap(x,m))
+        aa_mut = ("ORF8:L84S","ORF8:S84L")
[…]

### 2022-07-30 scripts/root_to_tip.py
add query func

@@ -20,9 +20,11 @@ def week_since2020_to_date(d):
 def week_since2020_to_numdate(d):
     return numeric_date(week_since2020_to_date(d))
 
-def filter_and_transform(d, clade_gt, min_date=None, max_date=None, completeness=None, swap_root=False):
+def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False):
     # filter for incomplete data
     d = d.loc[d.date.apply(lambda x:len(x)==10 and 'X' not in x)]
+    if query:
+        d = d.query(query)
     d['datetime'] = d.date.apply(lambda x: datetime.strptime(x, '%Y-%m-%d'))
     d['numdate'] = d.datetime.apply(lambda x: numeric_date(x))
     d['CW'] = d.datetime.apply(date_to_week_since2020)
@@ -146,6 +148,7 @@ if __name__=="__main__":
     parser.add_argument('--sub-clades', type=str, required=True, help="input data")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--max-date', type=float, help="input data")
+    parser.add_argument('--query', type=str, help="filters")
     parser.add_argument('--output-plot', type=str, help="plot file")
     parser.add_argument('--output-json', type=str, help="rate file")
     args = parser.parse_args()
@@ -153,7 +156,9 @@ if __name__=="__main__":
     clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.read_csv(args.metadata, sep='\t').fillna('')
-    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date, completeness=0, swap_root=args.clade_gts=='19B+')
+    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
+                                         query = args.query,
+                                         completeness=0, swap_root=args.clade_gts=='19B+')
 
     regression = linregress(filtered_data.numdate, filtered_data.divergence)
     filtered_data["residuals"] = filtered_data.apply(lambda x: x.divergence - (regression.intercept + regression.slope*x.numdate), axis=1)

### 2022-07-30 scripts/get_genotype_counts.py
add query func

@@ -17,6 +17,7 @@ if __name__=="__main__":
     parser.add_argument('--sub-clades', type=str, required=True, help="input data")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--max-date', type=float, help="input data")
+    parser.add_argument('--query', type=str, help="filters")
     parser.add_argument('--bin-size', type=float, help="input data")
     parser.add_argument('--output-json', type=str, help="rate file")
     args = parser.parse_args()
@@ -24,7 +25,9 @@ if __name__=="__main__":
     clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
-    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date, completeness=0, swap_root=args.clade_gts=='19B+')
+    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
+                                         query = args.query,
+                                         completeness=0, swap_root=args.clade_gts=='19B+')
     filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())
 
     print("clade", args.clade, "done filtering")

### 2022-07-30 scripts/explore.py
fix root-swap

@@ -1,0 +1,11 @@
+import pandas as pd
+from root_to_tip import filter_and_transform, get_clade_gts
+from treetime.utils import numeric_date, datestring_from_numeric
+from datetime import datetime
+# alpha
+clade_gt = get_clade_gts("data/clade_gts.json", "20I")
+
+alpha = pd.read_csv('subsets/20I.tsv', sep='\t')
+alpha_filtered = filter_and_transform(alpha, clade_gt, min_date=numeric_date(datetime(2020,8,1)))
+
+print(alpha_filtered.loc[alpha_filtered.date<"2020-10-01",["strain", "date", "intra_substitutions"]])

### 2022-07-30 scripts/get_genotype_counts.py

@@ -27,7 +27,7 @@ if __name__=="__main__":
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
                                          query = args.query,
-                                         completeness=0, swap_root=args.clade_gts=='19B+')
+                                         completeness=0, swap_root=args.clade=='19B+')
     filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())
 
     print("clade", args.clade, "done filtering")

### 2022-07-30 scripts/plot_genotype_counts.py

@@ -32,14 +32,14 @@ if __name__=="__main__":
     axs[0].legend()
 
     axs[1].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
-    for m in counts['mutations']:
+    for m in sorted(counts['mutations'].keys(), key=lambda w:int(m[1:-1])):
         axs[1].plot(dates, counts['mutations'][m], '-o', label=f'{m}')
 
     axs[1].set_yscale('log')
     axs[1].legend()
 
     axs[2].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
-    for m in counts['genotypes']:
+    for m in sorted(counts['genotypes'].keys(), key=lambda x: len(x)):
         axs[2].plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")
 
     axs[2].set_yscale('log')

### 2022-07-30 scripts/root_to_tip.py

@@ -158,7 +158,7 @@ if __name__=="__main__":
     d = pd.read_csv(args.metadata, sep='\t').fillna('')
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
                                          query = args.query,
-                                         completeness=0, swap_root=args.clade_gts=='19B+')
+                                         completeness=0, swap_root=args.clade=='19B+')
 
     regression = linregress(filtered_data.numdate, filtered_data.divergence)
     filtered_data["residuals"] = filtered_data.apply(lambda x: x.divergence - (regression.intercept + regression.slope*x.numdate), axis=1)

### 2022-07-30 scripts/plot_genotype_counts.py
fix root-swap

@@ -32,7 +32,7 @@ if __name__=="__main__":
     axs[0].legend()
 
     axs[1].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
-    for m in sorted(counts['mutations'].keys(), key=lambda w:int(m[1:-1])):
+    for m in sorted(counts['mutations'].keys(), key=lambda x:int(x[1:-1])):
         axs[1].plot(dates, counts['mutations'][m], '-o', label=f'{m}')
 
     axs[1].set_yscale('log')

### 2022-07-30 scripts/get_genotype_counts.py
select clones based on early genotypes

@@ -59,15 +59,17 @@ if __name__=="__main__":
 
     print("clade", args.clade, "done mutations")
 
-
-    intra_geno = filtered_data["intra_substitutions_str"].value_counts()
+    cutoff = args.min_date + (args.min_date + args.max_date)/2
+    ind = filtered_data.numdate<cutoff
+    n_early = ind.sum()
+    intra_geno = filtered_data.loc[ind,"intra_substitutions_str"].value_counts()
     genotypes = {}
     for x,i in intra_geno.items():
-        if i<10 or i<len(filtered_data)/10000:
+        if i<10 or i<n_early/10000:
             break
         nmuts = len(x.split(',')) if x else 0
         ind = filtered_data["intra_substitutions_str"]==x
-        if nmuts==0 or (nmuts==1 and ind.sum()>0.003*len(filtered_data)) or (nmuts<3 and ind.sum()>0.01*len(filtered_data)):
+        if nmuts==0 or (nmuts==1 and ind.sum()>0.03*n_early) or (nmuts<4 and ind.sum()>0.1*n_early):
             genotypes[x] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
 
     print("clade", args.clade, "done genotypes")

### 2022-07-30 scripts/get_genotype_counts.py
select clones based on early genotypes

@@ -45,8 +45,11 @@ if __name__=="__main__":
     print("clade", args.clade, "done mutation_number")
 
     mutation_counts = defaultdict(int)
-    date_cutoff = args.min_date + 0.5*(args.max_date-args.min_date)
-    for muts in filtered_data.loc[filtered_data.numdate<date_cutoff, "intra_substitutions"]:
+    cutoff = args.min_date + (args.min_date + args.max_date)/2
+    ind = filtered_data.numdate<cutoff
+    n_early = ind.sum()
+
+    for muts in filtered_data.loc[ind, "intra_substitutions"]:
         for m in muts:
             mutation_counts[m] +=1
 
@@ -59,9 +62,6 @@ if __name__=="__main__":
 
     print("clade", args.clade, "done mutations")
 
-    cutoff = args.min_date + (args.min_date + args.max_date)/2
-    ind = filtered_data.numdate<cutoff
-    n_early = ind.sum()
     intra_geno = filtered_data.loc[ind,"intra_substitutions_str"].value_counts()
     genotypes = {}
     for x,i in intra_geno.items():

### 2022-07-30 scripts/get_genotype_counts.py
select clones based on early genotypes

@@ -45,11 +45,11 @@ if __name__=="__main__":
     print("clade", args.clade, "done mutation_number")
 
     mutation_counts = defaultdict(int)
-    cutoff = args.min_date + (args.min_date + args.max_date)/2
-    ind = filtered_data.numdate<cutoff
-    n_early = ind.sum()
+    cutoff = args.min_date + (args.max_date - args.min_date)/2
+    ind_early = filtered_data.numdate<cutoff
+    n_early = ind_early.sum()
 
-    for muts in filtered_data.loc[ind, "intra_substitutions"]:
+    for muts in filtered_data.loc[ind_early, "intra_substitutions"]:
         for m in muts:
             mutation_counts[m] +=1
 
@@ -62,14 +62,14 @@ if __name__=="__main__":
 
     print("clade", args.clade, "done mutations")
 
-    intra_geno = filtered_data.loc[ind,"intra_substitutions_str"].value_counts()
+    intra_geno = filtered_data.loc[ind_early,"intra_substitutions_str"].value_counts()
     genotypes = {}
     for x,i in intra_geno.items():
         if i<10 or i<n_early/10000:
             break
         nmuts = len(x.split(',')) if x else 0
         ind = filtered_data["intra_substitutions_str"]==x
-        if nmuts==0 or (nmuts==1 and ind.sum()>0.03*n_early) or (nmuts<4 and ind.sum()>0.1*n_early):
+        if nmuts==0 or (nmuts==1 and ind.sum()>0.02*n_early) or (nmuts<4 and ind.sum()>0.05*n_early):
             genotypes[x] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
 
     print("clade", args.clade, "done genotypes")

### 2022-07-30 scripts/get_genotype_counts.py
export mutation spectra

@@ -46,12 +46,30 @@ if __name__=="__main__":
 
     mutation_counts = defaultdict(int)
     cutoff = args.min_date + (args.max_date - args.min_date)/2
-    ind_early = filtered_data.numdate<cutoff
-    n_early = ind_early.sum()
+    window = cutoff - 14/365, cutoff + 14/365
 
+    ind_early = filtered_data.numdate<cutoff
+    ind_window = filtered_data.numdate>=window[0] & filtered_data.numdate<window[1]
+    n_early = ind_early.sum()
+    n_window = ind_window.sum()
+
+    # Find common mutations
     for muts in filtered_data.loc[ind_early, "intra_substitutions"]:
         for m in muts:
             mutation_counts[m] +=1
+
+    relevant_muts = [x[0] for x in sorted(list(mutation_counts.items()), key=lambda k:k[1])[-10:]]
+    mutations = {}
+    nmax = 5
+    for mut in relevant_muts:
+        ind = filtered_data["intra_substitutions"].apply(lambda x: mut in x)
+        mutations[mut] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
+
+    # Get mutation _spectrum
+    mutation_spectrum = defaultdict(float)
+    for muts in filtered_data.loc[ind_window, "intra_substitutions"]:
+        for m in muts:
+            mutation_spectrum[m] += 1.0/n_window
 
     relevant_muts = [x[0] for x in sorted(list(mutation_counts.items()), key=lambda k:k[1])[-10:]]
     mutations = {}
@@ -77,5 +95,6 @@ if __name__=="__main__":
         json.dump({'bins':[int(x) for x in bins],
                    'all_samples':[int(x) for x in all_sequences],
                    'mutation_number': {k:[int(x) for x in v] for k,v in mutation_number.items()},
+                   'mutation_spectrum': {k: float(v) for k,v in mutation_spectrum.items()},
                    'mutations': {k:[int(x) for x in v] for k,v in mutations.items()},
                    'genotypes': {k:[int(x) for x in v] for k,v in genotypes.items()}}, fh)

### 2022-07-30 scripts/get_genotype_counts.py
export mutation spectra

@@ -49,7 +49,7 @@ if __name__=="__main__":
     window = cutoff - 14/365, cutoff + 14/365
 
     ind_early = filtered_data.numdate<cutoff
-    ind_window = filtered_data.numdate>=window[0] & filtered_data.numdate<window[1]
+    ind_window = (filtered_data.numdate>=window[0]) & (filtered_data.numdate<window[1])
     n_early = ind_early.sum()
     n_window = ind_window.sum()
 

### 2022-07-30 scripts/get_genotype_counts.py
fix 22A/B window

@@ -82,13 +82,15 @@ if __name__=="__main__":
 
     intra_geno = filtered_data.loc[ind_early,"intra_substitutions_str"].value_counts()
     genotypes = {}
+    gt_count = 0
     for x,i in intra_geno.items():
-        if i<10 or i<n_early/10000:
-            break
         nmuts = len(x.split(',')) if x else 0
         ind = filtered_data["intra_substitutions_str"]==x
-        if nmuts==0 or (nmuts==1 and ind.sum()>0.02*n_early) or (nmuts<4 and ind.sum()>0.05*n_early):
+        if nmuts==0 or (nmuts==1 and ind.sum()>0.02*n_early) or (ind.sum()>0.05*n_early):
             genotypes[x] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
+            gt_count +=1
+            if gt_count>10:
+                break
 
     print("clade", args.clade, "done genotypes")
     with open(args.output_json, 'w') as fh:

### 2022-07-30 scripts/plot_af.py

@@ -1,0 +1,42 @@
+import argparse,json
+from datetime import datetime
+import numpy as np
+import matplotlib as mpl
+mpl.rcParams['axes.formatter.useoffset'] = False
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+
+    parser.add_argument('--counts', nargs='+', type=str, required=True, help="input data")
+    parser.add_argument('--output-plot', type=str, help="figure file")
+    args = parser.parse_args()
+
+    fig, axs = plt.subplots(1,1, figsize = (8,6))
+    ax = axs
+    ls = ['-', '-.', '--']
+    for fi,fname in enumerate(args.counts):
+        with open(fname) as fh:
+            counts = json.load(fh)
+
+        clade = fname.split('/')[-1].split('_')[0]
+        dates = [datetime.fromordinal(x) for x in counts['bins'][:-1]]
+
+        ax.plot(sorted(counts["mutation_spectrum"].values()), np.linspace(1,0,len(counts["mutation_spectrum"])+1)[:-1],
+                       label=clade, ls=ls[fi//10])
+
+    ax.set_yscale('log')
+    ax.set_xscale('log')
+    ax.set_xlabel('mutation frequency')
+    ax.set_ylabel('fraction above')
+    x = np.logspace(-4,-0.3,21)
+    ax.plot([1e-5,1e-1], [1e-0,1e-4], c='k', lw=3, alpha=0.5, label='1/x')
+    # ax.plot(x, np.log(x)/np.log(x[0]), c='k', lw=3, alpha=0.5, label='~log(x)')
+    plt.legend(ncol=3)
+    plt.savefig(args.output_plot)

### 2022-07-30 scripts/plot_genotype_counts.py

@@ -1,5 +1,6 @@
 import argparse,json
 from datetime import datetime
+import numpy as np
 import matplotlib as mpl
 mpl.rcParams['axes.formatter.useoffset'] = False
 
@@ -22,28 +23,35 @@ if __name__=="__main__":
 
 
     dates = [datetime.fromordinal(x) for x in counts['bins'][:-1]]
-    fig, axs = plt.subplots(3,1, figsize = (8,20), sharex=True)
+    fig, axs = plt.subplots(1,3, figsize = (18,6), sharey=True)
 
-    axs[0].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
+    ax = axs[0]
+    ax.plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
     for m in counts['mutation_number']:
-        axs[0].plot(dates, counts['mutation_number'][m], '-o', label=f'{m} mutations')
+        ax.plot(dates, counts['mutation_number'][m], '-o', label=f'{m} mutations')
 
-    axs[0].set_yscale('log')
-    axs[0].legend()
+    ax.set_yscale('log')
+    ax.legend()
 
-    axs[1].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
+    ax = axs[1]
+    ax.plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
     for m in sorted(counts['mutations'].keys(), key=lambda x:int(x[1:-1])):
-        axs[1].plot(dates, counts['mutations'][m], '-o', label=f'{m}')
+        ax.plot(dates, counts['mutations'][m], '-o', label=f'{m}')
 
-    axs[1].set_yscale('log')
-    axs[1].legend()
+    ax.set_yscale('log')
+    ax.legend()
 
-    axs[2].plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
+    ax = axs[2]
+    ax.plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
     for m in sorted(counts['genotypes'].keys(), key=lambda x: len(x)):
-        axs[2].plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")
+        ax.plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")
 
-    axs[2].set_yscale('log')
-    axs[2].legend()
+    ax.set_yscale('log')
+    ax.legend()
[…]

### 2022-07-30 scripts/root_to_tip.py

@@ -142,7 +142,7 @@ if __name__=="__main__":
         formatter_class=argparse.ArgumentDefaultsHelpFormatter
     )
 
-    parser.add_argument('--metadata', type=str, required=True, help="input data")
+    parser.add_argument('--metadata',  type=str, required=True, help="input data")
     parser.add_argument('--clade-gts', type=str, required=True, help="input data")
     parser.add_argument('--clade', type=str, required=True, help="input data")
     parser.add_argument('--sub-clades', type=str, required=True, help="input data")
@@ -175,7 +175,7 @@ if __name__=="__main__":
     regression_clean_syn = regression_by_week(filtered_data.loc[ind], "synDivergence")
     regression_clean_spike = regression_by_week(filtered_data.loc[ind], "spikeDivergence")
 
-    fig, axs = plt.subplots(1,3, figsize=(15,6), sharex=True, sharey=True)
+    fig, axs = plt.subplots(1,3, figsize=(18,6), sharex=True, sharey=True)
     ymax = 20
     bins = bins=(20,np.arange(-0.5,ymax+0.5))
     sns.histplot(x=filtered_data.numdate, y=np.minimum(ymax*1.5, filtered_data.divergence), bins=bins, ax=axs[0])

### 2022-07-30 scripts/get_genotype_counts.py
filter for good

@@ -86,6 +86,8 @@ if __name__=="__main__":
     for x,i in intra_geno.items():
         nmuts = len(x.split(',')) if x else 0
         ind = filtered_data["intra_substitutions_str"]==x
+        if i<10 and nmuts:
+            continue
         if nmuts==0 or (nmuts==1 and ind.sum()>0.02*n_early) or (ind.sum()>0.05*n_early):
             genotypes[x] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
             gt_count +=1

### 2022-07-30 scripts/root_to_tip.py

@@ -23,6 +23,7 @@ def week_since2020_to_numdate(d):
 def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False):
     # filter for incomplete data
     d = d.loc[d.date.apply(lambda x:len(x)==10 and 'X' not in x)]
+    d = d.loc[d.QC_overall_status=='good']
     if query:
         d = d.query(query)
     d['datetime'] = d.date.apply(lambda x: datetime.strptime(x, '%Y-%m-%d'))

### 2022-07-30 scripts/get_genotype_counts.py
add max groups argument

@@ -17,6 +17,7 @@ if __name__=="__main__":
     parser.add_argument('--sub-clades', type=str, required=True, help="input data")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--max-date', type=float, help="input data")
+    parser.add_argument('--max-group', type=int, default=100, help="input data")
     parser.add_argument('--query', type=str, help="filters")
     parser.add_argument('--bin-size', type=float, help="input data")
     parser.add_argument('--output-json', type=str, help="rate file")
@@ -26,7 +27,7 @@ if __name__=="__main__":
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
-                                         query = args.query,
+                                         query = args.query, max_group=args.max_group,
                                          completeness=0, swap_root=args.clade=='19B+')
     filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())
 

### 2022-07-30 scripts/root_to_tip.py

@@ -20,7 +20,7 @@ def week_since2020_to_date(d):
 def week_since2020_to_numdate(d):
     return numeric_date(week_since2020_to_date(d))
 
-def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False):
+def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False, max_group=None):
     # filter for incomplete data
     d = d.loc[d.date.apply(lambda x:len(x)==10 and 'X' not in x)]
     d = d.loc[d.QC_overall_status=='good']
@@ -69,6 +69,9 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
     # filter
     if completeness is not None:
         return d.loc[d.missing_subs<=completeness]
+
+    if max_group:
+        return d.groupby(['CW', 'country']).sample(max_group, replace=True).drop_duplicates(subset='strain')
 
     return d
 
@@ -149,6 +152,7 @@ if __name__=="__main__":
     parser.add_argument('--sub-clades', type=str, required=True, help="input data")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--max-date', type=float, help="input data")
+    parser.add_argument('--max-group', type=int, default=100, help="input data")
     parser.add_argument('--query', type=str, help="filters")
     parser.add_argument('--output-plot', type=str, help="plot file")
     parser.add_argument('--output-json', type=str, help="rate file")
@@ -158,7 +162,7 @@ if __name__=="__main__":
 
     d = pd.read_csv(args.metadata, sep='\t').fillna('')
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
-                                         query = args.query,
+                                         query = args.query, max_group=args.max_group,
                                          completeness=0, swap_root=args.clade=='19B+')
 
     regression = linregress(filtered_data.numdate, filtered_data.divergence)

### 2022-07-30 scripts/root_to_tip.py
fix subsampling

@@ -68,7 +68,7 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
 
     # filter
     if completeness is not None:
-        return d.loc[d.missing_subs<=completeness]
+        d = d.loc[d.missing_subs<=completeness]
 
     if max_group:
         return d.groupby(['CW', 'country']).sample(max_group, replace=True).drop_duplicates(subset='strain')

### 2022-07-30 scripts/get_genotype_counts.py
make qc exception for 21H

@@ -27,7 +27,7 @@ if __name__=="__main__":
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
-                                         query = args.query, max_group=args.max_group,
+                                         query = args.query, max_group=args.max_group, QC_threshold=80 if args.clade=='21H' else 30,
                                          completeness=0, swap_root=args.clade=='19B+')
     filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())
 

### 2022-07-30 scripts/root_to_tip.py

@@ -20,10 +20,10 @@ def week_since2020_to_date(d):
 def week_since2020_to_numdate(d):
     return numeric_date(week_since2020_to_date(d))
 
-def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False, max_group=None):
+def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False, max_group=None, QC_threshold=30):
     # filter for incomplete data
     d = d.loc[d.date.apply(lambda x:len(x)==10 and 'X' not in x)]
-    d = d.loc[d.QC_overall_status=='good']
+    d = d.loc[d.QC_overall_score<QC_threshold]
     if query:
         d = d.query(query)
     d['datetime'] = d.date.apply(lambda x: datetime.strptime(x, '%Y-%m-%d'))
@@ -162,7 +162,7 @@ if __name__=="__main__":
 
     d = pd.read_csv(args.metadata, sep='\t').fillna('')
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
-                                         query = args.query, max_group=args.max_group,
+                                         query = args.query, max_group=args.max_group, QC_threshold=80 if args.clade=='21H' else 30,
                                          completeness=0, swap_root=args.clade=='19B+')
 
     regression = linregress(filtered_data.numdate, filtered_data.divergence)

### 2022-07-31 manuscript/variant_rates.tex
allow root swap
fix bib

@@
 \begin{figure*}[tb]
-    \includegraphics[width=\textwidth]{figures/20I_rtt.png}
+    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.png}
     \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
@@
 Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
+Analogous figures for all variants are included in the appendix.
     \label{fig:rate_alpha}}
@@
     \includegraphics[width=0.5\textwidth]{figures/rate_progression.pdf}
-    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades.
+    \caption{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades.
 Non-synonymous rate estimates are highest for clades 19A - 20C.
@@
 Epistatic interactions can further steepen the dependence on population size and stochasticity.
+Methods to infer emergence dates and locations of the early variants might therefore be biased or over-confident.
 
 \subsection*{Diversification within variants}
-The analysis of evolutionary rates above only considers the mean divergence within rates.
+The analysis of evolutionary rates above only considers the mean divergence within clades which, at least when variation is neutral, is independent of epidemiological parameters.
+The structure of within clade diversity does contain information on the rate of spread.
+A simple model for a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
+Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
+The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5days$ and a rate of $\mu = 15/year$ evaluates to $u=0.2$.
+Since these branching process models are linear, them mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation.
+\begin{equation}
+    \begin{split}
+        \frac{dn}{dt} &= \alpha n \\
+        \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
+        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
+        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
+        \cdots
+    \end{split}
+\end{equation}
+with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
+These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
+In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.
 
 \bibliography{bib}
+\appendix
+\include{appendix_rtt}
+\include{counts_appendix}
 \end{document}

### 2022-07-31 manuscript/appendix_rtt.tex
add reference, 19B++ clade

\section*{Divergence summaries}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/19A_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 19A.}
    \label{fig:19A_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/19B_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 19B.}
    \label{fig:19B_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/19B+_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 19B+.}
    This figure contains sequences in clades 19 A and B rooted on clade 19B.
    \label{fig:19B+_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/20A_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 20A.}
    \label{fig:20A_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/20B_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 20B.}
    \label{fig:20B_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/20C_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 20C.}
    \label{fig:20C_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/20A+_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 20A+.}
    This figure contains sequences in clades 20A,B,C,D rooted on clade 20A.
    \label{fig:20A+_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/20E_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 20E.}
    \label{fig:20E_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/20H_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 20H (Beta).}
    \label{fig:20H_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 20I (Alpha).}
    \label{fig:20I_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/20J_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 20J (Gamma).}
    \label{fig:20J_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/21D_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 21D (Eta).}
    \label{fig:21D_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/21G_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 21G (Lambda).}
    \label{fig:21G_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/21H_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 21H (Mu).}
    \label{fig:21H_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/21I_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 21I (Delta).}
    \label{fig:21I_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/21J_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 21J (Delta).}
    \label{fig:21J_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/21K_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 21K (Omicron).}
    \label{fig:21K_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/21L_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 21L (Omicron).}
    \label{fig:21L_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/22A_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 22A (Omicron).}
    \label{fig:22A_divergence}}
\end{figure*}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rtt/22B_rtt.png}
    \caption{{\bf Divergence increases linearly with time in clade 22B (Omicron).}
    \label{fig:22B_divergence}}
\end{figure*}

### 2022-07-31 manuscript/variant_rates.tex
add reference, 19B++ clade

@@
 Fig.~\ref{fig:rate_alpha} shows the increasing intra-clade divergence for clade 20I corresponding to VoC Alpha.
+Both synonymous and non-synonymous within-clade average divergence increases linearly over time allowing for a robust estimate of the rate.
 
@@
 In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.
+
+Nextstrain clades tend to be defined by long branches leading to a large polytomy.
+It is thus expected that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
+This effect might be particularly important early on in the pandemic where branches with as little two mutations were used to define new clades.
+We thus also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
+While we estimate somewhat higher rates for these composite clades (in particular 19B++), the overall pattern remains consistent.
 
@@
 This applies both to the changes within clades that were ultimately displaced by more successful variants, and inter-variant changes where the fraction of adaptive changes is even higher.
-Adaptive evolution complicates phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology.
-This assumption is true for neutral mutations that occur along every lineage with the same rate.
-Adaptive and deleterious mutations, however, affect the ability of the virus to spread.
-Deleterious mutations are relatively straightforward to account for: Strongly deleterious mutations don't spread and lead to an evolutionary rate that is smaller than the mutation rates.
-Weakly deleterious mutations can spread, but lineages that carry them tend to be short-lived.
-A spectrum of mutations with different deleterious effects leads to time dependent effective evolutionary rates \citep{wertheim_purifying_2011}.
-The intra-clade rates in Fig.~\ref{fig:rate_progression} are all estimated over a similar time frame and the time dependent rate phenomenon should not apply.
-
-Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
-Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
-Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
-Epistatic interactions can further steepen the dependence on population size and stochasticity.
-Methods to infer emergence dates and locations of the early variants might therefore be biased or over-confident.
 
@@
 
+\section*{Discussion}
+The inferred evolutionary rate of RNA viruses often decreases with the time scale across which is it estimated \citep{wertheim_purifying_2011,ghafari_prisoner_2021}.
+This effect is particularly pronounced at the beginning of an outbreak following a host switch and has been attributed to methodological issues and incomplete purifying selection leading to inflated measures of diversity \citep{meyer_time_2015,ghafari_purifying_2022}.
+In addition to segregating deleterious mutations, early viral evolution after a host switch will also be affect by anomalously fast adaptive evolution as a dramatic change in environment, i.e.~a host switch, likely results in many mutations that increase fitness, possibly only in epistatic combinations with others.
+Such transient increases in the rate of adaptation is common in experimental evolution \citep{elena_virus_2007}.
+
+The analysis presented here suggests that the early evolutionary rate of SARS-CoV-2 was inflated by adaptive evolution.
+Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of 6 month, purifying selection has comparable efficiency for all clades.
+Nevertheless, the non-synonymous evolutionary of clades circulating in late 2019 and early 2020 is about twice as high as that of clades in 2021 and 2020.
+
+Adaptive evolution complicates phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology.
+This assumption is true for neutral mutations that occur along every lineage with the same rate.
+Adaptive and deleterious mutations, however, affect the ability of the virus to spread.
+
+Deleterious mutations are relatively straightforward to account for: Strongly deleterious mutations don't spread and are only observed on terminal branches, similar to sequencing errors.
+Weakly deleterious mutations can spread, but lineages that carry them tend to be short-lived.
+A spectrum of mutations with different deleterious effects leads to time dependent effective evolutionary rates \citep{wertheim_purifying_2011}.
+The intra-clade rates in Fig.~\ref{fig:rate_progression} are all estimated over a similar time frame and the time dependent rate phenomenon should not apply.
+
+Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
+Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
+Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
+Epistatic interactions can further steepen the dependence on population size and increase stochasticity.
+Methods to infer emergence dates and locations of the early variants might therefore be biased or over-confident.
+
 \bibliography{bib}

### 2022-07-31 scripts/get_genotype_counts.py

@@ -17,7 +17,7 @@ if __name__=="__main__":
     parser.add_argument('--sub-clades', type=str, required=True, help="input data")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--max-date', type=float, help="input data")
-    parser.add_argument('--max-group', type=int, default=100, help="input data")
+    parser.add_argument('--max-group', type=int, help="input data")
     parser.add_argument('--query', type=str, help="filters")
     parser.add_argument('--bin-size', type=float, help="input data")
     parser.add_argument('--output-json', type=str, help="rate file")
@@ -28,7 +28,7 @@ if __name__=="__main__":
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
                                          query = args.query, max_group=args.max_group, QC_threshold=80 if args.clade=='21H' else 30,
-                                         completeness=0, swap_root=args.clade=='19B+')
+                                         completeness=0, swap_root=args.clade.startswith('19B+'))
     filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())
 
     print("clade", args.clade, "done filtering")
@@ -38,11 +38,17 @@ if __name__=="__main__":
 
     mutation_number = {}
     nmax = 5
+    nmax_extra = 15
     for n in range(nmax):
         ind = filtered_data["divergence"]==n
         mutation_number[n] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
         cumulative_sum += mutation_number[n]
     mutation_number[f'{nmax}+'] = all_sequences - cumulative_sum
+    for n in range(nmax,nmax_extra):
+        ind = filtered_data["divergence"]==n
+        mutation_number[n] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
+        cumulative_sum += mutation_number[n]
+
     print("clade", args.clade, "done mutation_number")
 
     mutation_counts = defaultdict(int)
@@ -89,7 +95,7 @@ if __name__=="__main__":
         ind = filtered_data["intra_substitutions_str"]==x
         if i<10 and nmuts:
             continue
-        if nmuts==0 or (nmuts==1 and ind.sum()>0.02*n_early) or (ind.sum()>0.05*n_early):
[…]

### 2022-07-31 scripts/plot_genotype_counts.py

@@ -21,14 +21,24 @@ if __name__=="__main__":
     with open(args.counts) as fh:
         counts = json.load(fh)
 
+    plt.figure()
+    nmuts = [m for m in counts['mutation_number'] if m[-1]!='+']
+    for week in range(5,len(counts['mutation_number']['0'])):
+        c = np.array([counts['mutation_number'][str(n)][week] for n in nmuts])
+        if c.sum()>100:
+            plt.plot(range(len(nmuts)), c/c.sum(), '-')
 
-    dates = [datetime.fromordinal(x) for x in counts['bins'][:-1]]
-    fig, axs = plt.subplots(1,3, figsize = (18,6), sharey=True)
 
-    ax = axs[0]
-    ax.plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
-    for m in counts['mutation_number']:
-        ax.plot(dates, counts['mutation_number'][m], '-o', label=f'{m} mutations')
+    dates = np.array([datetime.fromordinal(x) for x in counts['bins'][:-1]])
+    fig, axs = plt.subplots(1,3, figsize = (18,6))
+
+    ax = axs[2]
+    # ax.plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
+    all_samples = np.array(counts['all_samples'])
+    for m in ['0', '1', '2', '3', '4', '5+']:
+        muts = np.array(counts['mutation_number'][m])
+        ind = muts>0
+        ax.plot(dates[ind], muts[ind]/all_samples[ind], '-o', label=f'{m} mutations')
 
     ax.set_yscale('log')
     ax.legend()
@@ -41,7 +51,7 @@ if __name__=="__main__":
     ax.set_yscale('log')
     ax.legend()
 
-    ax = axs[2]
+    ax = axs[0]
     ax.plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
     for m in sorted(counts['genotypes'].keys(), key=lambda x: len(x)):
         ax.plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")

### 2022-07-31 scripts/root_to_tip.py

@@ -152,7 +152,7 @@ if __name__=="__main__":
     parser.add_argument('--sub-clades', type=str, required=True, help="input data")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--max-date', type=float, help="input data")
-    parser.add_argument('--max-group', type=int, default=100, help="input data")
+    parser.add_argument('--max-group', type=int, help="input data")
     parser.add_argument('--query', type=str, help="filters")
     parser.add_argument('--output-plot', type=str, help="plot file")
     parser.add_argument('--output-json', type=str, help="rate file")
@@ -163,7 +163,7 @@ if __name__=="__main__":
     d = pd.read_csv(args.metadata, sep='\t').fillna('')
     filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
                                          query = args.query, max_group=args.max_group, QC_threshold=80 if args.clade=='21H' else 30,
-                                         completeness=0, swap_root=args.clade=='19B+')
+                                         completeness=0, swap_root=args.clade.startswith('19B+'))
 
     regression = linregress(filtered_data.numdate, filtered_data.divergence)
     filtered_data["residuals"] = filtered_data.apply(lambda x: x.divergence - (regression.intercept + regression.slope*x.numdate), axis=1)

### 2022-08-01 scripts/count_mutations.py
add fitness script

@@ -1,0 +1,102 @@
+import argparse,json
+import pandas as pd
+import numpy as np
+import matplotlib as mpl
+from Bio import SeqIO
+mpl.rcParams['axes.formatter.useoffset'] = False
+from collections import defaultdict
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+columns = ['seqName', 'Nextclade_pango', "privateNucMutations.unlabeledSubstitutions", 'qc.overallStatus']
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+    parser.add_argument('--metadata', type=str, required=True, help="input data")
+    parser.add_argument('--reference', type=str, required=True, help="input data")
+    parser.add_argument('--output-fitness', type=str, required=True, help="fitness figure")
+    parser.add_argument('--output-mutation', type=str, required=True, help="fitness figure")
+    args = parser.parse_args()
+
+    ref = SeqIO.read(args.reference, 'genbank')
+
+    d = pd.read_csv(args.metadata, sep='\t', usecols=columns).fillna('')
+    d = d.loc[d["qc.overallStatus"]=='good',:]
+
+    mutation_counter = defaultdict(lambda: defaultdict(int))
+
+    for r, row in d.iterrows():
+        pango = row.Nextclade_pango
+        if pango[0]=='X':
+            continue
+
+        muts = row["privateNucMutations.unlabeledSubstitutions"]
+        if muts:
+            for m in muts.split(','):
+                mutation_counter[pango][m] += 1
+
+
+    position_counter = defaultdict(lambda: defaultdict(list))
+
+    for pmuts in mutation_counter.values():
+        for mut,c in pmuts.items():
+            a, pos, d = mut[0], int(mut[1:-1])-1, mut[-1]
+            position_counter[pos][(a,d)].append(c)
+
+
+    cutoff=1
+    total_events = np.array([sum([len([y for y in x if y>cutoff]) for x in position_counter[pos].values()])
+                        if pos in position_counter else 0 for pos in range(len(ref))])
+
+    total_counts = np.array([sum([sum(x) for x in position_counter[pos].values()])
+                        if pos in position_counter else 0 for pos in range(len(ref))])
+
[…]

### 2022-08-01 scripts/plot_af.py

@@ -8,6 +8,24 @@ import matplotlib.pyplot as plt
 import seaborn as sns
 
 
+def fit_exp(n,k,t, order=0):
+    from scipy.optimize import minimize
+    eps = 1e-16
+    def binom(x,n,k,t, order=0):
+        p = x[0]**2*np.exp(-x[1]**2*t)
+        if order>0:
+            p *= x[1]**2*t
+        if order>1:
+            p *= x[1]**2*t/2
+        p = np.minimum(1.0,p)
+        return -np.sum((n-k)*np.log(1-p+eps) + k*np.log(p+eps))
+
+    sol = minimize(binom, (1,0.05), args=(n,k,t, order), method="Powell")
+
+    return {'offset':sol['x'][0]**2, 'rate':sol['x'][1]**2}
+
+
+
 if __name__=="__main__":
     parser = argparse.ArgumentParser(
         description="remove time info",
@@ -17,18 +35,17 @@ if __name__=="__main__":
     parser.add_argument('--counts', nargs='+', type=str, required=True, help="input data")
     parser.add_argument('--output-plot', type=str, help="figure file")
     args = parser.parse_args()
+    counts = {}
+    for fi,fname in enumerate(args.counts):
+        clade = fname.split('/')[-1].split('_')[0]
+        with open(fname) as fh:
+            counts[clade] = json.load(fh)
 
     fig, axs = plt.subplots(1,1, figsize = (8,6))
     ax = axs
     ls = ['-', '-.', '--']
-    for fi,fname in enumerate(args.counts):
-        with open(fname) as fh:
-            counts = json.load(fh)
-
-        clade = fname.split('/')[-1].split('_')[0]
-        dates = [datetime.fromordinal(x) for x in counts['bins'][:-1]]
-
-        ax.plot(sorted(counts["mutation_spectrum"].values()), np.linspace(1,0,len(counts["mutation_spectrum"])+1)[:-1],
+    for fi,(clade,d) in enumerate(counts.items()):
+        ax.plot(sorted(d["mutation_spectrum"].values()), np.linspace(1,0,len(d["mutation_spectrum"])+1)[:-1],
                        label=clade, ls=ls[fi//10])
 
     ax.set_yscale('log')
@@ -39,4 +56,39 @@ if __name__=="__main__":
     ax.plot([1e-5,1e-1], [1e-0,1e-4], c='k', lw=3, alpha=0.5, label='1/x')
[…]

### 2022-08-01 scripts/root_to_tip.py

@@ -186,9 +186,9 @@ if __name__=="__main__":
     sns.histplot(x=filtered_data.numdate, y=np.minimum(ymax*1.5, filtered_data.divergence), bins=bins, ax=axs[0])
     x = np.linspace(*axs[0].get_xlim(),101)
     axs[0].set_title('all differences')
-    axs[0].plot(x, regression.intercept + regression.slope*x + tolerance(x), lw=4)
     axs[0].plot(x, regression_clean["intercept"] + regression_clean["slope"]*x, lw=4, label=f"slope = {regression_clean['slope']:1.1f} subs/year")
     axs[0].errorbar(regression_clean["date"], regression_clean["mean"], regression_clean["stderr"])
+    axs[0].plot(x, regression.intercept + regression.slope*x + tolerance(x), lw=4)
 
     axs[1].set_title('amino acid differences')
     sns.histplot(x=filtered_data.numdate[ind], y=np.minimum(ymax*1.5, filtered_data.aaDivergence[ind]), bins=bins, ax=axs[1])

### 2022-08-01 scripts/plot_af.py
add poisson fits

@@ -8,21 +8,29 @@ import matplotlib.pyplot as plt
 import seaborn as sns
 
 
-def fit_exp(n,k,t, order=0):
+def fit_exp(n,k_list,t):
     from scipy.optimize import minimize
     eps = 1e-16
-    def binom(x,n,k,t, order=0):
-        p = x[0]**2*np.exp(-x[1]**2*t)
-        if order>0:
-            p *= x[1]**2*t
-        if order>1:
-            p *= x[1]**2*t/2
-        p = np.minimum(1.0,p)
-        return -np.sum((n-k)*np.log(1-p+eps) + k*np.log(p+eps))
+    def binom(x,n,k_list,t):
+        res = 0
+        tu = np.maximum(0,x[1]**2*(t-x[0]))
+        for order, k in enumerate(k_list):
+            p = np.exp(-tu)
+            if order>0:
+                p *= tu
+            if order>1:
+                p *= tu/2
+            if order>2:
+                p *= tu/3
+            if order>3:
+                p *= tu/4
+            p = np.minimum(1.0,p)
+            res -= np.sum((n-k)*np.log(1-p+eps) + k*np.log(p+eps))
+        return res
 
-    sol = minimize(binom, (1,0.05), args=(n,k,t, order), method="Powell")
+    sol = minimize(binom, (-10,0.25), args=(n,k_list,t), method="Powell")
 
-    return {'offset':sol['x'][0]**2, 'rate':sol['x'][1]**2}
+    return {'offset':sol['x'][0], 'rate':sol['x'][1]**2}
 
 
 
@@ -56,39 +64,42 @@ if __name__=="__main__":
     ax.plot([1e-5,1e-1], [1e-0,1e-4], c='k', lw=3, alpha=0.5, label='1/x')
     # ax.plot(x, np.log(x)/np.log(x[0]), c='k', lw=3, alpha=0.5, label='~log(x)')
     plt.legend(ncol=3)
-    # plt.savefig(args.output_plot)
+    plt.savefig(args.output_plot)
 
 
     fig, axs = plt.subplots(1,1, figsize = (8,6))
     ax = axs
     ls = ['-', '-.', '--']
     for fi,(clade,d) in enumerate(counts.items()):
+        fig = plt.figure()
         dates =   np.array([x for x in d['bins'][:-1]])
-        dates -= dates[0]
+        datetimes =   np.array([datetime.fromordinal(x) for x in d['bins'][:-1]])
+        t0 = dates[0]
[…]

### 2022-08-01 scripts/count_mutations.py
add rates output

@@ -19,7 +19,7 @@ if __name__=="__main__":
     parser.add_argument('--metadata', type=str, required=True, help="input data")
     parser.add_argument('--reference', type=str, required=True, help="input data")
     parser.add_argument('--output-fitness', type=str, required=True, help="fitness figure")
-    parser.add_argument('--output-mutation', type=str, required=True, help="fitness figure")
+    parser.add_argument('--output-mutations', type=str, required=True, help="fitness figure")
     args = parser.parse_args()
 
     ref = SeqIO.read(args.reference, 'genbank')

### 2022-08-01 scripts/plot_af.py

@@ -5,7 +5,7 @@ import matplotlib as mpl
 mpl.rcParams['axes.formatter.useoffset'] = False
 
 import matplotlib.pyplot as plt
-import seaborn as sns
+import pandas as pd
 
 
 def fit_exp(n,k_list,t):
@@ -42,6 +42,7 @@ if __name__=="__main__":
 
     parser.add_argument('--counts', nargs='+', type=str, required=True, help="input data")
     parser.add_argument('--output-plot', type=str, help="figure file")
+    parser.add_argument('--output-rates', type=str, help="table")
     args = parser.parse_args()
     counts = {}
     for fi,fname in enumerate(args.counts):
@@ -67,9 +68,8 @@ if __name__=="__main__":
     plt.savefig(args.output_plot)
 
 
-    fig, axs = plt.subplots(1,1, figsize = (8,6))
-    ax = axs
     ls = ['-', '-.', '--']
+    data = []
     for fi,(clade,d) in enumerate(counts.items()):
         fig = plt.figure()
         dates =   np.array([x for x in d['bins'][:-1]])
@@ -92,14 +92,17 @@ if __name__=="__main__":
         plt.plot(datetimes[ind], klist[4][ind]/total[ind], 'o', c='C4')
 
         tu = (dates[ind]-res['offset'])*res['rate']
-        plt.plot(datetimes[ind], np.exp(-tu), label=clade, ls=ls[fi//10], c='C0')
-        plt.plot(datetimes[ind], np.exp(-tu)*tu ,label=clade, ls=ls[fi//10], c='C1')
-        plt.plot(datetimes[ind], np.exp(-tu)*tu**2/2, label=clade, ls=ls[fi//10], c='C2')
-        plt.plot(datetimes[ind], np.exp(-tu)*tu**3/6 ,label=clade, ls=ls[fi//10], c='C3')
-        plt.plot(datetimes[ind], np.exp(-tu)*tu**4/24, label=clade, ls=ls[fi//10], c='C4')
+        plt.plot(datetimes[ind], np.exp(-tu), label=clade,         ls='-', c='C0')
+        plt.plot(datetimes[ind], np.exp(-tu)*tu ,label=clade,      ls='-', c='C1')
+        plt.plot(datetimes[ind], np.exp(-tu)*tu**2/2, label=clade, ls='-', c='C2')
+        plt.plot(datetimes[ind], np.exp(-tu)*tu**3/6 ,label=clade, ls='-', c='C3')
[…]

### 2022-08-02 manuscript/appendix.tex
start fitness section

\section*{Supplementary figures}

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/mutation_distribution.pdf}
    \caption{{\bf The relative rate of different mutations in SARS-CoV-2.}
    These rates are measured from rare low frequency mutations probably subject to little purifying selection.
    \label{fig:mutation_distribution}}
\end{figure*}

### 2022-08-02 manuscript/variant_rates.tex
add fitness script
start fitness section

@@
 \begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rate_summary.png}
+    \includegraphics[width=\textwidth]{figures/rate_summary.pdf}
     \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
@@
 
-\subsection*{Diversification within variants}
+\subsection*{Diversification within clades}
 The analysis of evolutionary rates above only considers the mean divergence within clades which, at least when variation is neutral, is independent of epidemiological parameters.
@@
 with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
+
+For some clades, especially those that are well sampled soon after their emergence, this Poisson models is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clades}.
+For other clades, especially when the founder sequence is barely observed and dominant subclades developed, such fits are not robust.
+This is particularly true for the early clades 19A and 19B, as well as 21J (Delta), 21K (Omicron, BA.1), and 22B (Omicron, BA.5).
+
+\begin{figure*}
+    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
+    \caption{{\bf Diversification within clades.} \label{fig:within_clades}}
+\end{figure*}
+
 These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
 In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.
+
+\section*{Mutation tolerance}
+
+Above, we found that the rate of synonymous mutations is comparable within and between clades without any striking signal of temporal variation.
+This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements).
+To assess the level of constraint on the SARS-CoV-2 genome, we made use of the ``rare mutations'' annotation by Nextclade.
+Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
+For each pango lineage (as determined by Nextclade), we could how often these ``rare mutations'' are observed, normalize this against the relative rate of mutation of the reference nucleotide (see Supp.~Fig.~?), and use this as a proxy for mutational tolerance.
+Simply splitting mutations into those that occur in 1st, 2nd, and 3rd positions of codons already reveals strong signatures of purifying selection.
+
+ \begin{figure}
+    \includegraphics[width=\textwidth]{figures/fitness_cost.pdf}
+    \caption{{\bf Constraints on SARS-CoV-2 mutations.}
+    Almost all 3rd codon positions tolerate mutations, while 1st and 2nd positions are strongly constrained.
+    About half of the 1st and 2nd codon positions are less variable than the most constrained 10\% of 3rd positions (gray line).
+    \label{fig:fitness_costs}}
+ \end{figure}
 
@@
 \appendix
+\include{appendix}
 \include{appendix_rtt}
-\include{counts_appendix}
+\include{appendix_counts}
 \end{document}

### 2022-08-02 scripts/count_mutations.py
split fitness by gene

@@ -2,14 +2,30 @@ import argparse,json
 import pandas as pd
 import numpy as np
 import matplotlib as mpl
-from Bio import SeqIO
+from Bio import SeqIO,Seq
 mpl.rcParams['axes.formatter.useoffset'] = False
 from collections import defaultdict
+from scipy.stats import scoreatpercentile
 
 import matplotlib.pyplot as plt
 import seaborn as sns
 
 columns = ['seqName', 'Nextclade_pango', "privateNucMutations.unlabeledSubstitutions", 'qc.overallStatus']
+
+def get_sequence(mods, root_seq):
+    seq = root_seq.copy()
+    for mut in mods['nuc']:
+        a,pos,d = mut[0], int(mut[1:-1])-1, mut[-1]
+        if a!=seq[pos]:
+            print(seq[pos], mut)
+        seq[pos]=d
+    return seq
+
+def translate(codon):
+    try:
+        return Seq.translate("".join(codon))
+    except:
+        return 'X'
 
 if __name__=="__main__":
     parser = argparse.ArgumentParser(
@@ -18,19 +34,47 @@ if __name__=="__main__":
     )
     parser.add_argument('--metadata', type=str, required=True, help="input data")
     parser.add_argument('--reference', type=str, required=True, help="input data")
+    parser.add_argument('--pango-gts', type=str, required=True, help="input data")
     parser.add_argument('--output-fitness', type=str, required=True, help="fitness figure")
+    parser.add_argument('--output-fitness-by-gene', type=str, required=True, help="fitness figure")
     parser.add_argument('--output-mutations', type=str, required=True, help="fitness figure")
     args = parser.parse_args()
 
+    ## load reference
     ref = SeqIO.read(args.reference, 'genbank')
+    ref_array = np.array(ref.seq)
+    base_content = {x:ref.seq.count(x) for x in 'ACGT'}
 
+    ## make a map of codon positions, ignore orf9b (overlaps N)
+    codon_pos = np.zeros(len(ref))
+    map_to_gene = {}
+    gene_position = {}
+    gene_length = {}
+    for feat in ref.features:
+        if feat.type=='CDS':
+            gene_name = feat.qualifiers['gene'][0]
+            gene_position[gene_name] = feat.location
+            gene_length[gene_name] = (feat.location.end - feat.location.start)//3
+            if gene_name!='ORF9b':
+                for gpos, pos in enumerate(feat.location):
[…]

### 2022-08-02 scripts/get_genotype_counts.py

@@ -93,12 +93,12 @@ if __name__=="__main__":
     for x,i in intra_geno.items():
         nmuts = len(x.split(',')) if x else 0
         ind = filtered_data["intra_substitutions_str"]==x
-        if i<10 and nmuts:
+        if i<10 and nmuts or nmuts>4:
             continue
         if nmuts==0 or (nmuts==1 and ind.sum()>0.005*n_early) or (ind.sum()>0.01*n_early):
             genotypes[x] = np.histogram(filtered_data.loc[ind,"day"], bins=bins)[0]
             gt_count +=1
-            if gt_count>10:
+            if gt_count>8:
                 break
 
     print("clade", args.clade, "done genotypes")

### 2022-08-02 scripts/get_genotypes_pango.py

@@ -1,0 +1,81 @@
+import pandas as pd
+from collections import defaultdict
+import argparse,json
+
+def genotype_struct():
+    return {'nuc':{}, 'aa':defaultdict(dict)}
+
+def assign_genotypes(n):
+    gt = n["genotype"]
+    if "children" in n:
+        for c in n["children"]:
+            cgt = genotype_struct()
+
+            cgt["nuc"].update({k:v for k,v in gt["nuc"].items()})
+            if "nuc" in c["branch_attrs"]["mutations"]:
+                for mut in c["branch_attrs"]["mutations"]["nuc"]:
+                    a,pos,d = mut[0], int(mut[1:-1]), mut[-1]
+                    cgt["nuc"][pos] = d
+
+
+            for gene in set(c["branch_attrs"]["mutations"].keys()).union(set(gt["aa"].keys())):
+                if gene=='nuc':
+                    continue
+                cgt["aa"][gene].update({k:v for k,v in gt["aa"][gene].items()})
+                if gene in c["branch_attrs"]["mutations"]:
+                    for mut in c["branch_attrs"]["mutations"][gene]:
+                        a,pos,d = mut[0], int(mut[1:-1]), mut[-1]
+                        cgt['aa'][gene][pos] = d
+            c["genotype"] = cgt
+            assign_genotypes(c)
+
+def get_pango_genotypes(n, pango_gts):
+    name = n["name"]
+    if "children" in n and len(n["children"]):
+        for c in n["children"]:
+            get_pango_genotypes(c, pango_gts)
+    elif (name in ['A', 'B'] or '.' in name) and not ('/' in name):
+        pango_gts[n["name"]] = n["genotype"]
+
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="get variant genotypes",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+
+    parser.add_argument('--tree', type=str, required=True, help="input data")
+    parser.add_argument('--root', type=str, required=True, help="input data")
+    parser.add_argument('--output', type=str, required=True, help="output data")
+    args = parser.parse_args()
+
+    with open(args.tree) as fh:
+        tree = json.load(fh)['tree']
+
+    with open(args.root) as fh:
+        root_sequence = json.load(fh)
+
+    tree["genotype"] = genotype_struct()
[…]

### 2022-08-02 scripts/plot_af.py

@@ -7,31 +7,7 @@ mpl.rcParams['axes.formatter.useoffset'] = False
 import matplotlib.pyplot as plt
 import pandas as pd
 
-
-def fit_exp(n,k_list,t):
-    from scipy.optimize import minimize
-    eps = 1e-16
-    def binom(x,n,k_list,t):
-        res = 0
-        tu = np.maximum(0,x[1]**2*(t-x[0]))
-        for order, k in enumerate(k_list):
-            p = np.exp(-tu)
-            if order>0:
-                p *= tu
-            if order>1:
-                p *= tu/2
-            if order>2:
-                p *= tu/3
-            if order>3:
-                p *= tu/4
-            p = np.minimum(1.0,p)
-            res -= np.sum((n-k)*np.log(1-p+eps) + k*np.log(p+eps))
-        return res
-
-    sol = minimize(binom, (-10,0.25), args=(n,k_list,t), method="Powell")
-
-    return {'offset':sol['x'][0], 'rate':sol['x'][1]**2}
-
+from plot_genotype_counts import fit_poisson
 
 
 if __name__=="__main__":
@@ -71,38 +47,20 @@ if __name__=="__main__":
     ls = ['-', '-.', '--']
     data = []
     for fi,(clade,d) in enumerate(counts.items()):
-        fig = plt.figure()
         dates =   np.array([x for x in d['bins'][:-1]])
         datetimes =   np.array([datetime.fromordinal(x) for x in d['bins'][:-1]])
         t0 = dates[0]
         dates -= t0
         total =   np.array(d["all_samples"])
-        klist = [ np.array(d["mutation_number"]['0']),
-                  np.array(d["mutation_number"]['1']),
-                  np.array(d["mutation_number"]['2']),
-                  np.array(d["mutation_number"]['3']),
-                  np.array(d["mutation_number"]['4'])]
+        k_vecs = { i: np.array(d["mutation_number"][f'{i}']) for i in ['0', '1', '2', '3', '4', '5+']}
+
         ind = total>0
-        res = fit_exp(total[ind], [x[ind] for x in klist], dates[ind])
-        print(clade, res)
-        plt.plot(datetimes[ind], klist[0][ind]/total[ind], 'o',c='C0')
[…]

### 2022-08-02 scripts/plot_genotype_counts.py

@@ -1,11 +1,48 @@
 import argparse,json
 from datetime import datetime
+from sys import exec_prefix
 import numpy as np
 import matplotlib as mpl
 mpl.rcParams['axes.formatter.useoffset'] = False
 
 import matplotlib.pyplot as plt
 import seaborn as sns
+
+def poisson(lam, order):
+    p = np.exp(-lam)
+    try:
+        for i in range(0,int(order)):
+            p *= lam/(i+1)
+    except:
+        p_cum = np.copy(p)
+        for i in range(0,int(order[:-1])-1):
+            p *= lam/(i+1)
+            p_cum += p
+        p = 1 - p_cum
+
+    return np.maximum(0.0, np.minimum(1.0,p))
+
+def fit_poisson(n,k_vecs,t, rate=None):
+    from scipy.optimize import minimize
+    eps = 1e-16
+    def binom(x,n,k_vecs,t, rate):
+        res = 0
+        mu = rate or x[1]**2
+        tu = np.maximum(0,mu*(t-x[0]))
+        for order, k in k_vecs.items():
+            p = poisson(tu, order)
+            res -= np.sum(((n-k)*np.log(1-p+eps) + k*np.log(p+eps)))
+
+        return res
+
+    if rate is None:
+        sol = minimize(binom, (-10,0.25), args=(n,k_vecs,t,rate), method="Powell")
+        res = {'offset':sol['x'][0], 'rate':sol['x'][1]**2}
+    else:
+        sol = minimize(binom, (-10,), args=(n,k_vecs,t,rate), method="Powell")
+        res = {'offset':sol['x'][0], 'rate':rate}
+
+    return res
 
 
 if __name__=="__main__":
@@ -30,21 +67,26 @@ if __name__=="__main__":
 
 
     dates = np.array([datetime.fromordinal(x) for x in counts['bins'][:-1]])
+    t0 = counts['bins'][0]
+    rel_date = np.array([x-t0 for x in counts['bins'][:-1]])
     fig, axs = plt.subplots(1,3, figsize = (18,6))
 
     ax = axs[2]
     # ax.plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
-    all_samples = np.array(counts['all_samples'])
-    for m in ['0', '1', '2', '3', '4', '5+']:
-        muts = np.array(counts['mutation_number'][m])
[…]

### 2022-08-06 manuscript/appendix.tex
adapt cutoff dates

@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/mutation_distribution.pdf}
+\setcounter{figure}{0}
+\renewcommand{\figurename}{Figure S}
+\setcounter{table}{0}
+\renewcommand{\tablename}{Table S}
+
+\begin{figure}[h]
+    \includegraphics[width=0.5\textwidth]{figures/mutation_distribution.pdf}
     \caption{{\bf The relative rate of different mutations in SARS-CoV-2.}
@@
     \label{fig:mutation_distribution}}
-\end{figure*}
+\end{figure}
+
+\begin{figure}[h]
+    \includegraphics[width=0.5\textwidth]{figures/fitness_cost_by_gene.pdf}
+    \caption{{\bf Constraints on SARS-CoV-2 mutations by gene.}
+    The top panel quantifies constraint at 1st and 2nd positions in codons of open reading frames.
+    The bottom panel show the analogous distributions at 3rd codon positions.
+    The latter distributions are very similar across genes, with only \emph{E} showing somewhat less variation.
+    In contrast, mutation tolerance at 1st and 2nd positions differs markedly between genes.
+    In ORF3a, ORF6, ORF7a, ORF7b, and ORF8 the distribution of mutations at 1st and 2nd positions is very similar to the distribution at 3rd positions, while the remaining genes show clear signs of constraint.
+    \label{fig:fitness_costs_supp}}
+ \end{figure}

### 2022-08-06 manuscript/variant_rates.tex
adapt cutoff dates

@@
 In contrast to influenza virus evolution, the evolution of SARS-CoV-2 has been characterized by unusually large jumps in sequence space.
-New variants with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, most dramatically with the emergence of Omicron in later 2021 \citep{viana_rapid_2022}.
+New variants with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
 The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are frequently observed in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
 Onward transmission from such chronic infections has also been documented \citep{gonzalez-reiche_intrahost_2022}.
-However, to date there is no direct evidence for the emergence of any variant and the degree to which chronic infections might have contributed is a matter of debate.
+However, to date there is no direct evidence for the mode of emergence of any variant.
 The case for chronic infection being an important contributor is strongest for the variants Alpha and Omicron
@@
 
+The saltatory nature of SARS-CoV-2 evolution implies that evolution within individual variants, clades, or lineages might have different characteristics from its overall pattern, which is dominated by atypical bursts of evolution leading to new variants.
 Here, we investigate the rate at which the SARS-CoV-2 genome changes within variants and compare this rate to the rate at which the variants themselves accumulate differences relative to reference sequence.
 This comparison reveals a clear dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
+Early variants display more rapid non-synonymous evolution that later variants.
 The rate of synonymous changes between and within variants agrees, while the non-synonymous rate within variants is far below the between variant rate.
-Early variants display more rapid non-synonymous evolution that later variants.
+We further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.
 
@@
 Traditionally, evolutionary rates and divergence times are estimates using phylogenetic frameworks for heterochronously sampled sequences.
-The volume and heterogeneity of SARS-CoV-2 data, however, mean that one has to dramatically down-sample the data and remove many low quality sequences that would otherwise fatally distort the analysis.
+These methods, however, can not handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
+Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any problematic sequence can fatally distort the analysis, in particular if placed early in tree.
 
 To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and robust procedures to analyze the evolutionary patterns.
-We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain \citep{hadfield_nextstrain_2018} clades and analyze each clade separately.
+We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018} and analyze each clade separately.
+Sequences belonging to recombinant lineages are excluded.
 These clades represent well defined groups of sequences with little non-vertical evolution within them.
 Furthermore, we define a ``founder'' genotype for each clade and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
+This founder sequence is manually curated for each clade considered.
 This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with reference sequence.
-For this reduced set of sequences, we determine the mutations they carry on top of the founder genotype.
+For this reduced set of sequences, we determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
 The latter step is done for nucleotide changes as well as for amino acid changes.
@@
 Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
-To further exclude outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the variance by 3.
+For the majority of sequences, this is approximately true, but some problematic sequences are many more mutations than expected.
+To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the variance by 3.
 
@@
 Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each bin is weighted with the fourth root of the number of sequences in the bins.
-The exact functional form of this weighing does not have a big influence on the results, but a strongly sublinear weighing helps to counter the large variation in sequencing effort and the natural imbalance due to the fact that few sequences are available early.
+The exact functional form of this weighing does not have a big influence on the results, but a sublinear weighing helps to counter the large variation in sequencing effort across countries and the natural imbalance due to the fact that few sequences are available early.
 
-Due to shared ancestry, a divergence (or root-to-tip) regression against time is not a suitable method to determine the evolutionary rate.
-However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from a basal polytomy.
-Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate.
+Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suited to estimate evolutionary rates in general.
+In particular, confidence intervals are difficult to obtain.
+However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from one or several basal polytomies.
+Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate robustly.
 Fig.~\ref{fig:rate_alpha} shows the increasing intra-clade divergence for clade 20I corresponding to VoC Alpha.
@@
 Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
-Analogous figures for all variants are included in the appendix.
+Analogous figures for all clades considered are included in the appendix.
     \label{fig:rate_alpha}}
@@
 For almost all major Nextstrain clades, the average divergence increases linearly in time (similar patterns as Fig.~\ref{fig:rate_alpha}) allowing us to estimate clade specific evolutionary rates for amino acid and silent changes.
-These rates are summarized in Fig.~\ref{fig:rate_summary}.
+These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
 
@@
 
-Rates of synonymous change are very consistent across variants and also agree with the overall rate of synonymous changes of about 5-8 changes per genome per year.
-Rates of non-synonymous changes are much more variable.
+Rates of synonymous change are very consistent across variants and also agree with the overall rate of synonymous changes of about 5-8 changes per genome per year, while rates of non-synonymous changes are much more variable (Fig.~\ref{fig:rate_summary}D).
 Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
-Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year.
+Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year Fig.~\ref{fig:rate_progression}.
 In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.
@@
 We thus also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
-While we estimate somewhat higher rates for these composite clades (in particular 19B++), the overall pattern remains consistent.
+The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
+The latter is due to the rapid expansion and subsequent dominance of clade 20A and its descendants \citep{korber_tracking_2020}.
 
@@
     \hline
-      19A &     19.21 &   11.85 &     7.36 &     0 &       0 &        0 \\
-      19B &     24.40 &   14.50 &     9.90 &     2 &       1 &        1 \\
-      20A &     20.79 &   11.44 &     9.35 &     4 &       2 &        2 \\
-      20B &     14.64 &    8.42 &     6.22 &     7 &       4 &        3 \\
-      20C &     19.22 &   11.64 &     7.57 &     6 &       4 &        2 \\
-     20A+ &     18.48 &   10.26 &     8.22 &     4 &       2 &        2 \\
-      20E &     13.77 &    6.04 &     7.73 &    11 &       4 &        7 \\
-      20H &     11.16 &    5.22 &     5.94 &    19 &      15 &        4 \\
-      20I &     11.65 &    5.77 &     5.88 &    28 &      18 &       10 \\
-      20J &     10.93 &    2.66 &     8.28 &    33 &      21 &       12 \\
-      21I &     16.53 &    8.88 &     7.65 &    28 &      23 &        5 \\
-      21J &     16.20 &    8.19 &     8.00 &    32 &      27 &        5 \\
-      21K &     12.74 &    5.41 &     7.34 &    54 &      45 &        9 \\
-      21L &      8.66 &    4.67 &     3.99 &    66 &      49 &       17 \\
-      22A &      8.46 &    2.92 &     5.55 &    69 &      51 &       18 \\
-      22B &     13.78 &    4.13 &     9.65 &    64 &      49 &       15 \\
+    19A &     19.76 &    12.79 &      6.97 &        0 &       0 &        0 \\
+    19B &     19.47 &    12.37 &      7.10 &        2 &       1 &        1 \\
+   19B+ &     18.55 &    11.90 &      6.65 &        2 &       1 &        1 \\
+  19B++ &     28.21 &    16.37 &     11.84 &        2 &       1 &        1 \\
+    20A &     15.66 &     8.40 &      7.25 &        4 &       2 &        2 \\
+    20B &     14.14 &     7.61 &      6.54 &        7 &       4 &        3 \\
+    20C &     16.30 &     9.52 &      6.78 &        6 &       4 &        2 \\
+   20A+ &     17.15 &     9.81 &      7.33 &        4 &       2 &        2 \\
+    20E &     13.65 &     5.74 &      7.91 &       11 &       4 &        7 \\
+    20H &     14.50 &     8.03 &      6.47 &       19 &      15 &        4 \\
+    20I &     12.18 &     5.85 &      6.33 &       28 &      18 &       10 \\
+    20J &     13.31 &     6.84 &      6.47 &       33 &      21 &       12 \\
+    21D &     11.36 &     6.29 &      5.07 &       25 &      12 &       13 \\
+    21G &     12.05 &     6.52 &      5.53 &       26 &      18 &        8 \\
+    21H &     18.17 &    10.00 &      8.17 &       29 &      20 &        9 \\
+    21I &     16.33 &     8.67 &      7.66 &       28 &      23 &        5 \\
+    21J &     12.25 &     6.32 &      5.93 &       32 &      27 &        5 \\
+    21K &     14.37 &     6.70 &      7.67 &       54 &      45 &        9 \\
+    21L &      9.02 &     3.80 &      5.22 &       66 &      49 &       17 \\
+    22A &      8.10 &     2.79 &      5.31 &       69 &      51 &       18 \\
+    22B &     12.07 &     4.97 &      7.11 &       64 &      49 &       15 \\
       \hline
     \end{tabular}
-\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.} \label{tab:rates}}
+\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
+    The last three columns give the distances of the clade founder sequence from the reference Wuhan/Hu-1.
+\label{tab:rates}}
 \end{table*}
@@
 
-The apparent time dependence of non-synonymous rate suggests that at the beginning a large fraction of amino acid changes were beneficial to viral replication, not just the D614G mutation in the spike protein \citep{korber_tracking_2020}.
-This applies both to the changes within clades that were ultimately displaced by more successful variants, and inter-variant changes where the fraction of adaptive changes is even higher.
+% The apparently higher non-synonymous rate suggests that at the beginning a large fraction of amino acid changes were beneficial to viral replication, not just the D614G mutation in the spike protein \citep{korber_tracking_2020}.
+% This applies both to the changes within clades that were ultimately displaced by more successful variants, and inter-variant changes where the fraction of adaptive changes is even higher.
 
 \subsection*{Diversification within clades}
-The analysis of evolutionary rates above only considers the mean divergence within clades which, at least when variation is neutral, is independent of epidemiological parameters.
-The structure of within clade diversity does contain information on the rate of spread.
 A simple model for a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
@@
 
-For some clades, especially those that are well sampled soon after their emergence, this Poisson models is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clades}.
+For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clades}.
 For other clades, especially when the founder sequence is barely observed and dominant subclades developed, such fits are not robust.
@@
 Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
-For each pango lineage (as determined by Nextclade), we could how often these ``rare mutations'' are observed, normalize this against the relative rate of mutation of the reference nucleotide (see Supp.~Fig.~?), and use this as a proxy for mutational tolerance.
-Simply splitting mutations into those that occur in 1st, 2nd, and 3rd positions of codons already reveals strong signatures of purifying selection.
+For each pango lineage (as determined by Nextclade), we could how often these ``rare mutations'' are observed, normalize this against the relative rate of mutation of the reference nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
+
+Simply splitting mutations into those that occur in 1st, 2nd, and 3rd positions of codons already reveals strong signatures of purifying selection, see Fig.~\ref{fig:fitness_costs}.
+Between 15-20\% of positions at first and second codons show almost no variation, while half of these sites are less variable then the most constrained 10\% of third positions.
+The median of variation at 3rd position is more than double that at other 1st and 2nd positions.
+
+When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp}), the most constrained are ORF1b and M, while ORF3a, ORF6, ORF7a, ORF7b, and ORF8 show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
+N shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
+Rare variation at 3rd positions is very comparable between genes with only E showing slightly less variation at 3rd positions than other genes.
 
 \begin{figure}
-    \includegraphics[width=\textwidth]{figures/fitness_cost.pdf}
+    \includegraphics[width=0.5\textwidth]{figures/fitness_cost.pdf}
     \caption{{\bf Constraints on SARS-CoV-2 mutations.}

### 2022-08-06 scripts/count_mutations.py

@@ -176,15 +176,19 @@ if __name__=="__main__":
     plt.savefig(args.output_fitness)
 
     # figure with pdfs instead of cdfs
-    # measure = total_events_rescaled
     # plt.figure()
-    # bins = np.logspace(0, np.ceil(np.log10(measure.max())), 31)
-    # bc = np.sqrt(bins[:-1]*bins[1:])
-    # bins[0] = 0
-    # for p in range(4):
-    #     ind = codon_pos==p
-    #     y,x = np.histogram(measure[ind], bins=bins)
-    #     plt.plot(bc,y/y.sum(), label = 'non-coding' if p==0 else f"codon pos={p}")
+    measure = total_events_rescaled
+    bins = np.logspace(0, np.ceil(np.log10(measure.max())), 101)
+    bc = np.sqrt(bins[:-1]*bins[1:])
+    bins[0] = 0
+    rate_estimate = {}
+    for p in range(4):
+        ind = codon_pos==p
+        y,x = np.histogram(measure[ind], bins=bins)
+        rate_estimate[p] = {"mean": np.sum(bc*y/y.sum()),
+                            "geo-mean": np.exp(np.sum(np.log(bc)*y/y.sum())),
+                            "median": np.median(measure[ind])}
+        # plt.plot(bc,y/y.sum(), label = 'non-coding' if p==0 else f"codon pos={p}")
 
     # plt.xscale('log')
     # plt.ylabel("distribution")
@@ -206,6 +210,7 @@ if __name__=="__main__":
     fig, axs = plt.subplots(2,1, figsize = (6,10), sharex=True)
     pos = np.arange(len(ref_array))
     for i,gene in enumerate(gene_position):
+        if gene=='ORF9b': continue
         c = f"C{i}"
         ls = '--' if i>9 else '-'
         gene_range = gene_position[gene]
@@ -219,6 +224,10 @@ if __name__=="__main__":
     ind = codon_pos==0
     axs[1].plot(sorted(measure[ind]), np.linspace(0,1,ind.sum()),
                     label = 'non-coding', c='k')
+
+    for ax in axs:
+        for x in ax.xticks:
+            ax.plot([x,x], [0,1], alpha=0.3, c='k', lw=1)
 
     axs[1].legend(ncol=2)
     axs[1].set_title("3rd codon positions or non-coding")

### 2022-08-06 scripts/plot_genotype_counts.py

@@ -83,6 +83,7 @@ if __name__=="__main__":
         plt.plot(dates[ind], poisson(tu,m), ls='-', c=f'C{i}')
 
     ax.set_yscale('log')
+    ax.set_ylim(8e-5, 2)
     ax.legend()
 
     ax = axs[1]

### 2022-08-06 manuscript/variant_rates.tex
split fitness scipt

@@
 \begin{figure*}[tb]
-    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.png}
+    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.pdf}
     \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
@@
 Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
-For each pango lineage (as determined by Nextclade), we could how often these ``rare mutations'' are observed, normalize this against the relative rate of mutation of the reference nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
+For each pango lineage (as determined by Nextclade), we count how often these ``rare mutations'' are observed, normalize this against the relative rate of mutation of the reference nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
 

### 2022-08-06 scripts/count_mutations.py

@@ -35,9 +35,9 @@ if __name__=="__main__":
     parser.add_argument('--metadata', type=str, required=True, help="input data")
     parser.add_argument('--reference', type=str, required=True, help="input data")
     parser.add_argument('--pango-gts', type=str, required=True, help="input data")
-    parser.add_argument('--output-fitness', type=str, required=True, help="fitness figure")
-    parser.add_argument('--output-fitness-by-gene', type=str, required=True, help="fitness figure")
-    parser.add_argument('--output-mutations', type=str, required=True, help="fitness figure")
+    parser.add_argument('--rare-cutoff', type=int, default=2, help="minimal number of occurrences to count mutations")
+    parser.add_argument('--output-fitness', type=str, required=True, help="fitness table")
+    parser.add_argument('--output-mutations', type=str, required=True, help="fitness table")
     args = parser.parse_args()
 
     ## load reference
@@ -71,24 +71,34 @@ if __name__=="__main__":
 
     # glob all rare mutations by pango lineage
     mutation_counter = defaultdict(lambda: defaultdict(int))
-    # pango_counter = defaultdict(int)
+    pango_counter = defaultdict(int)
     for r, row in d.iterrows():
         pango = row.Nextclade_pango
-        # pango_counter[pango] += 1
         if pango[0]=='X':
             continue
 
+        pango_counter[pango] += 1
         muts = row["privateNucMutations.unlabeledSubstitutions"]
         if muts:
             for m in muts.split(','):
                 mutation_counter[pango][m] += 1
 
+    lineage_size_cutoff = 100
+    big_lineages = [pango for pango in pango_counter if pango_counter[pango]>lineage_size_cutoff]
+    nlin = len(big_lineages)
+    tmp_number_of_pango_muts = defaultdict(int)
+    for pango, muts in pango_gts.items():
+        if pango_counter[pango]>lineage_size_cutoff:
+            for m in muts['nuc']:
+                tmp_number_of_pango_muts[int(m[1:-1])-1] += 1
+    number_of_pango_muts = np.array([tmp_number_of_pango_muts.get(pos,0)/nlin for pos in range(len(ref))])
+
 
[…]

### 2022-08-06 scripts/plot_fitness.py

@@ -1,0 +1,126 @@
+import argparse,json
+import pandas as pd
+import numpy as np
+from collections import defaultdict
+import matplotlib.pyplot as plt
+from scipy.stats import scoreatpercentile
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+    parser.add_argument('--fitness', type=str, required=True, help="input data")
+    parser.add_argument('--mutations', type=str, required=True, help="input data")
+    parser.add_argument('--output-fitness', type=str, help="fitness figure")
+    parser.add_argument('--output-fitness-by-gene', type=str, help="fitness figure")
+    parser.add_argument('--output-mutations', type=str, help="fitness figure")
+    args = parser.parse_args()
+
+    mutations = pd.read_csv(args.mutations, sep='\t', index_col=0)
+    fitness = pd.read_csv(args.fitness, sep='\t', index_col=0)
+
+    ## figure with mutation distributions
+    plt.figure()
+    sorted_muts = mutations.sort_values('scaled_rate', ascending=False)
+    mut_sum = np.sum(sorted_muts.scaled_rate)
+    muts = [(t,v/mut_sum) for t,v in sorted_muts.scaled_rate.items()]
+    plt.bar(np.arange(len(muts)), height=[m[1] for m in muts])
+    plt.xticks(np.arange(len(muts)), [m[0] for m in muts], rotation=30)
+    plt.ylabel('fraction')
+    plt.savefig(args.output_mutations)
+
+    ## Figure with 1st, 2nd, 3rd positions
+    plt.figure()
+    for p in range(1,4):
+        ind = fitness.pos_in_codon==p
+        plt.plot(sorted(fitness.tolerance[ind]), np.linspace(0,1,ind.sum()),
+                        label = f"codon pos={p}")
+
+    ind = fitness.pos_in_codon.isna()
+    plt.plot(sorted(fitness.tolerance[ind]), np.linspace(0,1,ind.sum()),
+                    label = f"non-codon")
+
+    ind = fitness.pos_in_codon==3
+    syn_cutoff = scoreatpercentile(fitness.tolerance[ind],10)
[…]

### 2022-08-06 scripts/count_mutations.py
include reversions in fitness

@@ -10,7 +10,7 @@ from scipy.stats import scoreatpercentile
 import matplotlib.pyplot as plt
 import seaborn as sns
 
-columns = ['seqName', 'Nextclade_pango', "privateNucMutations.unlabeledSubstitutions", 'qc.overallStatus']
+columns = ['seqName', 'Nextclade_pango', "privateNucMutations.unlabeledSubstitutions", 'qc.overallStatus', 'privateNucMutations.reversionSubstitutions']
 
 def get_sequence(mods, root_seq):
     seq = root_seq.copy()
@@ -37,6 +37,7 @@ if __name__=="__main__":
     parser.add_argument('--pango-gts', type=str, required=True, help="input data")
     parser.add_argument('--rare-cutoff', type=int, default=2, help="minimal number of occurrences to count mutations")
     parser.add_argument('--output-fitness', type=str, required=True, help="fitness table")
+    parser.add_argument('--output-events', type=str, required=True, help="events table")
     parser.add_argument('--output-mutations', type=str, required=True, help="fitness table")
     args = parser.parse_args()
 
@@ -78,10 +79,12 @@ if __name__=="__main__":
             continue
 
         pango_counter[pango] += 1
-        muts = row["privateNucMutations.unlabeledSubstitutions"]
-        if muts:
-            for m in muts.split(','):
-                mutation_counter[pango][m] += 1
+        muts = row["privateNucMutations.unlabeledSubstitutions"].split(',')
+        rev_muts = row["privateNucMutations.reversionSubstitutions"].split(',')
+        if len(muts) + len(rev_muts):
+            for m in muts + rev_muts:
+                if m:
+                    mutation_counter[pango][m] += 1
 
     lineage_size_cutoff = 100
     big_lineages = [pango for pango in pango_counter if pango_counter[pango]>lineage_size_cutoff]
@@ -132,6 +135,19 @@ if __name__=="__main__":
     def total_len_filtered_lists(l, cutoff):
         return sum([len([y for y in x if y>=cutoff]) for x in l])
 
+    sorted_transitions = []
+    for a in 'ACGT':
+        for d in 'ACGT':
+            if a!=d:
+                sorted_transitions.append((a,d))
+
+    all_events = []
+    for pos in range(len(ref)):
+        tmp = []
+        for a, d in sorted_transitions:
[…]

### 2022-08-06 scripts/plot_fitness.py

@@ -44,9 +44,11 @@ if __name__=="__main__":
     ind = fitness.pos_in_codon==3
     syn_cutoff = scoreatpercentile(fitness.tolerance[ind],10)
     plt.plot([syn_cutoff, syn_cutoff], [0,1], c='k', alpha=0.3)
-    for i in range(3):
+    for i in range(1,3):
         ind = fitness.pos_in_codon==i
-        print(np.mean(fitness.tolerance[ind]<syn_cutoff))
+        print("codon pos", i, np.mean(fitness.tolerance[ind]<syn_cutoff))
+    ind = fitness.pos_in_codon.isna()
+    print("non coding", np.mean(fitness.tolerance[ind]<syn_cutoff))
 
     plt.xscale('log')
     plt.ylabel("fraction less")
@@ -85,7 +87,7 @@ if __name__=="__main__":
 
     ind = fitness.pos_in_codon.isna()
     axs[1].plot(sorted(fitness.tolerance[ind]), np.linspace(0,1,ind.sum()),
-                    label = 'non-coding', c='k')
+                       label = 'non-coding', c='k')
 
     plt.xscale('log')
     plt.xlim(0.01, 10)
@@ -102,25 +104,25 @@ if __name__=="__main__":
 
     # fitness_cost = np.array(fitness_cost)
     # plt.figure()
-    # pos = np.arange(len(ref_array))
-    # ws = 10
-    # w = np.ones(ws)/ws
-    # for i,gene in enumerate(gene_position):
-    #     if gene=='ORF9b': continue
-
-    #     gene_range = gene_position[gene]
-    #     plt.figure()
-    #     for cp in range(3):
-    #         #ind = (codon_pos==(cp+1)) & (pos>=gene_range.start) & (pos<gene_range.end) & (~np.isnan(fitness_cost).all(axis=1))
-    #         #plt.hist(np.min(np.log(fitness_cost[ind]), axis=1))
-    #         ind = (codon_pos==(cp+1)) & (pos>=gene_range.start) & (pos<gene_range.end)
-    #         gene_pos = (np.convolve(pos[ind],w, mode='valid') - gene_range.start)/3
-    #         plt.plot(gene_pos,
-    #                  np.convolve(np.log10(measure[ind]+.03), w, mode='valid'),
-    #                             c=f'C{cp}', label=f'codon pos {cp+1}')
[…]

### 2022-08-07 scripts/intra_host.py
add panel labels

@@ -1,0 +1,49 @@
+import argparse,glob
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+
+def get_freq(d, cov_cutoff=1000):
+    cov = d.Good_depth
+    freq = np.array([d[f"Count_{n}"]/cov for n in 'ACGT']).T
+    freq[cov<cov_cutoff] = np.nan
+    return freq
+
+def get_diversity(freq):
+    return 1-np.sum(freq**2, axis=1)
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="plot within host evolution",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+    parser.add_argument('--allele-counts-dir', type=str, required=True, help="input data")
+    parser.add_argument('--output-intra-host', type=str, help="fitness figure")
+    args = parser.parse_args()
+
+    fnames = glob.glob(args.allele_counts_dir + "/*")
+    freqs = []
+    div = []
+    for fname in fnames:
+        d = pd.read_csv(fname, sep='\t')
+        freqs.append(get_freq(d))
+        div.append(get_diversity(freqs[-1]))
+
+    masked_diversity = np.ma.array(div, mask=np.isnan(div))
+
+
+    median_div = np.ma.median(masked_diversity, axis=0)
+
+    smoothed_diversity = np.exp(np.ma.convolve(np.log(median_div.filled(fill_value=1e-5)+1e-5), np.ones(30)/30, mode='same'))
+
+    minor_variants = [[] for i in range(4)]
+    for f in freqs:
+        cons = np.ma.argmax(f, axis=1)
+        valid = ~np.isnan(np.ma.max(f, axis=1))
+        for i in range(4):
+            ind = (cons==i)&valid
+            minor_variants[i].append(np.median(f[ind,:], axis=0))
+
+    M = np.array([np.mean(minor_variants[i], axis=0) for i in range(4)])
+    np.fill_diagonal(M,0)
+

### 2022-08-07 scripts/plot_fitness_landscape.py

@@ -1,0 +1,73 @@
+import argparse
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+    parser.add_argument('--fitness', type=str, required=True, help="input data")
+    parser.add_argument('--output-fitness-landscape', type=str, help="fitness figure")
+    args = parser.parse_args()
+
+    fitness = pd.read_csv(args.fitness, sep='\t', index_col=0)
+    genes = [x for x in fitness.gene.unique() if not pd.isna(x)]
+    pos = np.arange(len(fitness))
+
+    fig = plt.figure(figsize=(15,12))
+    gene_groups = [ ['ORF3a', 'E', 'M', 'ORF6','ORF7a', 'ORF7b', 'ORF8'],['S', 'N'],['ORF1b'],['ORF1a'] ]
+    gene_length = {k:(fitness.gene==k).sum() for k in genes}
+
+    fs = 14
+    ws = 20
+    w = np.ones(ws)/ws
+    ws_fine = 7
+    w_fine = np.ones(ws_fine)/ws_fine
+    lower_fitness_cutoff = 0.03
+    n_rows = len(gene_groups)
+    h_spread = 0.003
+    h_margin = 0.05
+    v_spread = 0.02
+    v_margin = 0.05
+    height = (1 - v_spread*n_rows - v_margin)/n_rows
+    for row,group in enumerate(gene_groups):
+        axis_bottom = row*(v_spread + height) + v_margin
+        available_width = 1 - len(group)*h_spread - h_margin
+        left = h_margin
+        total_genome = np.sum([gene_length[g] for g in group])
+        for gi,gene in enumerate(group):
+            width = gene_length[gene]/total_genome*available_width
+            ax = fig.add_axes((left,axis_bottom, width, height))
+            left += width + h_spread
+            for cp in range(3):
+                ind = (fitness.pos_in_codon==(cp+1)) & (fitness.gene==gene)
+                gene_pos = (fitness.codon[ind]*3 + cp - 1)/3
+                gene_pos_smooth = np.convolve(gene_pos, w, mode='valid')
+                ax.plot(gene_pos_smooth,
+                        np.convolve(np.log10(fitness.tolerance[ind]+lower_fitness_cutoff), w, mode='valid'),
[…]

### 2022-08-07 scripts/plot_genotype_counts.py

@@ -7,7 +7,8 @@ mpl.rcParams['axes.formatter.useoffset'] = False
 
 import matplotlib.pyplot as plt
 import seaborn as sns
-
+from root_to_tip import add_panel_label
+fs=12
 def poisson(lam, order):
     p = np.exp(-lam)
     try:
@@ -82,25 +83,23 @@ if __name__=="__main__":
         ax.plot(dates[ind], k[ind]/total[ind], 'o', label=f'{m} mutations',c=f"C{i}")
         plt.plot(dates[ind], poisson(tu,m), ls='-', c=f'C{i}')
 
-    ax.set_yscale('log')
     ax.set_ylim(8e-5, 2)
-    ax.legend()
 
     ax = axs[1]
     ax.plot(dates, total, lw=3, c='k', alpha=0.3)
     for m in sorted(counts['mutations'].keys(), key=lambda x:int(x[1:-1])):
         ax.plot(dates, counts['mutations'][m], '-o', label=f'{m}')
 
-    ax.set_yscale('log')
-    ax.legend()
-
     ax = axs[0]
     ax.plot(dates, total, lw=3, c='k', alpha=0.3)
     for m in sorted(counts['genotypes'].keys(), key=lambda x: len(x)):
         ax.plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")
 
-    ax.set_yscale('log')
-    ax.legend()
     fig.autofmt_xdate()
+    for ax, label in zip(axs, 'DEF'):
+        ax.set_yscale('log')
+        ax.legend()
+        add_panel_label(ax, label, fs=fs*1.8)
+
 
     plt.savefig(args.output_plot)

### 2022-08-07 scripts/root_to_tip.py

@@ -10,6 +10,7 @@ import seaborn as sns
 from collections import defaultdict
 mpl.rcParams['axes.formatter.useoffset'] = False
 
+fs=14
 date_reference = datetime(2020,1,1).toordinal()
 def date_to_week_since2020(d):
     return (d.toordinal() - date_reference)//7
@@ -140,6 +141,9 @@ def get_clade_gts(all_gts, subclade_str):
 
     return clade_gt
 
+def add_panel_label(ax, t, fs=16):
+    ax.text(-0.1,1,t, fontsize=fs,transform=ax.transAxes)
+
 if __name__=="__main__":
     parser = argparse.ArgumentParser(
         description="remove time info",
@@ -185,34 +189,42 @@ if __name__=="__main__":
     bins = bins=(20,np.arange(-0.5,ymax+0.5))
     sns.histplot(x=filtered_data.numdate, y=np.minimum(ymax*1.5, filtered_data.divergence), bins=bins, ax=axs[0])
     x = np.linspace(*axs[0].get_xlim(),101)
-    axs[0].set_title('all differences')
+    axs[0].set_title(f'all differences', fontsize=fs*1.2)
     axs[0].plot(x, regression_clean["intercept"] + regression_clean["slope"]*x, lw=4, label=f"slope = {regression_clean['slope']:1.1f} subs/year")
     axs[0].errorbar(regression_clean["date"], regression_clean["mean"], regression_clean["stderr"])
     axs[0].plot(x, regression.intercept + regression.slope*x + tolerance(x), lw=4)
 
-    axs[1].set_title('amino acid differences')
+    axs[1].set_title(f'amino acid differences', fontsize=fs*1.2)
     sns.histplot(x=filtered_data.numdate[ind], y=np.minimum(ymax*1.5, filtered_data.aaDivergence[ind]), bins=bins, ax=axs[1])
     axs[1].plot(x, regression_clean_aa["intercept"] + regression_clean_aa["slope"]*x, lw=4, label=f"slope = {regression_clean_aa['slope']:1.1f} subs/year")
     axs[1].errorbar(regression_clean_aa["date"], regression_clean_aa["mean"], regression_clean_aa["stderr"])
 
-    axs[2].set_title('synonymous differences')
+    axs[2].set_title(f'synonymous differences', fontsize=fs*1.2)
[…]

### 2022-08-07 manuscript/appendix.tex
edits

@@
  \end{figure}
+
+ \begin{figure}[tb]
+    \includegraphics[width=\textwidth]{figures/fitness_landscape.pdf}
+    \caption{{\bf Landscape of selective constraint along the SARS-CoV-2 genome.}
+    Solid lines show sliding window smoothing of estimated mutational tolerance at 1st, 2nd, and 3rd positions for a window size of 20 and 7 (faint lines) sites.
+    The markers below the lines are a qualitative indication of variable positions across SARS-CoV-2 diversity.
+    \label{fig:fitness_landscape_supp}}
+ \end{figure}

### 2022-08-07 manuscript/variant_rates.tex
edits

@@
 
-Since its emergence in late 2019, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large unobserved jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
+Since its emergence in late 2019, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants without observed intermediates \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
 Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
-Variants of Concern or Interest were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
+Such variants of Concern or Interest were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
 The branches leading to these variants are characterized by many amino acid changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.
+
 This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells in common among many RNA viruses and for example well studied in influenza virus A evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
-
-In contrast to influenza virus evolution, the evolution of SARS-CoV-2 has been characterized by unusually large jumps in sequence space.
-New variants with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
+Adaptive evolution in influenza tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
 The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are frequently observed in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
@@
 However, to date there is no direct evidence for the mode of emergence of any variant.
-The case for chronic infection being an important contributor is strongest for the variants Alpha and Omicron
-\citep{hill_origins_2022}.
+The case for chronic infection being an important contributor is strongest for the variants Alpha and Omicron \citep{hill_origins_2022}.
 
 The saltatory nature of SARS-CoV-2 evolution implies that evolution within individual variants, clades, or lineages might have different characteristics from its overall pattern, which is dominated by atypical bursts of evolution leading to new variants.
-Here, we investigate the rate at which the SARS-CoV-2 genome changes within variants and compare this rate to the rate at which the variants themselves accumulate differences relative to reference sequence.
-This comparison reveals a clear dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
+\citet{tay_emergence_2022} and \citet{hill_origins_2022} show that the rate of evolution along these inter-clade branches is up to four-fold higher than the background rate.
+Here, we build on these results and investigate the patterns of SARS-CoV-2 diversifications within variants and compare these to the global dynamics of evolution and adaptation.
+This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
+The rate of synonymous evolution does not differ between and within variants, while the non-synonymous rate within variants is far below the between variant rate.
 Early variants display more rapid non-synonymous evolution that later variants.
-The rate of synonymous changes between and within variants agrees, while the non-synonymous rate within variants is far below the between variant rate.
 We further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.
@@
 
-To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and robust procedures to analyze the evolutionary patterns.
+To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns.
 We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018} and analyze each clade separately.
@@
 For the majority of sequences, this is approximately true, but some problematic sequences are many more mutations than expected.
-To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the variance by 3.
+To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected variance by 3.
 
@@
 Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
-For each pango lineage (as determined by Nextclade), we count how often these ``rare mutations'' are observed, normalize this against the relative rate of mutation of the reference nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
+For each pango lineage (as determined by Nextclade), we count how often these ``rare mutations'' (including reversions to reference) are observed, normalize this against the relative rate of mutation away from the ancestral nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
 
@@
 
-When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp}), the most constrained are ORF1b and M, while ORF3a, ORF6, ORF7a, ORF7b, and ORF8 show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
+When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape_supp}), the most constrained are ORF1b and M, while ORF3a, ORF6, ORF7a, ORF7b, and ORF8 show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
 N shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
-Rare variation at 3rd positions is very comparable between genes with only E showing slightly less variation at 3rd positions than other genes.
+Rare variation at 3rd positions is very comparable between genes.
+Only E shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35.
 

### 2022-08-07 manuscript/appendix_rtt.tex
adapt cutoff dates
split fitness scipt
move legend, change captions

@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/19A_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/19A_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/19A_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 19A.}
+    Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
+    The green line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter.
+Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
+Analogous figures for all clades considered are included in the appendix.
+    Bottom: Panels D and E show the prevalence of specific genotypes (D) and specific mutations (E).
+In the case shown, the founder genotype initially dominates and no daughter genotype or mutation dominate.
+Panel F shows a Poisson model to the breakdown of the population into genotypes with different number of mutations over time.
     \label{fig:19A_divergence}}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/19B_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/19B_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/19B_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 19B.}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/19B+_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/19B+_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/19B+_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 19B+.}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/20A_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/19B++_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/19B++_counts.pdf}
+    \caption{{\bf Divergence increases linearly with time in clade 19B++.}
+    This figure contains sequences in clades 19A, 19B, 20A, 20B, 20C, and 20D rooted on clade 19B.
+    \label{fig:19B++_divergence}}
+\end{figure*}
+
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/20A_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/20A_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 20A.}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/20B_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/20B_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/20B_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 20B.}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/20C_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/20C_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/20C_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 20C.}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/20A+_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/20A+_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/20A+_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 20A+.}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/20E_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/20E_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/20E_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 20E.}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/20H_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/20H_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/20H_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 20H (Beta).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/20I_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/20I_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 20I (Alpha).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/20J_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/20J_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/20J_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 20J (Gamma).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/21D_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/21D_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/21D_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 21D (Eta).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/21G_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/21G_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/21H_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 21G (Lambda).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/21H_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/21H_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/21H_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 21H (Mu).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/21I_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/21I_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/21I_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 21I (Delta).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/21J_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/21J_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/21J_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 21J (Delta).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/21K_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/21K_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/21K_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 21K (Omicron).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/21L_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/21L_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/21L_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 21L (Omicron).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/22A_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/22A_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/22A_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 22A (Omicron).}
@@
 
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rtt/22B_rtt.png}
+\begin{figure*}[h]
+    \includegraphics[width=0.8\textwidth]{figures/rtt/22B_rtt.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/counts/22B_counts.pdf}
     \caption{{\bf Divergence increases linearly with time in clade 22B (Omicron).}

### 2022-08-07 scripts/combine_fits.py

@@ -48,11 +48,12 @@ if __name__=="__main__":
         ax.fill_between(x, y+std_dev, np.maximum(0, y-std_dev), fc='k', lw=3, alpha=0.1, ec=None)
         ax.set_xlim(x.min(), x.max())
         ax.set_ylim(0)
-        if mut_type == 'syn': ax.legend(ncol=2)
+        if mut_type == 'aa': ax.legend(ncol=2)
     for i, (mut_type, rate) in enumerate(inter_clade_rates.items()):
         axs[-1,-1].plot([i-0.4, i+0.4], [rate, rate], lw=3, c='k', alpha=0.5)
         clade_rates = rates[f"{mut_type}_rate"]
-        axs[-1,-1].scatter(i - 0.35 + np.random.random(size=len(clade_rates))*0.7, clade_rates, c=[f"C{ci%10}" for ci in range(len(clade_rates))])
+        axs[-1,-1].scatter(i - 0.35 + np.random.random(size=len(clade_rates))*0.7,
+                           clade_rates, c=[f"C{ci%10}" for ci in range(len(clade_rates))])
     axs[-1,-1].set_ylabel("substitutions per year")
     axs[-1,-1].set_xticks([0,1,2], ['nuc', 'aa', 'syn'])
     axs[-1,-1].set_ylim(0)

### 2022-08-07 scripts/root_to_tip.py

@@ -219,12 +219,19 @@ if __name__=="__main__":
         plt.show()
 
     aaCounter = defaultdict(int)
+    nucCounter = defaultdict(int)
     for subs in filtered_data.intra_aaSubstitutions:
         for a in subs: aaCounter[a]+=1
+    for subs in filtered_data.intra_substitutions:
+        for a in subs: nucCounter[a]+=1
 
-    top_subs = m = sorted(aaCounter.items(),key=lambda x:x[1], reverse=True)[:100]
+    total = len(filtered_data)
+    top_aaSubs = {a:v/total for a,v in sorted(aaCounter.items(),key=lambda x:x[1], reverse=True)[:100]}
+    top_nucSubs = {a:v/total for a,v in sorted(nucCounter.items(),key=lambda x:x[1], reverse=True)[:100]}
+
     rate_data = {'clade':args.clade, 'nuc':regression_clean, 'aa':regression_clean_aa,
-                 'syn':regression_clean_syn, 'spike':regression_clean_spike, "top_subs": top_subs}
+                 'syn':regression_clean_syn, 'spike':regression_clean_spike,
+                 "top_aaSubs": top_aaSubs,  "top_nucSubs": top_nucSubs}
     if args.output_json:
         with open(args.output_json, 'w') as fh:
             json.dump(rate_data, fh)

### 2022-08-07 manuscript/variant_rates.tex
edits

@@
 \date{\today}
+
+\begin{abstract}
+The evolution of SARS-CoV-2 has had profound impact on the course of the pandemic and multiple variants of concern have led to renewed waves of infections.
+Here we analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes within and between variants.
+The rate of synonymous mutations is found to be around 6 changes per year.
+Rates within variants vary little and are compatible with a global rate.
+The non-synonymous changes within variants, however, was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
+Globally, the different variants of SARS-CoV-2 are compatible with an overall rate of about 25 amino acid changes per year, highlighting the qualitatively different adaptive evolutionary process gave rise to the different variants.
+We further show that many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
+\end{abstract}
+
 \maketitle
@@
 Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate robustly.
-Fig.~\ref{fig:rate_alpha} shows the increasing intra-clade divergence for clade 20I corresponding to VoC Alpha.
+Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
 Both synonymous and non-synonymous within-clade average divergence increases linearly over time allowing for a robust estimate of the rate.
@@
     \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.pdf}
-    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
-    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter.
+    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
+    \caption{{\bf Within-clade divergence increases linearly with time.} Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
+    The green line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter.
 Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
 Analogous figures for all clades considered are included in the appendix.
-    \label{fig:rate_alpha}}
+    Bottom: Panels D and E show the prevalence of specific genotypes (D) and specific mutations (E).
+In the case shown, the founder genotype initially dominates and no daughter genotype or mutation dominate.
+Panel F shows a Poisson model to the breakdown of the population into genotypes with different number of mutations over time.
+    \label{fig:within_clade}}
 \end{figure*}
 
-For almost all major Nextstrain clades, the average divergence increases linearly in time (similar patterns as Fig.~\ref{fig:rate_alpha}) allowing us to estimate clade specific evolutionary rates for amino acid and silent changes.
-These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
+A simple model for diversity in a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
+Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
+The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5days$ and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$.
+
+When considering a rapidly growing well sampled outbreak, typically a single founder genotype will give rise to a large number of daughter lineages that evolve independently.
+In this case, the diversification processes is robustly described by its mean
+Since the above branching process is linear, the mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation:
+\begin{equation}
+    \begin{split}
+        \frac{dn}{dt} &= \alpha n \\
+        \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
+        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
+        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
+        \cdots
+    \end{split}
+\end{equation}
+with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
+At each time the number of mutations in the population is expected to Poisson distributed with mean $ut$, while the overall number of cases is $e^{\alpha t}$ or more generally $e^{\int \alpha(t) dt}$ if growth rate or ascertainment varies over time.
+
+For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clade}F for clade 20C.
+Analogous plots for all other clades considered are included in the appendix.
+
+In some cases, subclades do become dominant in the early stochastic dynamic.
+This can result in increased within clade divergence, effectively generating an offset to diversity.
+Extrapolation of the linear fit to zero divergence is thus not necessarily a good estimator of the emergence time of a variant.
+For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder like variants are a minority even in early data.
+
+These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
+In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.
+In variant 21K (Omicron, BA.1), a sublineage with mutation $S:R346K$ might have enjoyed a transmission advantage.
+A possible consequence of this advantage is visible in Supp.~Fig.~\ref{fig:21K_divergence} where genotypes with 0, 1, or 2 mutations are decaying more rapidly than those with more mutations.
 
@@
 \end{figure*}
+
+Despite these caveats, for almost all major Nextstrain clades, the average divergence increases linearly in time (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C) allowing us to estimate clade specific evolutionary rates for amino acid and silent changes.
+These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
 
@@
 The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
-The latter is due to the rapid expansion and subsequent dominance of clade 20A and its descendants \citep{korber_tracking_2020}.
+The latter is due to the rapid expansion and subsequent dominance of clade 20A and its descendants which rapidly added four additional mutations \citep{korber_tracking_2020}.
 
@@
     \hline
-    19A &     19.76 &    12.79 &      6.97 &        0 &       0 &        0 \\
-    19B &     19.47 &    12.37 &      7.10 &        2 &       1 &        1 \\
-   19B+ &     18.55 &    11.90 &      6.65 &        2 &       1 &        1 \\
-  19B++ &     28.21 &    16.37 &     11.84 &        2 &       1 &        1 \\
-    20A &     15.66 &     8.40 &      7.25 &        4 &       2 &        2 \\
-    20B &     14.14 &     7.61 &      6.54 &        7 &       4 &        3 \\
-    20C &     16.30 &     9.52 &      6.78 &        6 &       4 &        2 \\
-   20A+ &     17.15 &     9.81 &      7.33 &        4 &       2 &        2 \\
-    20E &     13.65 &     5.74 &      7.91 &       11 &       4 &        7 \\
-    20H &     14.50 &     8.03 &      6.47 &       19 &      15 &        4 \\
-    20I &     12.18 &     5.85 &      6.33 &       28 &      18 &       10 \\
-    20J &     13.31 &     6.84 &      6.47 &       33 &      21 &       12 \\
-    21D &     11.36 &     6.29 &      5.07 &       25 &      12 &       13 \\
-    21G &     12.05 &     6.52 &      5.53 &       26 &      18 &        8 \\
-    21H &     18.17 &    10.00 &      8.17 &       29 &      20 &        9 \\
-    21I &     16.33 &     8.67 &      7.66 &       28 &      23 &        5 \\
-    21J &     12.25 &     6.32 &      5.93 &       32 &      27 &        5 \\
-    21K &     14.37 &     6.70 &      7.67 &       54 &      45 &        9 \\
-    21L &      9.02 &     3.80 &      5.22 &       66 &      49 &       17 \\
-    22A &      8.10 &     2.79 &      5.31 &       69 &      51 &       18 \\
-    22B &     12.07 &     4.97 &      7.11 &       64 &      49 &       15 \\
+    19B &     19.45 &    12.37 &      7.09 &        0 &       0 &        0 \\
+    19B+ &     18.51 &    11.87 &      6.64 &        0 &       0 &        0 \\
+    19B++ &     28.17 &    16.32 &     11.84 &        0 &       0 &        0 \\
+    19A &     19.75 &    12.79 &      6.96 &        2 &       1 &        1 \\
+    20A &     15.59 &     8.33 &      7.26 &        6 &       3 &        3 \\
+     20B &     14.17 &     7.64 &      6.53 &        9 &       5 &        4 \\
+     20C &     16.43 &     9.63 &      6.79 &        8 &       5 &        3 \\
+    20A+ &     17.16 &     9.76 &      7.40 &        6 &       3 &        3 \\
+     20E &     13.88 &     5.84 &      8.04 &       13 &       5 &        8 \\
+     20H &     15.46 &     9.50 &      5.96 &       21 &      16 &        5 \\
+     20I &     11.78 &     5.72 &      6.06 &       30 &      19 &       11 \\
+     20J &     12.91 &     6.29 &      6.62 &       35 &      22 &       13 \\
+     21D &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
+     21G &     12.03 &     6.51 &      5.51 &       28 &      19 &        9 \\
+     21H &     18.24 &    10.05 &      8.19 &       31 &      21 &       10 \\
+     21I &     16.26 &     8.63 &      7.63 &       30 &      24 &        6 \\
+     21J &     13.26 &     6.91 &      6.36 &       34 &      28 &        6 \\
+     21K &     14.04 &     6.55 &      7.49 &       56 &      46 &       10 \\
+     21L &      9.40 &     4.12 &      5.28 &       68 &      50 &       18 \\
+     22A &      8.07 &     2.84 &      5.23 &       71 &      52 &       19 \\
+     22B &     12.06 &     4.96 &      7.10 &       66 &      50 &       16 \\
          \hline
@@
 \caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
-    The last three columns give the distances of the clade founder sequence from the reference Wuhan/Hu-1.
+    The last three columns give the distances of the clade founder sequence from putative MRCA of SARS-CoV-2.
 \label{tab:rates}}
@@
 
-\subsection*{Diversification within clades}
-A simple model for a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
-Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
-The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5days$ and a rate of $\mu = 15/year$ evaluates to $u=0.2$.
-Since these branching process models are linear, them mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation.
-\begin{equation}
-    \begin{split}
-        \frac{dn}{dt} &= \alpha n \\
-        \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
-        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
-        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
-        \cdots
-    \end{split}
-\end{equation}
-with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
-
-For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clades}.
-For other clades, especially when the founder sequence is barely observed and dominant subclades developed, such fits are not robust.
-This is particularly true for the early clades 19A and 19B, as well as 21J (Delta), 21K (Omicron, BA.1), and 22B (Omicron, BA.5).
-
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
-    \caption{{\bf Diversification within clades.} \label{fig:within_clades}}
-\end{figure*}
-
-These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
-In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.
-
 \section*{Mutation tolerance}
@@
 
-When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape_supp}), the most constrained are ORF1b and M, while ORF3a, ORF6, ORF7a, ORF7b, and ORF8 show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
-N shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
+When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape_supp}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
+\emph{N} shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
 Rare variation at 3rd positions is very comparable between genes.
-Only E shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35.
+Only \emph{E} shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35.
 
@@
 This effect is particularly pronounced at the beginning of an outbreak following a host switch and has been attributed to methodological issues and incomplete purifying selection leading to inflated measures of diversity \citep{meyer_time_2015,ghafari_purifying_2022}.
-In addition to segregating deleterious mutations, early viral evolution after a host switch will also be affect by anomalously fast adaptive evolution as a dramatic change in environment, i.e.~a host switch, likely results in many mutations that increase fitness, possibly only in epistatic combinations with others.
+In addition to segregating deleterious mutations, early viral evolution after a host switch will also be affected by anomalously fast adaptive evolution as a dramatic change in environment, i.e.~a host switch, likely results in many mutations that increase fitness, though sometimes these better adapted states require multiple epistatically interacting mutations \citep{martin_emergence_2021}.
 Such transient increases in the rate of adaptation is common in experimental evolution \citep{elena_virus_2007}.
 
-The analysis presented here suggests that the early evolutionary rate of SARS-CoV-2 was inflated by adaptive evolution.
 Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of 6 month, purifying selection has comparable efficiency for all clades.
-Nevertheless, the non-synonymous evolutionary of clades circulating in late 2019 and early 2020 is about twice as high as that of clades in 2021 and 2020.
+Sequencing artefacts are expected to increase divergence at all time points or decrease with time as sequencing workflows improve.
+Nevertheless, the estimated non-synonymous evolutionary rate of clades circulating in late 2019 and early 2020 is about twice as high as that of clades in 2021 and 2022, while the synonymous rate does not change (see Fig.~\ref{fig:rate_summary} and Table \ref{tab:rates}).
+One possible explanation is that the early evolutionary rate of SARS-CoV-2 was inflated by adaptive evolution.
 
-Adaptive evolution complicates phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology.
-This assumption is true for neutral mutations that occur along every lineage with the same rate.
-Adaptive and deleterious mutations, however, affect the ability of the virus to spread.
+Most of this apparent acceleration of within-variant evolution ceased in mid 2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 per year instead of 12.
+By that time, the number of non-synonymous mutations was SARS-CoV-2 small (5 in the case of 20E) and given what is observed in subsequent variants, it is implausible that this small change would have exhausted the pool of beneficial mutations.
+However, the different replication and transmission kinetics of variants with S:D614G might mean that virus is operating closer to the maximal capacity at which cells can produce virions and the strength of selection is generally reduced.
+Difference in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments \citep{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.
 
+An evolutionary rate of 6 synonymous changes at around 9000 positions per year corresponds to a per-site mutation rate of $6.6\times 10^{-4}$ per site and year or $1.8\times 10^{-6}$/day, slightly higher than the estimated base-line mutation rate of $1.3\times 10^{-6}$/day \citep{amicone_mutation_2022}.
+The overall rate in variants starting with 20E is in the range of 9-16 changes per year, corresponding to a per site rate of $3 - 5\times 10^{-4}$/site and year, consistent with recent estimates by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
+
+When considering only the putative founder genotype and date of origin of each variant, all variants so far are compatible with a back-bone evolutionary rate of 32 changes per year, corresponding to an per site rate of around $10^{-3}$/site/year (see Fig.~\ref{fig:rate_summary}).
+This rate is a composite of episode of cryptic accelerated evolution (probably in chronically infected individuals), and regular transmission chains of acute infections.
+This is thus lower than the branch specific rates estimated by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
+However, all clades and variants are compatible with a single ``back-bone'' molecular clock without postulating significant differences in the emergence of different variants.
+This observation of course does not rule out that different clades and variants emerged in qualitatively different ways.
+
+By analysing the rate mutations that spread only on short times scales within fine-grained Pango lineages, we estimated the level of constraint on different parts of the SARS-CoV-genome.
+The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
+The major open reading frames ORF1ab, S, N, E, M show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
+This result is consistent with similar analysis in HIV that also show little fitness costs for most synonymous mutations, while half of non-synonymous mutations are so deleterious that they are not even observed on short times scales \citep{zanini_vivo_2017}.
+The remaining ORFs show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs_supp}).
+Only a few regions show a clear signal of conservation at 3rd positions, notably a central region of \emph{E} and the ribosome slippage region at the beginning of \emph{ORF1b}.
+
+The heterogeneity in evolutionary rates and the combination of adaptive evolution, approximately neutral mutations, and purifying selection complicate the interpretation of phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology and independent at differents.
+These assumptions are (approximately) true for neutral mutations that occur along every lineage with the same rate.
 Deleterious mutations are relatively straightforward to account for: Strongly deleterious mutations don't spread and are only observed on terminal branches, similar to sequencing errors.
@@
 A spectrum of mutations with different deleterious effects leads to time dependent effective evolutionary rates \citep{wertheim_purifying_2011}.
-The intra-clade rates in Fig.~\ref{fig:rate_progression} are all estimated over a similar time frame and the time dependent rate phenomenon should not apply.
 
@@
 Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
-Epistatic interactions can further steepen the dependence on population size and increase stochasticity.
 Methods to infer emergence dates and locations of the early variants might therefore be biased or over-confident.
@@
 \include{appendix}
+\onecolumngrid
 \include{appendix_rtt}
-\include{appendix_counts}
+% \include{appendix_counts}
 \end{document}

### 2022-08-07 scripts/combine_fits.py

@@ -32,8 +32,10 @@ if __name__=="__main__":
             dt = np.linspace(0, 0.75,2)
             t = dt + row[f'{mut_type}_origin']
             slope = row[f'{mut_type}_rate']
-            ax.plot(t, dt*slope + row[f'{mut_type}_div'], label=clade, c=f"C{ci%10}")
-            ax.scatter([[row[f'{mut_type}_origin']]], [[row[f'{mut_type}_div']]], c=f"C{ci%10}")
+            ls = '-' if ci<10 else '--'
+            m = 'o' if ci<10 else 's'
+            ax.plot(t, dt*slope + row[f'{mut_type}_div'], label=clade, c=f"C{ci%10}", ls=ls)
+            ax.scatter([[row[f'{mut_type}_origin']]], [[row[f'{mut_type}_div']]], c=f"C{ci%10}", marker=m)
             if row[f'{mut_type}_origin']>2019.7 and row[f'{mut_type}_origin']<2022.7:
                 inter_clade.append([row[f'{mut_type}_origin'], row[f'{mut_type}_div']])
             ci += 1
@@ -52,8 +54,13 @@ if __name__=="__main__":
     for i, (mut_type, rate) in enumerate(inter_clade_rates.items()):
         axs[-1,-1].plot([i-0.4, i+0.4], [rate, rate], lw=3, c='k', alpha=0.5)
         clade_rates = rates[f"{mut_type}_rate"]
-        axs[-1,-1].scatter(i - 0.35 + np.random.random(size=len(clade_rates))*0.7,
-                           clade_rates, c=[f"C{ci%10}" for ci in range(len(clade_rates))])
+        x_offset = i + np.linspace(-.35, 0.35,len(clade_rates))
+        axs[-1,-1].scatter(x_offset[:10],
+                           clade_rates[:10], c=[f"C{ci%10}" for ci in range(10)],
+                           marker='o')
+        axs[-1,-1].scatter(x_offset[10:],
+                           clade_rates[10:], c=[f"C{ci%10}" for ci in range(len(clade_rates) - 10)],
+                           marker='s')
     axs[-1,-1].set_ylabel("substitutions per year")
     axs[-1,-1].set_xticks([0,1,2], ['nuc', 'aa', 'syn'])
     axs[-1,-1].set_ylim(0)

### 2022-08-09 manuscript/variant_rates.tex
edits to ms

@@
 The non-synonymous changes within variants, however, was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
-Globally, the different variants of SARS-CoV-2 are compatible with an overall rate of about 25 amino acid changes per year, highlighting the qualitatively different adaptive evolutionary process gave rise to the different variants.
+Globally, the different variants of SARS-CoV-2 are compatible with an overall rate of about 25 amino acid changes per year, highlighting that the different adaptive evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains.
 We further show that many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
@@
 
+\begin{figure*}[tb]
+    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.pdf}
+    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
+    \caption{{\bf Within-clade divergence increases linearly with time.} Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
+    The green line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter.
+Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
+Analogous figures for all clades considered are included in the appendix.
+    Bottom: Panels D and E show the prevalence of specific genotypes (D) and specific mutations (E).
+In the case shown, the founder genotype initially dominates and no daughter genotype or mutation dominate.
+Panel F shows a Poisson model fit to the breakdown of the population into genotypes with different number of mutations over time.
+    \label{fig:within_clade}}
+\end{figure*}
+
 \section*{Results}
 
-Traditionally, evolutionary rates and divergence times are estimates using phylogenetic frameworks for heterochronously sampled sequences.
+Evolutionary rates and divergence times are typically estimated using phylogenetic frameworks for heterochronously sampled sequences.
 These methods, however, can not handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
@@
 Both synonymous and non-synonymous within-clade average divergence increases linearly over time allowing for a robust estimate of the rate.
-
-\begin{figure*}[tb]
-    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.pdf}
-    \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
-    \caption{{\bf Within-clade divergence increases linearly with time.} Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
-    The green line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter.
-Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
-Analogous figures for all clades considered are included in the appendix.
-    Bottom: Panels D and E show the prevalence of specific genotypes (D) and specific mutations (E).
-In the case shown, the founder genotype initially dominates and no daughter genotype or mutation dominate.
-Panel F shows a Poisson model to the breakdown of the population into genotypes with different number of mutations over time.
-    \label{fig:within_clade}}
-\end{figure*}
 
@@
 \begin{figure*}
-    \includegraphics[width=\textwidth]{figures/rate_summary.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/rate_summary.pdf}
     \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
 In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
-    The standard deviation expected based on Poisson statistics is indicated as shared area.
-    Panel D summarizes the individual rate estimate (dots) and compares them to the estimate inter-clade rates (gray lines).
+    The standard deviation expected based on Poisson statistics is indicated as shaded area.
+    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate inter-clade rates (gray lines).
     \label{fig:rate_summary} }
@@
 
-\section*{Mutation tolerance}
+\section*{Purifying selection and mutation tolerance}
 
 Above, we found that the rate of synonymous mutations is comparable within and between clades without any striking signal of temporal variation.
-This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements).
+This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements or splice sites).
 To assess the level of constraint on the SARS-CoV-2 genome, we made use of the ``rare mutations'' annotation by Nextclade.
 Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
-For each pango lineage (as determined by Nextclade), we count how often these ``rare mutations'' (including reversions to reference) are observed, normalize this against the relative rate of mutation away from the ancestral nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
+For each Pango lineage (as determined by Nextclade) \citep{rambaut_dynamic_2020,aksamentov_nextclade_2021}, we count how often these ``rare mutations'' (including reversions to reference) are observed, normalize this against the relative rate of mutation away from the ancestral nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
 
@@
 
-When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape_supp}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
+When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
 \emph{N} shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
@@
 
+ \begin{figure*}[tb]
+    \includegraphics[width=\textwidth]{figures/fitness_landscape.pdf}
+    \caption{{\bf Landscape of selective constraint along the SARS-CoV-2 genome.}
+    Solid lines show sliding window smoothing of estimated mutational tolerance at 1st, 2nd, and 3rd positions for a window size of 20 and 7 (faint lines) sites.
+    The markers below the lines are a qualitative indication of variable positions across SARS-CoV-2 diversity.
+    \label{fig:fitness_landscape}}
+ \end{figure*}
+
 \section*{Discussion}
@@
 In addition to segregating deleterious mutations, early viral evolution after a host switch will also be affected by anomalously fast adaptive evolution as a dramatic change in environment, i.e.~a host switch, likely results in many mutations that increase fitness, though sometimes these better adapted states require multiple epistatically interacting mutations \citep{martin_emergence_2021}.
-Such transient increases in the rate of adaptation is common in experimental evolution \citep{elena_virus_2007}.
+Such transient increases in the rate of adaptation are common in experimental evolution \citep{elena_virus_2007}.
 
@@
 However, the different replication and transmission kinetics of variants with S:D614G might mean that virus is operating closer to the maximal capacity at which cells can produce virions and the strength of selection is generally reduced.
-Difference in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments \citep{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.
+Such ``diminishing return'' epistasis as been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
+Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments \citep{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.
 
@@
 
-By analysing the rate mutations that spread only on short times scales within fine-grained Pango lineages, we estimated the level of constraint on different parts of the SARS-CoV-genome.
+By analyzing the rate mutations that spread only on short times scales within fine-grained Pango lineages, we estimated the level of constraint on different parts of the SARS-CoV-genome.
 The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
@@
 This result is consistent with similar analysis in HIV that also show little fitness costs for most synonymous mutations, while half of non-synonymous mutations are so deleterious that they are not even observed on short times scales \citep{zanini_vivo_2017}.
-The remaining ORFs show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs_supp}).
+The remaining ORFs show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
 Only a few regions show a clear signal of conservation at 3rd positions, notably a central region of \emph{E} and the ribosome slippage region at the beginning of \emph{ORF1b}.
 
-The heterogeneity in evolutionary rates and the combination of adaptive evolution, approximately neutral mutations, and purifying selection complicate the interpretation of phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology and independent at differents.
+The heterogeneity in evolutionary rates and the combination of adaptive evolution, approximately neutral mutations, and purifying selection complicate the interpretation of phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology and that different sites evolve independently.
 These assumptions are (approximately) true for neutral mutations that occur along every lineage with the same rate.
@@
 
-Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
-Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
-Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
-Methods to infer emergence dates and locations of the early variants might therefore be biased or over-confident.
+Adaptive evolution, however, is much harder to account for properly.
+Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic -- it is not the typical events, but the rare and extreme events that determine the course of adaptive evolution.
+Unlike neutral evolution, the rate of adaptive evolution depends on the population size, stochastic nature of the transmission process, the environment, and previous adaptation.
 
 \bibliography{bib}
-\appendix
+%\appendix
 \include{appendix}

### 2022-08-09 scripts/combine_fits.py
change mutation normalization

@@ -15,7 +15,7 @@ if __name__=="__main__":
         description="remove time info",
         formatter_class=argparse.ArgumentDefaultsHelpFormatter
     )
-
+    fs=14
     parser.add_argument('--rate-table', type=str, required=True, help="input data")
     parser.add_argument('--output-plot', type=str, help="plot file")
     parser.add_argument('--output-plot-rates', type=str, help="plot file")
@@ -24,8 +24,10 @@ if __name__=="__main__":
     rates = pd.read_csv(args.rate_table, sep='\t', index_col='clade')
     inter_clade_rates = {}
     fig, axs = plt.subplots(2,2, figsize=(12,12))
-    for ax, mut_type, ax_label in zip(axs.flatten(),['nuc', 'aa', 'syn'],['total divergence', 'aa divergence', 'syn divergence']):
-        ax.set_ylabel(ax_label)
+    xticks = [2020,2020.5,2021,2021.5,2022,2022.5]
+    for ax, mut_type, ax_label in zip(axs.flatten(),['nuc', 'aa', 'syn'],
+                        ['total divergence', 'amino acid divergence', 'synonymous divergence']):
+        ax.set_ylabel(ax_label, fontsize=fs)
         inter_clade = []
         ci=0
         for clade, row in rates.iterrows():
@@ -51,6 +53,7 @@ if __name__=="__main__":
         ax.set_xlim(x.min(), x.max())
         ax.set_ylim(0)
         if mut_type == 'aa': ax.legend(ncol=2)
+        ax.set_xticks(xticks, [str(x) for x in xticks], fontsize=fs)
     for i, (mut_type, rate) in enumerate(inter_clade_rates.items()):
         axs[-1,-1].plot([i-0.4, i+0.4], [rate, rate], lw=3, c='k', alpha=0.5)
         clade_rates = rates[f"{mut_type}_rate"]
@@ -61,8 +64,9 @@ if __name__=="__main__":
         axs[-1,-1].scatter(x_offset[10:],
                            clade_rates[10:], c=[f"C{ci%10}" for ci in range(len(clade_rates) - 10)],
                            marker='s')
-    axs[-1,-1].set_ylabel("substitutions per year")
-    axs[-1,-1].set_xticks([0,1,2], ['nuc', 'aa', 'syn'])
+    axs[-1,-1].set_ylabel("substitutions per year", fontsize=fs)
[…]

### 2022-08-09 scripts/count_mutations.py

@@ -1,4 +1,5 @@
 import argparse,json
+from unittest.mock import NonCallableMagicMock
 import pandas as pd
 import numpy as np
 import matplotlib as mpl
@@ -131,30 +132,24 @@ if __name__=="__main__":
                 elif 'X' not in [aa,alt_aa]:
                     nonsyn[gene][codon_number][a].append(c)
 
-    # determine the total number of events and counts at each position
-    def total_len_filtered_lists(l, cutoff):
-        return sum([len([y for y in x if y>=cutoff]) for x in l])
-
     sorted_transitions = []
     for a in 'ACGT':
         for d in 'ACGT':
             if a!=d:
                 sorted_transitions.append((a,d))
 
+    # determine the total number of events and counts at each position
+    def total_len_filtered_lists(l, cutoff, rates=None):
+        if rates is None:
+            rates = {}
+        return sum([len([y for y in x if y>=cutoff])/rates.get(a,1.0) for (a,d),x in l.items()])
+
     all_events = []
     for pos in range(len(ref)):
         tmp = []
         for a, d in sorted_transitions:
-            tmp.append(total_len_filtered_lists([position_counter[pos][(a,d)]], args.rare_cutoff))
+            tmp.append(total_len_filtered_lists({(a,d): position_counter[pos][(a,d)]}, args.rare_cutoff))
         all_events.append(tmp)
-
-    total_events = np.array([total_len_filtered_lists(position_counter[pos].values(), args.rare_cutoff)
-                            if pos in position_counter else 0
-                            for pos in range(len(ref))])
-
-    total_counts = np.array([sum([sum(x) for x in position_counter[pos].values()])
-                             if pos in position_counter else 0
-                             for pos in range(len(ref))])
 
     # sort all mutations by type of mutation
     events_by_transition = defaultdict(int)
@@ -170,9 +165,23 @@ if __name__=="__main__":
             transition_rate[(a,d)] = events_by_transition[(a,d)]/base_content[a]/total_rate*len(ref)
             away_rate[a] += transition_rate[(a,d)]
 
+
+    total_events = np.array([total_len_filtered_lists(position_counter[pos], args.rare_cutoff)
+                            if pos in position_counter else 0
+                            for pos in range(len(ref))])
+
[…]

### 2022-08-13 manuscript/variant_rates.tex
edit pass

@@
 \begin{document}
-\title{Evolutionary rates of SARS-CoV-2}
+\title{Contributions of adaptive evolution and purifying selection to SARS-CoV-2 diversification}
 \author{Richard A.~Neher}
+\address{Biozentrum, University of Basel, Basel, Switzerland \looseness=-5}
+\address{Swiss Institute of Bioinformatics, Switzerland \looseness=-5}
+
 \date{\today}
@@
 \begin{abstract}
-The evolution of SARS-CoV-2 has had profound impact on the course of the pandemic and multiple variants of concern have led to renewed waves of infections.
-Here we analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes within and between variants.
+Continued evolution and adaptation of SARS-CoV-2 has lead to more transmissible and immune-evasive variants with profound impact on the course of the pandemic.
+Here we analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes.
+We estimate separate rates for evolution within clades, that is well defined mono-phyletic groups with gradual evolution, and for the pandemic overall.
 The rate of synonymous mutations is found to be around 6 changes per year.
-Rates within variants vary little and are compatible with a global rate.
-The non-synonymous changes within variants, however, was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
-Globally, the different variants of SARS-CoV-2 are compatible with an overall rate of about 25 amino acid changes per year, highlighting that the different adaptive evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains.
-We further show that many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
+Rates within variants vary little and are compatible with a overall rate.
+In contrast, the rate at which variants accumulate amino acid changes was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
+The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 25 amino acid changes per year.
+This 2-fold higher rate indicates that adaptive evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains.
+We further quantify the spectrum of mutations and purifying selection in different SARS-CoV proteins.
+Many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
+About half of mutations in other proteins are strongly deleterious and rarely observed, even at low frequency.
 \end{abstract}
@@
 
-Since its emergence in late 2019, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants without observed intermediates \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
+Since its emergence in late 2019 \citep{zhu_novel_2020}, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants without observed intermediates \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
 Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
-Such variants of Concern or Interest were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
+Specific Variants of Concern (VoC) or Interest (VoI) were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
 The branches leading to these variants are characterized by many amino acid changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.
 
-This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells in common among many RNA viruses and for example well studied in influenza virus A evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
-Adaptive evolution in influenza tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
+This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells is common among many RNA viruses and for example well studied in influenza virus A evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
+But adaptive evolution of influenza viruses tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
 The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are frequently observed in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
@@
 \begin{figure*}[tb]
-    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.pdf}
+    \includegraphics[width=\textwidth]{figures/rtt/20C_rtt.pdf}
     \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
@@
 These methods, however, can not handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
-Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any problematic sequence can fatally distort the analysis, in particular if placed early in tree.
+Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated or recombinant sequence can fatally distort the analysis, in particular if placed early in tree.
 
-To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns.
+To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
 We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018} and analyze each clade separately.
-Sequences belonging to recombinant lineages are excluded.
+Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
 These clades represent well defined groups of sequences with little non-vertical evolution within them.
@@
 The exact functional form of this weighing does not have a big influence on the results, but a sublinear weighing helps to counter the large variation in sequencing effort across countries and the natural imbalance due to the fact that few sequences are available early.
+Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
+Both synonymous and non-synonymous within-clade average divergence increases linearly over time allowing for a robust estimate of the rate.
 
@@
 Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate robustly.
-Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
-Both synonymous and non-synonymous within-clade average divergence increases linearly over time allowing for a robust estimate of the rate.
 
@@
 For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clade}F for clade 20C.
+In this case, the founder genotypes initially dominates, but is gradually replaced -- first single mutant genotypes, then double mutants, and so forth (comp.~Fig.~\ref{fig:within_clade}).
 Analogous plots for all other clades considered are included in the appendix.
 
-In some cases, subclades do become dominant in the early stochastic dynamic.
+In some cases, notable subclades do become dominant in the early stochastic dynamic.
 This can result in increased within clade divergence, effectively generating an offset to diversity.
@@
 In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.
-In variant 21K (Omicron, BA.1), a sublineage with mutation $S:R346K$ might have enjoyed a transmission advantage.
+In variant 21K (Omicron, BA.1), a sublineage with mutation \texttt{S:R346K} might have enjoyed a transmission advantage.
 A possible consequence of this advantage is visible in Supp.~Fig.~\ref{fig:21K_divergence} where genotypes with 0, 1, or 2 mutations are decaying more rapidly than those with more mutations.
+
+Founder genotypes of Omicron clades 21L (BA.2), 22A (BA.4) and 22B (BA.5) are barely sampled.
+These clades showed considerable diversity, including reversions to the reference, soon after their initial discovery \citep{tegally_emergence_2022}.
+This diversity and the complex, possibly recombinant, origin of the Omicron clades makes estimating their rate of evolution challenging.
+The lower estimates for these latest clades should therefore be interpreted with care.
 
@@
     \hline
-    19B &     19.45 &    12.37 &      7.09 &        0 &       0 &        0 \\
-    19B+ &     18.51 &    11.87 &      6.64 &        0 &       0 &        0 \\
-    19B++ &     28.17 &    16.32 &     11.84 &        0 &       0 &        0 \\
-    19A &     19.75 &    12.79 &      6.96 &        2 &       1 &        1 \\
-    20A &     15.59 &     8.33 &      7.26 &        6 &       3 &        3 \\
-     20B &     14.17 &     7.64 &      6.53 &        9 &       5 &        4 \\
-     20C &     16.43 &     9.63 &      6.79 &        8 &       5 &        3 \\
-    20A+ &     17.16 &     9.76 &      7.40 &        6 &       3 &        3 \\
-     20E &     13.88 &     5.84 &      8.04 &       13 &       5 &        8 \\
-     20H &     15.46 &     9.50 &      5.96 &       21 &      16 &        5 \\
-     20I &     11.78 &     5.72 &      6.06 &       30 &      19 &       11 \\
-     20J &     12.91 &     6.29 &      6.62 &       35 &      22 &       13 \\
+    19B &     19.46 &    12.37 &      7.09 &        0 &       0 &        0 \\
+   19B+ &     18.56 &    11.92 &      6.64 &        0 &       0 &        0 \\
+  19B++ &     27.91 &    16.41 &     11.50 &        0 &       0 &        0 \\
+  19A &     19.73 &    12.79 &      6.95 &        2 &       1 &        1 \\
+  20A &     15.08 &     7.94 &      7.15 &        6 &       3 &        3 \\
+    20B &     14.29 &     7.69 &      6.59 &        9 &       5 &        4 \\
+    20C &     16.73 &     9.92 &      6.81 &        8 &       5 &        3 \\
+   20A+ &     16.30 &     9.40 &      6.90 &        6 &       3 &        3 \\
+    20E &     14.04 &     5.65 &      8.39 &       13 &       5 &        8 \\
+    20H &     15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
+    20I &     11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
+    20J &     12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
     21D &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
-     21G &     12.03 &     6.51 &      5.51 &       28 &      19 &        9 \\
-     21H &     18.24 &    10.05 &      8.19 &       31 &      21 &       10 \\
-     21I &     16.26 &     8.63 &      7.63 &       30 &      24 &        6 \\
-     21J &     13.26 &     6.91 &      6.36 &       34 &      28 &        6 \\
-     21K &     14.04 &     6.55 &      7.49 &       56 &      46 &       10 \\
-     21L &      9.40 &     4.12 &      5.28 &       68 &      50 &       18 \\
-     22A &      8.07 &     2.84 &      5.23 &       71 &      52 &       19 \\
-     22B &     12.06 &     4.96 &      7.10 &       66 &      50 &       16 \\
+    21G &     12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
+    21H &     18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
+    21I &     16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
+    21J &     14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
+    21K &     13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
+    21L &      9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
+    22A &      7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
+    22B &     11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
            \hline
@@
 
-Above, we found that the rate of synonymous mutations is comparable within and between clades without any striking signal of temporal variation.
+Above, we found that the rate of synonymous mutations is comparable within and between clades without a strong signal of temporal variation.
 This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements or splice sites).
-To assess the level of constraint on the SARS-CoV-2 genome, we made use of the ``rare mutations'' annotation by Nextclade.
+To assess the level of constraint on the SARS-CoV-2 genome more generally, we made use of the ``rare mutations'' annotation by Nextclade.
 Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
@@
 Simply splitting mutations into those that occur in 1st, 2nd, and 3rd positions of codons already reveals strong signatures of purifying selection, see Fig.~\ref{fig:fitness_costs}.
-Between 15-20\% of positions at first and second codons show almost no variation, while half of these sites are less variable then the most constrained 10\% of third positions.
+Between 15-20\% of first and second positions in codons show almost no variation, while half of these sites are less variable then the most constrained 10\% of third positions.
 The median of variation at 3rd position is more than double that at other 1st and 2nd positions.
@@
 The inferred evolutionary rate of RNA viruses often decreases with the time scale across which is it estimated \citep{wertheim_purifying_2011,ghafari_prisoner_2021}.
-This effect is particularly pronounced at the beginning of an outbreak following a host switch and has been attributed to methodological issues and incomplete purifying selection leading to inflated measures of diversity \citep{meyer_time_2015,ghafari_purifying_2022}.
-In addition to segregating deleterious mutations, early viral evolution after a host switch will also be affected by anomalously fast adaptive evolution as a dramatic change in environment, i.e.~a host switch, likely results in many mutations that increase fitness, though sometimes these better adapted states require multiple epistatically interacting mutations \citep{martin_emergence_2021}.
+This effect can be particularly pronounced at the beginning of an outbreak following a host switch and has been attributed to incomplete purifying selection or methodological issues leading to inflated measures of diversity \citep{meyer_time_2015,ghafari_purifying_2022}.
+In addition to segregating deleterious mutations, early viral evolution after a host switch can also be driven by anomalously fast adaptation.
+A dramatic change in environment, e.g.~a host switch, likely results in many opportunities for mutations that increase fitness.
 Such transient increases in the rate of adaptation are common in experimental evolution \citep{elena_virus_2007}.
 
-Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of 6 month, purifying selection has comparable efficiency for all clades.
-Sequencing artefacts are expected to increase divergence at all time points or decrease with time as sequencing workflows improve.
+Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of 6 months, purifying selection has comparable efficiency and should affect our estimates equally for all clades.
+Sequencing artifacts are expected to increase divergence at all time points or decrease with time as sequencing workflows improve.
 Nevertheless, the estimated non-synonymous evolutionary rate of clades circulating in late 2019 and early 2020 is about twice as high as that of clades in 2021 and 2022, while the synonymous rate does not change (see Fig.~\ref{fig:rate_summary} and Table \ref{tab:rates}).
@@
 Most of this apparent acceleration of within-variant evolution ceased in mid 2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 per year instead of 12.
-By that time, the number of non-synonymous mutations was SARS-CoV-2 small (5 in the case of 20E) and given what is observed in subsequent variants, it is implausible that this small change would have exhausted the pool of beneficial mutations.
-However, the different replication and transmission kinetics of variants with S:D614G might mean that virus is operating closer to the maximal capacity at which cells can produce virions and the strength of selection is generally reduced.
+By that time, the number of non-synonymous differences relative to the reference was small (5 in the case of 20E) and it is implausible that this small change would have exhausted the pool of beneficial mutations.
+So why would the rate of adaptation evolution slow down?
+Viruses with the S:D614G showed a faster replication kinetics compared to earlier variants \citep{korber_tracking_2020}.
+The virus might therefore be operating closer to the maximal capacity at which cells can produce virions, reducing the scope for further optimization.
 Such ``diminishing return'' epistasis as been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
+Such global epistasis can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021}.
 Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments \citep{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.
@@
 However, all clades and variants are compatible with a single ``back-bone'' molecular clock without postulating significant differences in the emergence of different variants.
-This observation of course does not rule out that different clades and variants emerged in qualitatively different ways.
+This observation of course does not rule out that particular clades and variants emerged in qualitatively different ways, but exponential amplification of lineages with adaptive non-synonymous mutations results in elevated rates of evolution.
 
+In addition to neutral and potentially adaptive mutations, we also quantified purifying selection on the SARS-CoV-2 genome.
 By analyzing the rate mutations that spread only on short times scales within fine-grained Pango lineages, we estimated the level of constraint on different parts of the SARS-CoV-genome.
 The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
-The major open reading frames ORF1ab, S, N, E, M show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
-This result is consistent with similar analysis in HIV that also show little fitness costs for most synonymous mutations, while half of non-synonymous mutations are so deleterious that they are not even observed on short times scales \citep{zanini_vivo_2017}.
-The remaining ORFs show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
+The major open reading frames \emph{ORF1ab, S, N, E, M} show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
+This result is consistent with similar analysis in HIV-1 that also show small fitness costs for most synonymous mutations, while half of non-synonymous mutations are so deleterious that they are not even observed on short times scales \citep{zanini_vivo_2017}.
+The remaining SARS-CoV-2 ORFs show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
+This lack of strong purifying selection in these ORFs does not necessarily imply that they don't matter or are not a possible locus of adaptation -- but the exact amino acid sequence does not seem essential for their function.
 Only a few regions show a clear signal of conservation at 3rd positions, notably a central region of \emph{E} and the ribosome slippage region at the beginning of \emph{ORF1b}.
 
-The heterogeneity in evolutionary rates and the combination of adaptive evolution, approximately neutral mutations, and purifying selection complicate the interpretation of phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology and that different sites evolve independently.
+The heterogeneity in evolutionary rates and the combination of adaptive evolution, approximately neutral mutations, and purifying selection complicate the interpretation of phylodynamic analysis.
+Phylogenetics typically assumes that the mutation process is independent of the spread and epidemiology and that different sites evolve independently.
 These assumptions are (approximately) true for neutral mutations that occur along every lineage with the same rate.
-Deleterious mutations are relatively straightforward to account for: Strongly deleterious mutations don't spread and are only observed on terminal branches, similar to sequencing errors.
-Weakly deleterious mutations can spread, but lineages that carry them tend to be short-lived.
-A spectrum of mutations with different deleterious effects leads to time dependent effective evolutionary rates \citep{wertheim_purifying_2011}.
+Strongly deleterious mutations don't spread and are only observed on terminal branches, similar to sequencing errors.
+Weakly deleterious mutations can spread, but lineages that carry them tend to die out.
+Overall, the deleterious mutations result in increased diversity on short time scales compared to longer ones, which can lead to time-dependent effective evolutionary rates \citep{wertheim_purifying_2011}.
 
-Adaptive evolution, however, is much harder to account for properly.
+The effects of adaptive evolution are harder to control and account for.
 Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic -- it is not the typical events, but the rare and extreme events that determine the course of adaptive evolution.
 Unlike neutral evolution, the rate of adaptive evolution depends on the population size, stochastic nature of the transmission process, the environment, and previous adaptation.
+These factors make extrapolation -- whether it is reconstruction of past events or scenarios of future evolution -- uncertain.
+Standard phylogenetic will underestimate that uncertainty.
+
+Our analysis here has shown that the majority of non-synonymous changes by which SARS-CoV-2 variants differ from each other are likely of adaptive origin.
+If the excess of non-synonymous changes on the pandemic scale compared to within clade evolution is due to adaptive evolution, the suggests an adaptive rate of about 20 changes per year (plus insertion/deletions).
+This process is likely to continue even in the more and more diverse immunological landscape of the human population in which the virus adapts.
+But it is unclear whether SARS-CoV-2 continues to evolve in a saltatory fashion with repeated emergence of highly mutated variants, or whether we will see a transition to a more gradual adaptive process.
+We currently observe multiple lineage with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), suggesting that shift to a more influenza-like evolution.
+
+\section*{Materials and Methods}
+
+\subsubsection*{Data}
+This analysis is based on SARS-CoV-2 sequencing data shared by researchers around the world via GISAID \citep{shu_gisaid_2017} (EPISET\_ID \href{10.55876/gis8.220812zf}{EPI_SET_220812zf}).
+The bulk of the analysis can be reproduced using SARS-CoV-2 genomes shared via INSDC.
+These data are available from Nextstrain at \url{http://data.nextstrain.org/files/ncov/open/metadata.tsv.gz} and \url{http://data.nextstrain.org/files/ncov/open/nextclade.tsv.gz}.
+However, for some clades and variants, coverage in INSDC is low.
+
+\subsubsection*{Analysis}
+The code used to run this analysis is publicly available at \url{https://github.com/neherlab/SC2_variant_rates}.
+It is organized as a snakemake workflow \citep{koster_snakemakescalable_2012}.
+We used Nextclade data \texttt{2022-07-26T12:00:00Z} with reference sequence MN908947 (Wuhan-Hu-1).
+Tabular files with estimates of evolutionary rates, the mutation distribution, and fitness costs are included in the git repository.
+
+\section*{Acknowledgements}
+We gratefully acknowledge the work by 1000s of scientists and healthcare professionals that collect the samples, generate, and share the sequence data on which this analysis is built.
+We are also grateful for stimulating discussions with Cornelius Roemer and Emma Hodcroft on many aspects of SARS-CoV-2 evolution.
 
 \bibliography{bib}
-%\appendix
-\include{appendix}
+
+\clearpage
+\appendix
+\input{appendix}
+\clearpage
 \onecolumngrid
-\include{appendix_rtt}
+\input{appendix_rtt}
 % \include{appendix_counts}

### 2022-08-13 scripts/combine_fits.py

@@ -52,9 +52,12 @@ if __name__=="__main__":
         ax.fill_between(x, y+std_dev, np.maximum(0, y-std_dev), fc='k', lw=3, alpha=0.1, ec=None)
         ax.set_xlim(x.min(), x.max())
         ax.set_ylim(0)
+        ax.text( 0.6, 0.1,f"rate: {reg.slope:1.1f}/year", fontsize=fs, transform=ax.transAxes)
         if mut_type == 'aa': ax.legend(ncol=2)
         ax.set_xticks(xticks, [str(x) for x in xticks], fontsize=fs)
+
     for i, (mut_type, rate) in enumerate(inter_clade_rates.items()):
+        axs[-1,-1].fill_between([i-0.5, i+0.5], [35,35], facecolor='k', alpha=0.075*(2+i%2))
         axs[-1,-1].plot([i-0.4, i+0.4], [rate, rate], lw=3, c='k', alpha=0.5)
         clade_rates = rates[f"{mut_type}_rate"]
         x_offset = i + np.linspace(-.35, 0.35,len(clade_rates))
@@ -80,6 +83,7 @@ if __name__=="__main__":
     plt.plot(rates["nuc_origin"], np.ones_like(rates['nuc_origin'])*inter_clade_rates["aa"], label='inter-clade amino acid rate', c=f"C{0}", lw=3)
     plt.plot(rates["nuc_origin"], np.ones_like(rates['nuc_origin'])*inter_clade_rates["syn"], label='inter-clade synonymous rate', c=f"C{1}", lw=3)
     plt.ylabel('rate estimate [1/y]')
+    plt.ylim(0,25)
     plt.legend()
 
     if args.output_plot_rates:

### 2022-08-13 scripts/plot_fitness.py

@@ -39,7 +39,7 @@ if __name__=="__main__":
 
     ind = fitness.pos_in_codon.isna()
     plt.plot(sorted(fitness.tolerance[ind]), np.linspace(0,1,ind.sum()),
-                    label = f"non-codon")
+                    label = f"non-coding")
 
     ind = fitness.pos_in_codon==3
     syn_cutoff = scoreatpercentile(fitness.tolerance[ind],10)
@@ -51,7 +51,7 @@ if __name__=="__main__":
     print("non coding", np.mean(fitness.tolerance[ind]<syn_cutoff))
 
     plt.xscale('log')
-    plt.ylabel("fraction less")
+    plt.ylabel("fraction below")
     plt.xlabel("scaled number of lineages with mutations")
     plt.legend()
     plt.savefig(args.output_fitness)

### 2022-08-15 scripts/root_to_tip.py
add orf1ab to rtt

@@ -47,6 +47,7 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
     # define "with-in clade substitutions"
     d["intra_aaSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y] if x else [])
     d["intra_SpikeSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y and y[0]=='S'] if x else [])
+    d["intra_ORF1Substitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y and y[:4]=='ORF1'] if x else [])
 
     if swap_root:
         muts = [("C8782T","T8782C"), ("T28144C","C28144T")]
@@ -64,6 +65,7 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
     d["divergence"] =   d.intra_substitutions.apply(lambda x:   len(x))
     d["aaDivergence"] = d.intra_aaSubstitutions.apply(lambda x: len(x))
     d["spikeDivergence"] = d.intra_SpikeSubstitutions.apply(lambda x: len(x))
+    d["orf1Divergence"] = d.intra_ORF1Substitutions.apply(lambda x: len(x))
     d["synDivergence"] = d["divergence"] - d["aaDivergence"]
 
 
@@ -183,6 +185,7 @@ if __name__=="__main__":
     regression_clean_aa = regression_by_week(filtered_data.loc[ind], "aaDivergence")
     regression_clean_syn = regression_by_week(filtered_data.loc[ind], "synDivergence")
     regression_clean_spike = regression_by_week(filtered_data.loc[ind], "spikeDivergence")
+    regression_clean_ORF1 = regression_by_week(filtered_data.loc[ind], "orf1Divergence")
 
     fig, axs = plt.subplots(1,3, figsize=(18,6), sharex=True, sharey=True)
     ymax = 20
@@ -231,6 +234,7 @@ if __name__=="__main__":
 
     rate_data = {'clade':args.clade, 'nuc':regression_clean, 'aa':regression_clean_aa,
                  'syn':regression_clean_syn, 'spike':regression_clean_spike,
+                 'orf1':regression_clean_ORF1,
[…]

### 2022-08-15 manuscript/variant_rates.tex
add orf1ab to rtt
edits

@@
 \begin{document}
-\title{Contributions of adaptive evolution and purifying selection to SARS-CoV-2 diversification}
+\title{Contributions of adaptation and purifying selection to SARS-CoV-2 evolution}
 \author{Richard A.~Neher}
@@
 
-Since its emergence in late 2019 \citep{zhu_novel_2020}, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants without observed intermediates \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
+Since its emergence in late 2019 \citep{zhu_novel_2020}, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
 Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
@@
 This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells is common among many RNA viruses and for example well studied in influenza virus A evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
-But adaptive evolution of influenza viruses tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
-The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are frequently observed in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
+But adaptive evolution of influenza viruses tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with 10s of novel mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
+The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are common in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
 Onward transmission from such chronic infections has also been documented \citep{gonzalez-reiche_intrahost_2022}.
@@
 
-The saltatory nature of SARS-CoV-2 evolution implies that evolution within individual variants, clades, or lineages might have different characteristics from its overall pattern, which is dominated by atypical bursts of evolution leading to new variants.
+The saltatory nature of SARS-CoV-2 evolution implies that evolution within individual variants, clades, or lineages might have different characteristics from the global pattern, which is dominated by atypical bursts of evolution leading to new variants.
 \citet{tay_emergence_2022} and \citet{hill_origins_2022} show that the rate of evolution along these inter-clade branches is up to four-fold higher than the background rate.
-Here, we build on these results and investigate the patterns of SARS-CoV-2 diversifications within variants and compare these to the global dynamics of evolution and adaptation.
+Here, we build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
 This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
-The rate of synonymous evolution does not differ between and within variants, while the non-synonymous rate within variants is far below the between variant rate.
-Early variants display more rapid non-synonymous evolution that later variants.
+This different in evolutionary rate is only seen for non-synonymous changes -- the rate of synonymous evolution within variants is compatible with that seen between variants.
+Furthermore, early variants display more rapid non-synonymous evolution than later variants suggesting more ubiquitous adaptive evolution early on.
 We further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.
@@
 
-Evolutionary rates and divergence times are typically estimated using phylogenetic frameworks for heterochronously sampled sequences.
+Evolutionary rates and divergence times are typically estimated using phylogenetic \citep{drummond_relaxed_2006}.
 These methods, however, can not handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
-Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated or recombinant sequence can fatally distort the analysis, in particular if placed early in tree.
+Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated, recombinant, contaminated, or otherwise chimeric sequence can fatally distort the analysis, in particular if placed early in tree.
 
 To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
-We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018} and analyze each clade separately.
+We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots).
 Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
-These clades represent well defined groups of sequences with little non-vertical evolution within them.
+These clades represent well defined groups of sequences with little non-vertical evolution within them and we analyze data from each clade separately.
 Furthermore, we define a ``founder'' genotype for each clade and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
@@
 Within each clade, the number of mutations is expected to increase linearly in time and the variation around this mean would, in an ideal case, obey Poisson statistics.
-For the majority of sequences, this is approximately true, but some problematic sequences are many more mutations than expected.
-To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected variance by 3.
+For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
+To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3.
 
 After removing these outliers, we now bin the data by calendar week and determine the mean and standard deviation in each bin.
-Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each bin is weighted with the fourth root of the number of sequences in the bins.
-The exact functional form of this weighing does not have a big influence on the results, but a sublinear weighing helps to counter the large variation in sequencing effort across countries and the natural imbalance due to the fact that few sequences are available early.
+Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each time bin is weighted with the fourth root of the number of sequences in the bins.
+The exact functional form of this weighing does not have a big influence on the results, but a sublinear weighing helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early.
 Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
-Both synonymous and non-synonymous within-clade average divergence increases linearly over time allowing for a robust estimate of the rate.
+Both synonymous and non-synonymous within-clade average divergence increase linearly over time allowing for a robust estimate of the rate.
 
@@
 However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from one or several basal polytomies.
-Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate robustly.
+Along each of these lineages, mutation accumulation is independent.
+While not every sequence is an independent sample, the effective number of independent sampled is large and the steadily increasing average divergence allows to estimate the rate robustly.
 
-A simple model for diversity in a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
+A simple model for diversity within a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
 Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
@@
 When considering a rapidly growing well sampled outbreak, typically a single founder genotype will give rise to a large number of daughter lineages that evolve independently.
-In this case, the diversification processes is robustly described by its mean
+In this case, the diversification processes is robustly described by its mean.
 Since the above branching process is linear, the mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation:
@@
 For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clade}F for clade 20C.
-In this case, the founder genotypes initially dominates, but is gradually replaced -- first single mutant genotypes, then double mutants, and so forth (comp.~Fig.~\ref{fig:within_clade}).
-Analogous plots for all other clades considered are included in the appendix.
+In this case, the founder genotype initially dominates, but is gradually replaced -- first by single mutant genotypes, then double mutants, and so forth (comp.~Fig.~\ref{fig:within_clade}).
+Analogous graphs for all other clades considered are included in the appendix.
 
@@
 
-Founder genotypes of Omicron clades 21L (BA.2), 22A (BA.4) and 22B (BA.5) are barely sampled.
-These clades showed considerable diversity, including reversions to the reference, soon after their initial discovery \citep{tegally_emergence_2022}.
-This diversity and the complex, possibly recombinant, origin of the Omicron clades makes estimating their rate of evolution challenging.
-The lower estimates for these latest clades should therefore be interpreted with care.
-
 \begin{figure*}
@@
 
-Despite these caveats, for almost all major Nextstrain clades, the average divergence increases linearly in time (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C) allowing us to estimate clade specific evolutionary rates for amino acid and silent changes.
+Despite these caveats, for almost all major Nextstrain clades, slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade specific evolutionary rates for amino acid and synonymous changes.
 These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
+In addition, we estimate the rate at which the variants themselves accumulated amino acid and synonymous changes by regressing the number of changes of the variants (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the variants.
+These regressions are shown in thick gray lines in Fig.~\ref{fig:rate_summary}.
 
-Rates of synonymous change are very consistent across variants and also agree with the overall rate of synonymous changes of about 5-8 changes per genome per year, while rates of non-synonymous changes are much more variable (Fig.~\ref{fig:rate_summary}D).
+Rates of synonymous change are very consistent across variants (about 5-8 changes per genome per year) and also agree with the overall rate of synonymous changes of 7.3 changes per genome per year.
+The rates of non-synonymous changes are much more variable (Fig.~\ref{fig:rate_summary}D).
 Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
@@
 Nextstrain clades tend to be defined by long branches leading to a large polytomy.
-It is thus expected that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
+It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
 This effect might be particularly important early on in the pandemic where branches with as little two mutations were used to define new clades.
-We thus also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
+We therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
 The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
 The latter is due to the rapid expansion and subsequent dominance of clade 20A and its descendants which rapidly added four additional mutations \citep{korber_tracking_2020}.
+We argue that this is an early example of an accelerated global rate of evolution due to adaptive evolution, see Discussion.
+
+At the other extreme, some Omicron clades have very low rate estimates which should be interpreted carefully.
+Founder genotypes of Omicron clades 21L (BA.2), 22A (BA.4) and 22B (BA.5) are barely sampled.
+These clades showed considerable diversity, including reversions to the reference, soon after their initial discovery \citep{tegally_emergence_2022}.
+This diversity and the complex, possibly recombinant, origin of the Omicron clades makes estimating their rate of evolution challenging.
 
@@
 
-Above, we found that the rate of synonymous mutations is comparable within and between clades without a strong signal of temporal variation.
+Above, we found that the rate of synonymous mutations is comparable within and between clades without a strong indication that this rate might change over time.
 This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements or splice sites).
-To assess the level of constraint on the SARS-CoV-2 genome more generally, we made use of the ``rare mutations'' annotation by Nextclade.
+To quantify how much of the SARS-CoV-2 genome is constrained how strongly, we made use of the ``rare mutations'' annotation by Nextclade.
 Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
-For each Pango lineage (as determined by Nextclade) \citep{rambaut_dynamic_2020,aksamentov_nextclade_2021}, we count how often these ``rare mutations'' (including reversions to reference) are observed, normalize this against the relative rate of mutation away from the ancestral nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a proxy for mutational tolerance.
+For each Pango lineage (as determined by Nextclade) \citep{rambaut_dynamic_2020,aksamentov_nextclade_2021}, we count how often these ``rare mutations'' (including reversions to reference) are observed.
+This way, for each position in the genome, we obtain the fraction of lineage with minor variation (excluding singletons).
+We normalize this fraction against the relative rate of mutation away from the ancestral nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a semi-quantitative proxy for mutational tolerance.
 
@@
 
-When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed stop-mutations in these genes.
+When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed premature stop-mutations in some of these genes.
 \emph{N} shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
-Rare variation at 3rd positions is very comparable between genes.
-Only \emph{E} shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35.
+Variation at 3rd positions is common and comparable between genes.
+Only \emph{E} shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35 (see Fig.~\ref{fig:fitness_landscape}).
 
@@
 
-Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of 6 months, purifying selection has comparable efficiency and should affect our estimates equally for all clades.
-Sequencing artifacts are expected to increase divergence at all time points or decrease with time as sequencing workflows improve.
+We show that non-synonymous diversification within-clades slowed down during the first six months of the pandemic.
+Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of up to six months, purifying selection has comparable efficiency and should affect our estimates equally for all clades.
+Similarly, sequencing artifacts are expected to increase divergence at all time points and affect synonymous and non-synonymous rate in similar ways.
+The same applies to potential changes in the fidelity of the polymerase.
 Nevertheless, the estimated non-synonymous evolutionary rate of clades circulating in late 2019 and early 2020 is about twice as high as that of clades in 2021 and 2022, while the synonymous rate does not change (see Fig.~\ref{fig:rate_summary} and Table \ref{tab:rates}).
@@
 
-Most of this apparent acceleration of within-variant evolution ceased in mid 2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 per year instead of 12.
+Most of this apparent acceleration of within-variant evolution ceased in mid 2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 instead of 12 changes per year.
 By that time, the number of non-synonymous differences relative to the reference was small (5 in the case of 20E) and it is implausible that this small change would have exhausted the pool of beneficial mutations.
@@
 Such ``diminishing return'' epistasis as been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
-Such global epistasis can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021}.
+Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021}.
 Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments \citep{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.
 
-An evolutionary rate of 6 synonymous changes at around 9000 positions per year corresponds to a per-site mutation rate of $6.6\times 10^{-4}$ per site and year or $1.8\times 10^{-6}$/day, slightly higher than the estimated base-line mutation rate of $1.3\times 10^{-6}$/day \citep{amicone_mutation_2022}.
-The overall rate in variants starting with 20E is in the range of 9-16 changes per year, corresponding to a per site rate of $3 - 5\times 10^{-4}$/site and year, consistent with recent estimates by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
+An evolutionary rate of 6 synonymous changes at around 9700 positions per year corresponds to a per-site mutation rate of $6.2\times 10^{-4}$ per site and year or $1.7\times 10^{-6}$/day, slightly higher than the estimated base-line mutation rate of $1.3\times 10^{-6}$/day \citep{amicone_mutation_2022}.
+The overall evolutionary rate in variants after mid 2020 (starting with 20E) is in the range of 9-16 changes per year, corresponding to a per site rate of $3 - 5\times 10^{-4}$/site and year, consistent with recent estimates by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
 
@@
 This rate is a composite of episode of cryptic accelerated evolution (probably in chronically infected individuals), and regular transmission chains of acute infections.
-This is thus lower than the branch specific rates estimated by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
+This average rate is thus lower than the branch specific rates estimated by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
 However, all clades and variants are compatible with a single ``back-bone'' molecular clock without postulating significant differences in the emergence of different variants.
-This observation of course does not rule out that particular clades and variants emerged in qualitatively different ways, but exponential amplification of lineages with adaptive non-synonymous mutations results in elevated rates of evolution.
+This observation of course does not rule out that particular clades and variants emerged in qualitatively different ways, but does suggest that adaptive evolution resulted in similar rates of mutation accumulation.
 
 In addition to neutral and potentially adaptive mutations, we also quantified purifying selection on the SARS-CoV-2 genome.
-By analyzing the rate mutations that spread only on short times scales within fine-grained Pango lineages, we estimated the level of constraint on different parts of the SARS-CoV-genome.
+By analyzing the rate of mutations that spread only on short times scales within fine-grained Pango lineages, we estimated the level of constraint on different parts of the SARS-CoV-genome.
 The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
@@
 These factors make extrapolation -- whether it is reconstruction of past events or scenarios of future evolution -- uncertain.
-Standard phylogenetic will underestimate that uncertainty.
+Standard phylogenetic or phylodynamic methods will underestimate that uncertainty.
 
 Our analysis here has shown that the majority of non-synonymous changes by which SARS-CoV-2 variants differ from each other are likely of adaptive origin.
-If the excess of non-synonymous changes on the pandemic scale compared to within clade evolution is due to adaptive evolution, the suggests an adaptive rate of about 20 changes per year (plus insertion/deletions).
-This process is likely to continue even in the more and more diverse immunological landscape of the human population in which the virus adapts.
+If the excess of non-synonymous changes on the pandemic scale compared to within clade evolution is due to adaptive evolution, the suggests an adaptive rate of about 20 changes per year (plus insertion/deletions not considered here).
+This process is likely to continue even in the more and more diverse immunological landscape of the human population in which the viral population spread and adapts.
 But it is unclear whether SARS-CoV-2 continues to evolve in a saltatory fashion with repeated emergence of highly mutated variants, or whether we will see a transition to a more gradual adaptive process.
-We currently observe multiple lineage with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), suggesting that shift to a more influenza-like evolution.
+We currently observe multiple lineage with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), suggesting a shift to a more influenza-like evolution.
 
@@
 We used Nextclade data \texttt{2022-07-26T12:00:00Z} with reference sequence MN908947 (Wuhan-Hu-1).
+The codon analysis ignores \emph{ORF9b} which is fully contained in \emph{N}.
 Tabular files with estimates of evolutionary rates, the mutation distribution, and fitness costs are included in the git repository.

### 2022-08-18 manuscript/variant_rates.tex
comments by louis and Jesse

@@
 The rate of synonymous mutations is found to be around 6 changes per year.
-Rates within variants vary little and are compatible with a overall rate.
-In contrast, the rate at which variants accumulate amino acid changes was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
+Rates within variants vary little and are compatible with the overall rate.
+In contrast, the rate at which variants accumulate amino acid changes (non-synonymous mutation) was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
 The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 25 amino acid changes per year.
@@
 Many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
-About half of mutations in other proteins are strongly deleterious and rarely observed, even at low frequency.
+About half of the mutations in other proteins are strongly deleterious and rarely observed, even at low frequency.
 \end{abstract}
@@
 Specific Variants of Concern (VoC) or Interest (VoI) were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
-The branches leading to these variants are characterized by many amino acid changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.
+The branches leading to these variants are characterized by many amino acid-changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.
 
-This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells is common among many RNA viruses and for example well studied in influenza virus A evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
-But adaptive evolution of influenza viruses tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with 10s of novel mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
+This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells is common among many RNA viruses and for example well studied in influenza A virus evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.
+But adaptive evolution of influenza viruses tends to be gradual without large jumps in sequence space, while new variants of SARS-CoV-2 with tens of novel mutations emerged suddenly without intermediate genomes being observed, the most dramatic being the emergence of Omicron in late 2021 \citep{viana_rapid_2022}.
 The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are common in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
@@
 
-The saltatory nature of SARS-CoV-2 evolution implies that evolution within individual variants, clades, or lineages might have different characteristics from the global pattern, which is dominated by atypical bursts of evolution leading to new variants.
-\citet{tay_emergence_2022} and \citet{hill_origins_2022} show that the rate of evolution along these inter-clade branches is up to four-fold higher than the background rate.
+The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within clades or variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who show that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
 Here, we build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
 This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
-This different in evolutionary rate is only seen for non-synonymous changes -- the rate of synonymous evolution within variants is compatible with that seen between variants.
+This difference in evolutionary rate is only seen for non-synonymous changes -- the rate of synonymous evolution within variants is compatible with that seen between variants.
 Furthermore, early variants display more rapid non-synonymous evolution than later variants suggesting more ubiquitous adaptive evolution early on.
@@
     \includegraphics[width=\textwidth]{figures/counts/20C_counts.pdf}
-    \caption{{\bf Within-clade divergence increases linearly with time.} Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
+    \caption{{\bf Within-clade divergence increases linearly with time (example clade 20C).} Top: Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
     The green line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter.
 Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
-Analogous figures for all clades considered are included in the appendix.
-    Bottom: Panels D and E show the prevalence of specific genotypes (D) and specific mutations (E).
-In the case shown, the founder genotype initially dominates and no daughter genotype or mutation dominate.
+    Bottom: Panels D and E show the prevalence of the most common genotypes (D) and mutations (E) during the first 3 months.
+In the case shown (20C), the founder genotype initially dominates and no specific mutation ever reaches high frequency.
+Note that specific genotypes invariably decrease in frequency due to subsequent mutations, while mutations irrespective of their genetic background persist.
 Panel F shows a Poisson model fit to the breakdown of the population into genotypes with different number of mutations over time.
+Analogous figures for other clades are included in the appendix.
     \label{fig:within_clade}}
@@
 
-Evolutionary rates and divergence times are typically estimated using phylogenetic \citep{drummond_relaxed_2006}.
-These methods, however, can not handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
+Evolutionary rates and divergence times are typically estimated using phylogenetic approaches \citep{drummond_relaxed_2006}.
+These methods, however, cannot handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
 Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated, recombinant, contaminated, or otherwise chimeric sequence can fatally distort the analysis, in particular if placed early in tree.
+Problematic sequences can be particularly common when a new variant takes over since sequencing protocols need adjusting \citep{deMaio_issues_2020}.
 
-To circumvent many of the above mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
+To circumvent many of the above-mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
 We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots).
 Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
-These clades represent well defined groups of sequences with little non-vertical evolution within them and we analyze data from each clade separately.
+These clades represent well-defined groups of sequences with little evidence of recombination within them and we analyze data from each clade separately.
 Furthermore, we define a ``founder'' genotype for each clade and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
 This founder sequence is manually curated for each clade considered.
-This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with reference sequence.
+This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence.
 For this reduced set of sequences, we determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
@@
 For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
-To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3.
+To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations.
 
@@
 Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each time bin is weighted with the fourth root of the number of sequences in the bins.
-The exact functional form of this weighing does not have a big influence on the results, but a sublinear weighing helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early.
+The exact functional form of this weighting does not have a big influence on the results, but a sublinear weighting helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early after a new variant emerges.
 Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
@@
 Along each of these lineages, mutation accumulation is independent.
-While not every sequence is an independent sample, the effective number of independent sampled is large and the steadily increasing average divergence allows to estimate the rate robustly.
+While not every sequence is an independent sample, the effective number of independent samples is large and the steadily increasing average divergence allows to estimate the rate robustly.
 
@@
 Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
-The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5days$ and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$.
+The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5$ days and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$ (note that this rate excludes strongly deleterious mutations, see Discussion).
 
@@
 with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
-At each time the number of mutations in the population is expected to Poisson distributed with mean $ut$, while the overall number of cases is $e^{\alpha t}$ or more generally $e^{\int \alpha(t) dt}$ if growth rate or ascertainment varies over time.
+At each time the number of mutations in the population is expected to be Poisson distributed with mean $ut$, while the overall number of cases is $e^{\alpha t}$ or more generally $e^{\int \alpha(t) dt}$ if growth rate or ascertainment varies over time.
 
-For some clades, especially those that are well sampled soon after their emergence, this Poisson model is good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clade}F for clade 20C.
+For some clades, especially those that are well sampled soon after their emergence, this Poisson model is a good fit to diversity accumulation and yields estimates of rates and time of origin that are compatible with the divergence regression, see Fig.~\ref{fig:within_clade}F for clade 20C.
 In this case, the founder genotype initially dominates, but is gradually replaced -- first by single mutant genotypes, then double mutants, and so forth (comp.~Fig.~\ref{fig:within_clade}).
@@
 
-In some cases, notable subclades do become dominant in the early stochastic dynamic.
-This can result in increased within clade divergence, effectively generating an offset to diversity.
+In other cases, notable subclades do become dominant in the early stochastic dynamic.
+This can result in increased within-clade divergence, effectively generating an offset to diversity.
 Extrapolation of the linear fit to zero divergence is thus not necessarily a good estimator of the emergence time of a variant.
-For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder like variants are a minority even in early data.
+For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder-like variants are a minority even in early data.
 
 These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
-In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.
+In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or because they carry adaptive mutations.
 In variant 21K (Omicron, BA.1), a sublineage with mutation \texttt{S:R346K} might have enjoyed a transmission advantage.
@@
     \includegraphics[width=0.8\textwidth]{figures/rate_summary.pdf}
-    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
+    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot or square) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
     In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
     The standard deviation expected based on Poisson statistics is indicated as shaded area.
-    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate inter-clade rates (gray lines).
+    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate ofinter-clade rates (gray lines).
+    In panel D, clades are in alphabetical order, which is similar to their (uncertain) order of emergence.
     \label{fig:rate_summary} }
@@
 
-Despite these caveats, for almost all major Nextstrain clades, slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade specific evolutionary rates for amino acid and synonymous changes.
+Despite these caveats, for almost all major Nextstrain clades, the slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade-specific evolutionary rates for amino acid and synonymous changes.
 These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
@@
 Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
-Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year Fig.~\ref{fig:rate_progression}.
+Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year (see Fig.~\ref{fig:rate_progression}).
 In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.
@@
 It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
-This effect might be particularly important early on in the pandemic where branches with as little two mutations were used to define new clades.
+This effect might be particularly important early on in the pandemic where branches with as few as two mutations were used to define new clades.
 We therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
 The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
-The latter is due to the rapid expansion and subsequent dominance of clade 20A and its descendants which rapidly added four additional mutations \citep{korber_tracking_2020}.
+The latter is due to the rapid expansion and subsequent dominance of clade 20A, the first clade with spike mutation $D614G$, and its descendants which rapidly fixed four additional mutations \citep{korber_tracking_2020}.
 We argue that this is an early example of an accelerated global rate of evolution due to adaptive evolution, see Discussion.
@@
 At the other extreme, some Omicron clades have very low rate estimates which should be interpreted carefully.
-Founder genotypes of Omicron clades 21L (BA.2), 22A (BA.4) and 22B (BA.5) are barely sampled.
+Founder genotypes of Omicron clades 21L (BA.2), 22A (BA.4) and 22B (BA.5) are rarely sampled.
 These clades showed considerable diversity, including reversions to the reference, soon after their initial discovery \citep{tegally_emergence_2022}.
-This diversity and the complex, possibly recombinant, origin of the Omicron clades makes estimating their rate of evolution challenging.
+This diversity and the complex, possibly recombinant, origin of the Omicron clades makes estimating their rates of evolution challenging.
 
@@
     20E &     14.04 &     5.65 &      8.39 &       13 &       5 &        8 \\
-    20H &     15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
-    20I &     11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
-    20J &     12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
-    21D &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
-    21G &     12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
-    21H &     18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
-    21I &     16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
-    21J &     14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
-    21K &     13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
-    21L &      9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
-    22A &      7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
-    22B &     11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
+    20H (Beta) &     15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
+    20I (Alpha) &     11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
+    20J (Gamma) &     12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
+    21D (Eta) &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
+    21G (Lambda) &     12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
+    21H (Mu) &     18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
+    21I (Delta) &     16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
+    21J (Delta) &     14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
+    21K (Omicron) &     13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
+    21L (Omicron) &      9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
+    22A (Omicron) &      7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
+    22B (Omicron) &     11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
            \hline
@@
     \caption{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades.
-Non-synonymous rate estimates are highest for clades 19A - 20C.
+Non-synonymous rate estimates are highest for clades 19A to 20C that emerged early in the pandemic.
     \label{fig:rate_progression} }
@@
 This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements or splice sites).
-To quantify how much of the SARS-CoV-2 genome is constrained how strongly, we made use of the ``rare mutations'' annotation by Nextclade.
+To quantify how much of the SARS-CoV-2 genome is constrained and how strongly purifying selection operates on different genomic regions, we made use of the ``rare mutations'' annotation provided by Nextclade.
 Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
-For each Pango lineage (as determined by Nextclade) \citep{rambaut_dynamic_2020,aksamentov_nextclade_2021}, we count how often these ``rare mutations'' (including reversions to reference) are observed.
-This way, for each position in the genome, we obtain the fraction of lineage with minor variation (excluding singletons).
+For each Pango lineage (as determined by Nextclade) \citep{rambaut_dynamic_2020,aksamentov_nextclade_2021}, we count how often these ``rare mutations'' (including reversions to the reference) are observed.
+This way, for each position in the genome, we obtain the fraction of the lineage with minor variation (excluding singletons).
 We normalize this fraction against the relative rate of mutation away from the ancestral nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a semi-quantitative proxy for mutational tolerance.
@@
 Simply splitting mutations into those that occur in 1st, 2nd, and 3rd positions of codons already reveals strong signatures of purifying selection, see Fig.~\ref{fig:fitness_costs}.
-Between 15-20\% of first and second positions in codons show almost no variation, while half of these sites are less variable then the most constrained 10\% of third positions.
-The median of variation at 3rd position is more than double that at other 1st and 2nd positions.
+Between 15 and 20\% of first and second positions in codons show almost no variation, while half of these sites are less variable than the most constrained 10\% of third positions.
+The median of variation at 3rd positions is more than double that at 1st and 2nd positions.
 
-When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and \ref{fig:fitness_landscape}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed premature stop-mutations in some of these genes.
+When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and Supp.~Fig.~\ref{fig:fitness_landscape}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed premature stop-mutations in some of these genes.
 \emph{N} shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
@@
     Solid lines show sliding window smoothing of estimated mutational tolerance at 1st, 2nd, and 3rd positions for a window size of 20 and 7 (faint lines) sites.
-    The markers below the lines are a qualitative indication of variable positions across SARS-CoV-2 diversity.
+    The markers at the bottom of the panels show qualitatively how many Pango lineages are differ at this position from the Wuhan-Hu-1 reference sequence.
     \label{fig:fitness_landscape}}
@@
 \section*{Discussion}
-The inferred evolutionary rate of RNA viruses often decreases with the time scale across which is it estimated \citep{wertheim_purifying_2011,ghafari_prisoner_2021}.
+The inferred evolutionary rate of RNA viruses often decreases with the time scale across which it is estimated \citep{wertheim_purifying_2011,ghafari_prisoner_2021}.
 This effect can be particularly pronounced at the beginning of an outbreak following a host switch and has been attributed to incomplete purifying selection or methodological issues leading to inflated measures of diversity \citep{meyer_time_2015,ghafari_purifying_2022}.
@@
 We show that non-synonymous diversification within-clades slowed down during the first six months of the pandemic.
-Since we analyzed evolution within short lived clades of SARS-CoV-2 over a span of up to six months, purifying selection has comparable efficiency and should affect our estimates equally for all clades.
+Since we analyzed evolution within short-lived clades of SARS-CoV-2 over a span of up to six months, purifying selection has comparable efficiency and should affect our estimates equally for all clades.
 Similarly, sequencing artifacts are expected to increase divergence at all time points and affect synonymous and non-synonymous rate in similar ways.
@@
 
-Most of this apparent acceleration of within-variant evolution ceased in mid 2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 instead of 12 changes per year.
+Most of this apparent acceleration of within-variant evolution ceased in mid-2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 instead of 12 changes per year.
 By that time, the number of non-synonymous differences relative to the reference was small (5 in the case of 20E) and it is implausible that this small change would have exhausted the pool of beneficial mutations.
-So why would the rate of adaptation evolution slow down?
-Viruses with the S:D614G showed a faster replication kinetics compared to earlier variants \citep{korber_tracking_2020}.
+So why would the rate of adaptive evolution slow down?
+Viruses with the S:D614G mutation showed faster replication kinetics compared to earlier variants \citep{korber_tracking_2020}.
 The virus might therefore be operating closer to the maximal capacity at which cells can produce virions, reducing the scope for further optimization.
-Such ``diminishing return'' epistasis as been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
+Such ``diminishing returns'' epistasis has been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
 Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021}.
-Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments \citep{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.
+Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments by \citet{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.
 
-An evolutionary rate of 6 synonymous changes at around 9700 positions per year corresponds to a per-site mutation rate of $6.2\times 10^{-4}$ per site and year or $1.7\times 10^{-6}$/day, slightly higher than the estimated base-line mutation rate of $1.3\times 10^{-6}$/day \citep{amicone_mutation_2022}.
-The overall evolutionary rate in variants after mid 2020 (starting with 20E) is in the range of 9-16 changes per year, corresponding to a per site rate of $3 - 5\times 10^{-4}$/site and year, consistent with recent estimates by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
+An evolutionary rate of 6 synonymous changes per year at around 9700 positions corresponds to a per-site evolutionary rate of $6.2\times 10^{-4}$ per site and year or $1.7\times 10^{-6}$/day, slightly higher than the estimated base-line mutation rate of $1.3\times 10^{-6}$/day \citep{amicone_mutation_2022}.
+The overall evolutionary rate in variants after mid 2020 (starting with 20E) is in the range of 9-16 changes per year, corresponding to a per site rate of $3 - 5\times 10^{-4}$/site/year, consistent with recent estimates by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
 
 When considering only the putative founder genotype and date of origin of each variant, all variants so far are compatible with a back-bone evolutionary rate of 32 changes per year, corresponding to an per site rate of around $10^{-3}$/site/year (see Fig.~\ref{fig:rate_summary}).
-This rate is a composite of episode of cryptic accelerated evolution (probably in chronically infected individuals), and regular transmission chains of acute infections.
-This average rate is thus lower than the branch specific rates estimated by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
+This rate is a composite of episode of cryptic accelerated evolution (possibly in chronically infected individuals), and regular transmission chains of acute infections.
+This estimate for this average rate is thus consistent with results by \citet{hill_origins_2022} and \citet{tay_emergence_2022}, who estimated even higher rates specifically for the branches that gave rise to variants of concern.
 However, all clades and variants are compatible with a single ``back-bone'' molecular clock without postulating significant differences in the emergence of different variants.
@@
 The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
-The major open reading frames \emph{ORF1ab, S, N, E, M} show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
-This result is consistent with similar analysis in HIV-1 that also show small fitness costs for most synonymous mutations, while half of non-synonymous mutations are so deleterious that they are not even observed on short times scales \citep{zanini_vivo_2017}.
+The major open reading frames \emph{ORF1ab, S, N, E} and \emph{M} show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
+Purifying selection analyzed there is operating over the time rare lineage circulate, which is typically a few weeks.
+On shorter time scales, more strongly deleterious mutations will be observed.
+This likely explains the observation that the rate of intra-host diversification during acute infections is estimated to about twice as large ($1.1 \times 10^{-3}$ per site and year, \citep{braun_acute_2021) as the base line rate estimated here.
+
+The degree of purifying selection is consistent with a similar analysis in HIV-1 that also show small fitness costs for most synonymous mutations, while half of the non-synonymous mutations are so deleterious that they are not even observed on short timescales \citep{zanini_vivo_2017}.
 The remaining SARS-CoV-2 ORFs show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
-This lack of strong purifying selection in these ORFs does not necessarily imply that they don't matter or are not a possible locus of adaptation -- but the exact amino acid sequence does not seem essential for their function.
+This lack of strong purifying selection in these ORFs does not necessarily imply that they don't matter or are not a possible loci of adaptation -- but the exact amino acid sequence does not seem essential for their function.
 Only a few regions show a clear signal of conservation at 3rd positions, notably a central region of \emph{E} and the ribosome slippage region at the beginning of \emph{ORF1b}.
@@
 
-Our analysis here has shown that the majority of non-synonymous changes by which SARS-CoV-2 variants differ from each other are likely of adaptive origin.
-If the excess of non-synonymous changes on the pandemic scale compared to within clade evolution is due to adaptive evolution, the suggests an adaptive rate of about 20 changes per year (plus insertion/deletions not considered here).
+If the excess of non-synonymous changes on the pandemic scale compared to within clade evolution is due to adaptive evolution, it would imply an adaptive rate of about 20 changes per year (plus insertion/deletions not considered here).
 This process is likely to continue even in the more and more diverse immunological landscape of the human population in which the viral population spread and adapts.
 But it is unclear whether SARS-CoV-2 continues to evolve in a saltatory fashion with repeated emergence of highly mutated variants, or whether we will see a transition to a more gradual adaptive process.
-We currently observe multiple lineage with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), suggesting a shift to a more influenza-like evolution.
+We currently observe multiple lineages with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), suggesting a shift to a more influenza-like evolution.
 

### 2022-08-18 scripts/combine_fits.py
edits to graphs

@@ -1,10 +1,10 @@
 import argparse
-from turtle import fillcolor
 import numpy as np
 import pandas as pd
 from scipy.stats import linregress
 import matplotlib as mpl
 mpl.rcParams['axes.formatter.useoffset'] = False
+from root_to_tip import add_panel_label
 
 import matplotlib.pyplot as plt
 import seaborn as sns
@@ -25,9 +25,10 @@ if __name__=="__main__":
     inter_clade_rates = {}
     fig, axs = plt.subplots(2,2, figsize=(12,12))
     xticks = [2020,2020.5,2021,2021.5,2022,2022.5]
-    for ax, mut_type, ax_label in zip(axs.flatten(),['nuc', 'aa', 'syn'],
-                        ['total divergence', 'amino acid divergence', 'synonymous divergence']):
+    for ax, mut_type, ax_label, panel in zip(axs.flatten(),['nuc', 'aa', 'syn'],
+                        ['total divergence', 'amino acid divergence', 'synonymous divergence'], ['A', 'B', 'C']):
         ax.set_ylabel(ax_label, fontsize=fs)
+        add_panel_label(ax, panel, fs=fs*1.8)
         inter_clade = []
         ci=0
         for clade, row in rates.iterrows():
@@ -52,7 +53,7 @@ if __name__=="__main__":
         ax.fill_between(x, y+std_dev, np.maximum(0, y-std_dev), fc='k', lw=3, alpha=0.1, ec=None)
         ax.set_xlim(x.min(), x.max())
         ax.set_ylim(0)
-        ax.text( 0.6, 0.1,f"rate: {reg.slope:1.1f}/year", fontsize=fs, transform=ax.transAxes)
+        ax.text( 0.6, 0.06,f"overall rate: {reg.slope:1.1f}/year", fontsize=fs, transform=ax.transAxes)
         if mut_type == 'aa': ax.legend(ncol=2)
         ax.set_xticks(xticks, [str(x) for x in xticks], fontsize=fs)
 
@@ -71,6 +72,7 @@ if __name__=="__main__":
     axs[-1,-1].set_xticks([0,1,2], ['total', 'amino acid', 'synonymous'], rotation=20,
                           ha='center', fontsize=fs)
     axs[-1,-1].set_ylim(0)
+    add_panel_label(axs[-1,-1], 'D', fs=fs*1.8)
 
     if args.output_plot:
         plt.savefig(args.output_plot)

### 2022-08-18 scripts/plot_genotype_counts.py

@@ -73,6 +73,7 @@ if __name__=="__main__":
     fig, axs = plt.subplots(1,3, figsize = (18,6))
 
     ax = axs[2]
+    ax.set_title("mutations per genome", fontsize=1.2*fs)
     # ax.plot(dates, counts['all_samples'], lw=3, c='k', alpha=0.3)
     total = np.array(counts['all_samples'])
     ind = total>0
@@ -86,11 +87,13 @@ if __name__=="__main__":
     ax.set_ylim(8e-5, 2)
 
     ax = axs[1]
+    ax.set_title("mutation frequencies", fontsize=1.2*fs)
     ax.plot(dates, total, lw=3, c='k', alpha=0.3)
     for m in sorted(counts['mutations'].keys(), key=lambda x:int(x[1:-1])):
         ax.plot(dates, counts['mutations'][m], '-o', label=f'{m}')
 
     ax = axs[0]
+    ax.set_title("genotype frequencies", fontsize=1.2*fs)
     ax.plot(dates, total, lw=3, c='k', alpha=0.3)
     for m in sorted(counts['genotypes'].keys(), key=lambda x: len(x)):
         ax.plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")

### 2022-08-19 scripts/combine_fits.py
add by gene figure to rates script

@@ -19,6 +19,7 @@ if __name__=="__main__":
     parser.add_argument('--rate-table', type=str, required=True, help="input data")
     parser.add_argument('--output-plot', type=str, help="plot file")
     parser.add_argument('--output-plot-rates', type=str, help="plot file")
+    parser.add_argument('--output-plot-rates-genes', type=str, help="plot file")
     args = parser.parse_args()
 
     rates = pd.read_csv(args.rate_table, sep='\t', index_col='clade')
@@ -53,7 +54,7 @@ if __name__=="__main__":
         ax.fill_between(x, y+std_dev, np.maximum(0, y-std_dev), fc='k', lw=3, alpha=0.1, ec=None)
         ax.set_xlim(x.min(), x.max())
         ax.set_ylim(0)
-        ax.text( 0.6, 0.06,f"overall rate: {reg.slope:1.1f}/year", fontsize=fs, transform=ax.transAxes)
+        ax.text( 0.4, 0.06,f"overall rate: {reg.slope:1.1f}/year", fontsize=fs, transform=ax.transAxes)
         if mut_type == 'aa': ax.legend(ncol=2)
         ax.set_xticks(xticks, [str(x) for x in xticks], fontsize=fs)
 
@@ -90,5 +91,21 @@ if __name__=="__main__":
 
     if args.output_plot_rates:
         plt.savefig(args.output_plot_rates)
+    else:
+        plt.show()
+
+    plt.figure()
+    plt.plot(rates["aa_rate"], 'o-', label='Overall acid rate')
+    plt.plot(rates["spike_rate"], 's-', label='spike protein')
+    plt.plot(rates["orf1ab_rate"], 'd-', label='ORF1ab')
+    plt.plot(rates["aa_rate"] - rates["spike_rate"] - rates["orf1ab_rate"],
+             'v-', label='other ORFs')
+    plt.ylabel('rate estimate [subs/y]')
+    plt.legend()
+    plt.xticks(range(len(rates)), rates.index, rotation=60, ha='right')
+    plt.tight_layout()
+
+    if args.output_plot_rates_genes:
+        plt.savefig(args.output_plot_rates_genes)
     else:
         plt.show()

### 2022-08-19 scripts/root_to_tip.py
add ENM rate to script

@@ -46,8 +46,9 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
                                                      if all([y not in clade_gt['nuc'], int(y[1:-1])>150,  int(y[1:-1])<29753])] if x else [])
     # define "with-in clade substitutions"
     d["intra_aaSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y] if x else [])
-    d["intra_SpikeSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y and y[0]=='S'] if x else [])
-    d["intra_ORF1Substitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y and y[:4]=='ORF1'] if x else [])
+    d["intra_SpikeSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0]=='S'] if x else [])
+    d["intra_ORF1Substitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:4]=='ORF1'] if x else [])
+    d["intra_ENMSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0] in ['E','N','M'] if x else [])
 
     if swap_root:
         muts = [("C8782T","T8782C"), ("T28144C","C28144T")]
@@ -66,6 +67,7 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
     d["aaDivergence"] = d.intra_aaSubstitutions.apply(lambda x: len(x))
     d["spikeDivergence"] = d.intra_SpikeSubstitutions.apply(lambda x: len(x))
     d["orf1Divergence"] = d.intra_ORF1Substitutions.apply(lambda x: len(x))
+    d["enmDivergence"] = d.intra_ENMSubstitutions.apply(lambda x: len(x))
     d["synDivergence"] = d["divergence"] - d["aaDivergence"]
 
 
@@ -186,6 +188,7 @@ if __name__=="__main__":
     regression_clean_syn = regression_by_week(filtered_data.loc[ind], "synDivergence")
[…]

### 2022-08-19 scripts/root_to_tip.py
fix

@@ -48,7 +48,7 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
     d["intra_aaSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y] if x else [])
     d["intra_SpikeSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0]=='S'] if x else [])
     d["intra_ORF1Substitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:4]=='ORF1'] if x else [])
-    d["intra_ENMSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0] in ['E','N','M'] if x else [])
+    d["intra_ENMSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0] in ['E','N','M']] if x else [])
 
     if swap_root:
         muts = [("C8782T","T8782C"), ("T28144C","C28144T")]

### 2022-08-19 manuscript/appendix.tex
edits to ms
add figures to repo

--- 2022-08-06 manuscript/appendix.tex
@@
 \renewcommand{\tablename}{Table S}
+
+\begin{figure}[h]
+    \includegraphics[width=0.5\textwidth]{figures/rate_progression_by_gene.pdf}
+    \caption{{\bf Within-clade non-synonymous evolution in different parts of the genome.}
+    The decreasing trend in the within-clade non-synonymous rate of evolution is most evident in the largest \emph{ORF1ab}.
+    Accessory proteins and \emph{N,E,M} show a weak trend.
+    \label{fig:rate_progression_by_gene}}
+\end{figure}
 

### 2022-08-19 manuscript/variant_rates.tex
add orf1ab
add figures to repo

@@
 The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within clades or variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who show that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
-Here, we build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
+Here, we\footnote[1]{I and the computer} build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
 This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
@@
         \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
-        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
-        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
+        \frac{dm_1}{dt} &= u m_0 + (\alpha - u) m_1 \\
+        \frac{dm_2}{dt} &= u m_1 + (\alpha - u) m_2 \\
         \cdots
@@
 In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.
+The spike protein does not contribute to this decreasing trend in the rate of within-clade non-synonymous evolution, which is most evident in \emph{ORF1ab} and to a lesser degree in accessory proteins and \emph{N} (see Supp.Fig.~\ref{fig:rate_progression_by_gene}).
 
@@
 On shorter time scales, more strongly deleterious mutations will be observed.
-This likely explains the observation that the rate of intra-host diversification during acute infections is estimated to about twice as large ($1.1 \times 10^{-3}$ per site and year, \citep{braun_acute_2021) as the base line rate estimated here.
+This likely explains the observation that the rate of intra-host diversification during acute infections is estimated to about twice as large ($1.1 \times 10^{-3}$ per site and year, \citep{braun_acute_2021}) as the base line rate estimated here.
 
@@
 We used Nextclade data \texttt{2022-07-26T12:00:00Z} with reference sequence MN908947 (Wuhan-Hu-1).
-The codon analysis ignores \emph{ORF9b} which is fully contained in \emph{N}.
+The codon analysis ignores \emph{ORF9b} which is fully contained in \emph{N} but in a different reading frame.
 Tabular files with estimates of evolutionary rates, the mutation distribution, and fitness costs are included in the git repository.
@@
 \section*{Acknowledgements}
-We gratefully acknowledge the work by 1000s of scientists and healthcare professionals that collect the samples, generate, and share the sequence data on which this analysis is built.
-We are also grateful for stimulating discussions with Cornelius Roemer and Emma Hodcroft on many aspects of SARS-CoV-2 evolution.
+I gratefully acknowledge the work by 1000s of scientists and healthcare professionals that collect the samples, generate, and share the sequence data on which this analysis is built.
+I am also grateful to Jesse Bloom and Louis du Plessis provided valuable feed-back on the manuscript and to Cornelius Roemer and Emma Hodcroft for stimulating discussions on many aspects of SARS-CoV-2 evolution.
 

### 2022-08-19 scripts/combine_fits.py

@@ -98,7 +98,8 @@ if __name__=="__main__":
     plt.plot(rates["aa_rate"], 'o-', label='Overall acid rate')
     plt.plot(rates["spike_rate"], 's-', label='spike protein')
     plt.plot(rates["orf1ab_rate"], 'd-', label='ORF1ab')
-    plt.plot(rates["aa_rate"] - rates["spike_rate"] - rates["orf1ab_rate"],
+    plt.plot(rates["enm_rate"], 'd-', label='E,M,N')
+    plt.plot(rates["aa_rate"] - rates["spike_rate"] - rates["orf1ab_rate"]- rates["enm_rate"],
              'v-', label='other ORFs')
     plt.ylabel('rate estimate [subs/y]')
     plt.legend()

### 2022-08-20 manuscript/variant_rates.tex
add some methods

@@
     \includegraphics[width=0.8\textwidth]{figures/rate_summary.pdf}
-    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot or square) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
+    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B\&C show the estimated divergence of the founder genotype of each clade (big dot or square) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
     In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
@@
 
+The fitness costs of each positions are estimated from the Nextclade columns \texttt{unlabeledSubstitutions} and \texttt{reversionSubstitutions} for each sequence that has QC-status \texttt{good}.
+For each Pango lineage, we count how often each specific mutation was observed.
+We then calculate how many lineages show minor variation for each mutation.
+To avoid loss of signal through sequencing errors, we drop lineages at which only a single mutation
+To turn this into a measure of mutational tolerance, we normalize the number of lineages with minor variation by the rate at which mutation rate of the ancestral base at this position.
+
 \section*{Acknowledgements}

### 2022-08-22 manuscript/variant_rates.tex
fix typo

@@
 Continued evolution and adaptation of SARS-CoV-2 has lead to more transmissible and immune-evasive variants with profound impact on the course of the pandemic.
-Here we analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes.
-We estimate separate rates for evolution within clades, that is well defined mono-phyletic groups with gradual evolution, and for the pandemic overall.
+Here I analyze the evolution of the virus over 2.5 years since its emergence and estimate rates of evolution for synonymous and non-synonymous changes separately for evolution within clades -- well defined mono-phyletic groups with gradual evolution -- and for the pandemic overall.
 The rate of synonymous mutations is found to be around 6 changes per year.
-Rates within variants vary little and are compatible with the overall rate.
+Synonymous rates within variants vary little from variant to variant and are compatible with the overall rate.
 In contrast, the rate at which variants accumulate amino acid changes (non-synonymous mutation) was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
 The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 25 amino acid changes per year.
-This 2-fold higher rate indicates that adaptive evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains.
-We further quantify the spectrum of mutations and purifying selection in different SARS-CoV proteins.
+This 2-fold higher rate indicates that the evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains and likely dominated by adaptive evolution.
+I further quantify the spectrum of mutations and purifying selection in different SARS-CoV-2 proteins.
 Many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
-About half of the mutations in other proteins are strongly deleterious and rarely observed, even at low frequency.
+About half of the mutations in other proteins are strongly deleterious and rarely observed, not even at low frequency.
 \end{abstract}
@@
 
-The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within clades or variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who show that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
-Here, we\footnote[1]{I and the computer} build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
+The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within clades or variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who showed that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
+Here, I build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
 This comparison reveals a consistent dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
@@
 Furthermore, early variants display more rapid non-synonymous evolution than later variants suggesting more ubiquitous adaptive evolution early on.
-We further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.
+I further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.
 
@@
 
-To circumvent many of the above-mentioned problems and still use the majority of the available data, we use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
-We first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots).
+To circumvent many of the above-mentioned problems and still use the majority of the available data, I use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
+I first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots).
 Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
-These clades represent well-defined groups of sequences with little evidence of recombination within them and we analyze data from each clade separately.
-Furthermore, we define a ``founder'' genotype for each clade and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
+These clades represent well-defined groups of sequences with little evidence of recombination within them and are analyzed separately.
+For each clade, I define a ``founder'' genotype and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
 This founder sequence is manually curated for each clade considered.
 This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence.
-For this reduced set of sequences, we determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
+For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
 The latter step is done for nucleotide changes as well as for amino acid changes.
@@
 For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
-To exclude these outliers, we perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations.
+To exclude these outliers, I perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations.
 
-After removing these outliers, we now bin the data by calendar week and determine the mean and standard deviation in each bin.
+After removing these outliers, the data are binned by calendar week.
 Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each time bin is weighted with the fourth root of the number of sequences in the bins.
@@
 
-Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suited to estimate evolutionary rates in general.
+Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suitable to estimate evolutionary rates in general.
 In particular, confidence intervals are difficult to obtain.
@@
 with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
-At each time the number of mutations in the population is expected to be Poisson distributed with mean $ut$, while the overall number of cases is $e^{\alpha t}$ or more generally $e^{\int \alpha(t) dt}$ if growth rate or ascertainment varies over time.
+At time $t$ after the emergence of the variant, the number of mutations in the population is expected to be Poisson distributed with mean $ut$.
+The overall number of cases is $e^{\alpha t}$ or more generally $e^{\int \alpha(t) dt}$ if growth rate or ascertainment varies over time.
 
@@
 
-In other cases, notable subclades do become dominant in the early stochastic dynamic.
+In other cases, notable subclades did become dominant during the early stochastic dynamic.
 This can result in increased within-clade divergence, effectively generating an offset to diversity.
@@
 \begin{figure*}
-    \includegraphics[width=0.8\textwidth]{figures/rate_summary.pdf}
+    \includegraphics[width=0.9\textwidth]{figures/rate_summary.pdf}
     \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B\&C show the estimated divergence of the founder genotype of each clade (big dot or square) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively.
@@
     In panel D, clades are in alphabetical order, which is similar to their (uncertain) order of emergence.
+    The red outlier is `composite' clade 19++ with inflated rates due to adaptive mutations.
     \label{fig:rate_summary} }
@@
 
-Despite these caveats, for almost all major Nextstrain clades, the slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade-specific evolutionary rates for amino acid and synonymous changes.
+Despite these caveats, for almost all Nextstrain clades, the slope at which diversity increases is robust (similar patterns as Fig.~\ref{fig:within_clade}A-C for clade 20C), allowing us to estimate clade-specific evolutionary rates for amino acid and synonymous changes.
 These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
-In addition, we estimate the rate at which the variants themselves accumulated amino acid and synonymous changes by regressing the number of changes of the variants (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the variants.
+In addition, I estimated the rate at which the variants themselves accumulated amino acid and synonymous changes by regressing the number of changes of the variants (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the variants.
 These regressions are shown in thick gray lines in Fig.~\ref{fig:rate_summary}.
@@
 This effect might be particularly important early on in the pandemic where branches with as few as two mutations were used to define new clades.
-We therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
+I therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
 The estimates for composite clades 19B+ and 20A+ are consistent with the estimates of the individual clades, while apparent rates of 19B++ are considerably higher.
 The latter is due to the rapid expansion and subsequent dominance of clade 20A, the first clade with spike mutation $D614G$, and its descendants which rapidly fixed four additional mutations \citep{korber_tracking_2020}.
-We argue that this is an early example of an accelerated global rate of evolution due to adaptive evolution, see Discussion.
+This is an early example of an accelerated global rate of evolution due to adaptive evolution, see Discussion.
 
@@
 
-Above, we found that the rate of synonymous mutations is comparable within and between clades without a strong indication that this rate might change over time.
+As described above, the rate of synonymous mutations is comparable within and between clades without a strong indication that this rate might have changed over time.
 This is expected, as synonymous positions are rarely a locus of adaptation and tend to have a small effects on fitness in large parts of the genomes of RNA viruses \citep{zanini_vivo_2017} (outside of specific regions with important RNA elements or splice sites).
-To quantify how much of the SARS-CoV-2 genome is constrained and how strongly purifying selection operates on different genomic regions, we made use of the ``rare mutations'' annotation provided by Nextclade.
+To quantify how much of the SARS-CoV-2 genome is constrained and how strongly purifying selection operates on different genomic regions, I made use of the ``rare mutations'' annotation provided by Nextclade.
 Nextclade attaches each sequence to a reference tree and determines by which mutations it differs from the attachment point.
-For each Pango lineage (as determined by Nextclade) \citep{rambaut_dynamic_2020,aksamentov_nextclade_2021}, we count how often these ``rare mutations'' (including reversions to the reference) are observed.
-This way, for each position in the genome, we obtain the fraction of the lineage with minor variation (excluding singletons).
-We normalize this fraction against the relative rate of mutation away from the ancestral nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a semi-quantitative proxy for mutational tolerance.
+For each Pango lineage (as determined by Nextclade) \citep{rambaut_dynamic_2020,aksamentov_nextclade_2021}, I count how often these ``rare mutations'' (including reversions to the reference) are observed.
+This way, for each position in the genome, one obtains the fraction of lineages with minor variation (excluding singletons).
+I normalize this fraction against the relative rate of mutation away from the ancestral nucleotide (see Supp.~Fig.~\ref{fig:mutation_distribution}), and use this as a semi-quantitative proxy of mutational tolerance.
 
-Simply splitting mutations into those that occur in 1st, 2nd, and 3rd positions of codons already reveals strong signatures of purifying selection, see Fig.~\ref{fig:fitness_costs}.
+Simply splitting the genome into at 1st, 2nd, and 3rd positions of codons already reveals strong signatures of purifying selection, see Fig.~\ref{fig:fitness_costs}.
 Between 15 and 20\% of first and second positions in codons show almost no variation, while half of these sites are less variable than the most constrained 10\% of third positions.
@@
 
-When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and Supp.~Fig.~\ref{fig:fitness_landscape}), the most constrained are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed premature stop-mutations in some of these genes.
+When split by open reading frame (see Supp.~Fig.~\ref{fig:fitness_costs_supp} and Supp.~Fig.~\ref{fig:fitness_landscape}), the most constrained regions are \emph{ORF1b} and \emph{M}, while \emph{ORF3a, ORF6, ORF7a, ORF7b}, and \emph{ORF8} show little evidence of constraint, consistent with frequently observed premature stop-mutations in some of these genes.
 \emph{N} shows an intermediate pattern, possibly reflecting its mix of structured and unstructured regions.
@@
 
-We show that non-synonymous diversification within-clades slowed down during the first six months of the pandemic.
-Since we analyzed evolution within short-lived clades of SARS-CoV-2 over a span of up to six months, purifying selection has comparable efficiency and should affect our estimates equally for all clades.
+Here, I showed that the deceleration during the first six months of the pandemic is only observed for non-synonymous mutations.
+Since I analyzed evolution within short-lived clades of SARS-CoV-2 over a span of up to six months, purifying selection has comparable efficiency and should affect our estimates equally for all clades.
 Similarly, sequencing artifacts are expected to increase divergence at all time points and affect synonymous and non-synonymous rate in similar ways.
@@
 
-Most of this apparent acceleration of within-variant evolution ceased in mid-2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 instead of 12 changes per year.
-By that time, the number of non-synonymous differences relative to the reference was small (5 in the case of 20E) and it is implausible that this small change would have exhausted the pool of beneficial mutations.
+Most of this apparent initial acceleration of within-variant evolution ceased in mid-2020 and even early variants like 20E accumulated non-synonymous changes at a rate of about 6 instead of 12 changes per year.
+By that time, the number of non-synonymous differences relative to the root of the tree was small (5 in the case of 20E) and it is implausible that this small change would have exhausted the pool of beneficial mutations.
 So why would the rate of adaptive evolution slow down?
+Maybe it is just chance, after all it is only clades 19A and 19B where the rate is substantially higher.
+Another possible explanation could be \emph{diminishing returns} epistasis.
 Viruses with the S:D614G mutation showed faster replication kinetics compared to earlier variants \citep{korber_tracking_2020}.
 The virus might therefore be operating closer to the maximal capacity at which cells can produce virions, reducing the scope for further optimization.
-Such ``diminishing returns'' epistasis has been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
+Other mutations then have smaller benefits, rise in frequency more slowly, and the effect of adaptation does not manifest itself over the life time of the clades studied here.
+Such diminishing returns epistasis has been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
 Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021}.
-Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments by \citet{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate.
+Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments by \citet{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate but had selective origin.
 
-An evolutionary rate of 6 synonymous changes per year at around 9700 positions corresponds to a per-site evolutionary rate of $6.2\times 10^{-4}$ per site and year or $1.7\times 10^{-6}$/day, slightly higher than the estimated base-line mutation rate of $1.3\times 10^{-6}$/day \citep{amicone_mutation_2022}.
-The overall evolutionary rate in variants after mid 2020 (starting with 20E) is in the range of 9-16 changes per year, corresponding to a per site rate of $3 - 5\times 10^{-4}$/site/year, consistent with recent estimates by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
+An evolutionary rate of 6 synonymous changes per year at around 9700 positions corresponds to a per-site evolutionary rate of $6.2\times 10^{-4}$/site/year or $1.7\times 10^{-6}$/day, slightly higher than the estimated base-line mutation rate of $1.3\times 10^{-6}$/day \citep{amicone_mutation_2022}.
+The total evolutionary rate within variants after mid 2020 (starting with 20E) is in the range of 9-16 changes per year, corresponding to a per site rate of $3 - 5\times 10^{-4}$/site/year, consistent with recent estimates by \citet{hill_origins_2022} and \citet{tay_emergence_2022}.
 
@@
 This rate is a composite of episode of cryptic accelerated evolution (possibly in chronically infected individuals), and regular transmission chains of acute infections.
-This estimate for this average rate is thus consistent with results by \citet{hill_origins_2022} and \citet{tay_emergence_2022}, who estimated even higher rates specifically for the branches that gave rise to variants of concern.
-However, all clades and variants are compatible with a single ``back-bone'' molecular clock without postulating significant differences in the emergence of different variants.
-This observation of course does not rule out that particular clades and variants emerged in qualitatively different ways, but does suggest that adaptive evolution resulted in similar rates of mutation accumulation.
+This estimate for the average rate is thus not inconsistent with results by \citet{hill_origins_2022} and \citet{tay_emergence_2022}, who estimated even higher rates specifically for the branches that gave rise to variants of concern.
+Different clades and variants probably emerged in different ways under different circumstances.
+Nevertheless, all clades and variants are compatible with a single ``back-bone'' molecular clock that runs substantially faster than the ``within variant clock''.
+This accelerated back-bone clock is likely driven by exponential amplification of beneficial mutations.
 
-In addition to neutral and potentially adaptive mutations, we also quantified purifying selection on the SARS-CoV-2 genome.
-By analyzing the rate of mutations that spread only on short times scales within fine-grained Pango lineages, we estimated the level of constraint on different parts of the SARS-CoV-genome.
+In addition to neutral and potentially adaptive mutations, I also quantified purifying selection on the SARS-CoV-2 genome.
+By analyzing the rate of mutations that spread only on short times scales within fine-grained Pango lineages, I estimated the level of constraint on different parts of the SARS-CoV-genome.
 The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
@@
 On shorter time scales, more strongly deleterious mutations will be observed.
-This likely explains the observation that the rate of intra-host diversification during acute infections is estimated to about twice as large ($1.1 \times 10^{-3}$ per site and year, \citep{braun_acute_2021}) as the base line rate estimated here.
+The degree of purifying selection is consistent with a similar analysis in HIV-1 that also show small fitness costs for most synonymous mutations, while half of the non-synonymous mutations are so deleterious that they are not even observed on short timescales \citep{zanini_vivo_2017}.
 
-The degree of purifying selection is consistent with a similar analysis in HIV-1 that also show small fitness costs for most synonymous mutations, while half of the non-synonymous mutations are so deleterious that they are not even observed on short timescales \citep{zanini_vivo_2017}.
-The remaining SARS-CoV-2 ORFs show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
-This lack of strong purifying selection in these ORFs does not necessarily imply that they don't matter or are not a possible loci of adaptation -- but the exact amino acid sequence does not seem essential for their function.
+The SARS-CoV-2 \emph{ORF3a, ORF6, ORF7a, ORF7b} and \emph{ORF8} show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
+This lack of strong purifying selection in these ORFs does not necessarily imply that they don't matter or are not possible loci of adaptation -- but the exact amino acid sequence does not seem essential for their function.
 Only a few regions show a clear signal of conservation at 3rd positions, notably a central region of \emph{E} and the ribosome slippage region at the beginning of \emph{ORF1b}.
@@
 The heterogeneity in evolutionary rates and the combination of adaptive evolution, approximately neutral mutations, and purifying selection complicate the interpretation of phylodynamic analysis.
-Phylogenetics typically assumes that the mutation process is independent of the spread and epidemiology and that different sites evolve independently.
+Phylodynamics typically assumes that the mutation process is independent of the spread and epidemiology and that different sites evolve independently.
 These assumptions are (approximately) true for neutral mutations that occur along every lineage with the same rate.
@@
 Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic -- it is not the typical events, but the rare and extreme events that determine the course of adaptive evolution.
-Unlike neutral evolution, the rate of adaptive evolution depends on the population size, stochastic nature of the transmission process, the environment, and previous adaptation.
+Unlike neutral evolution, the rate of adaptive evolution depends on the population size, stochastic nature of the transmission process, the environment, and previous adaptation \citep{neher_genetic_2013}.
 These factors make extrapolation -- whether it is reconstruction of past events or scenarios of future evolution -- uncertain.
@@
 This process is likely to continue even in the more and more diverse immunological landscape of the human population in which the viral population spread and adapts.
-But it is unclear whether SARS-CoV-2 continues to evolve in a saltatory fashion with repeated emergence of highly mutated variants, or whether we will see a transition to a more gradual adaptive process.
-We currently observe multiple lineages with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), suggesting a shift to a more influenza-like evolution.
+But it is unclear whether SARS-CoV-2 continues to evolve in a saltatory fashion with repeated emergence of highly mutated variants, or whether I will see a transition to a more gradual adaptive process.
+I currently observe multiple lineages with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), which could indicated a shift to a more influenza-like evolution.
 
@@
 It is organized as a snakemake workflow \citep{koster_snakemakescalable_2012}.
-We used Nextclade data \texttt{2022-07-26T12:00:00Z} with reference sequence MN908947 (Wuhan-Hu-1).
+I used Nextclade data \texttt{2022-07-26T12:00:00Z} with reference sequence MN908947 (Wuhan-Hu-1).
 The codon analysis ignores \emph{ORF9b} which is fully contained in \emph{N} but in a different reading frame.
@@
 The fitness costs of each positions are estimated from the Nextclade columns \texttt{unlabeledSubstitutions} and \texttt{reversionSubstitutions} for each sequence that has QC-status \texttt{good}.
-For each Pango lineage, we count how often each specific mutation was observed.
-We then calculate how many lineages show minor variation for each mutation.
-To avoid loss of signal through sequencing errors, we drop lineages at which only a single mutation
-To turn this into a measure of mutational tolerance, we normalize the number of lineages with minor variation by the rate at which mutation rate of the ancestral base at this position.
+For each Pango lineage, I count how often each specific mutation was observed.
+I then calculate how many lineages show minor variation for each mutation.
+To avoid loss of signal through sequencing errors, I drop lineages at which only a single mutation
+To turn this into a measure of mutational tolerance, I normalize the number of lineages with minor variation by the rate at which mutation rate of the ancestral base at this position.
 

### 2022-08-22 scripts/combine_fits.py
fix typo

@@ -95,7 +95,7 @@ if __name__=="__main__":
         plt.show()
 
     plt.figure()
-    plt.plot(rates["aa_rate"], 'o-', label='Overall acid rate')
+    plt.plot(rates["aa_rate"], 'o-', label='Overall amino acid rate')
     plt.plot(rates["spike_rate"], 's-', label='spike protein')
     plt.plot(rates["orf1ab_rate"], 'd-', label='ORF1ab')
     plt.plot(rates["enm_rate"], 'd-', label='E,M,N')

### 2022-09-01 manuscript/variant_rates.tex
update manuscript

@@
 
+\begin{figure}
+    \includegraphics[width=0.5\textwidth]{figures/clade_overview.png}
+    \caption{{\bf Schematic summary of Nextstrain clades.}\label{fig:clade_overview}}
+\end{figure}
+
 \begin{figure*}[tb]
@@
     Bottom: Panels D and E show the prevalence of the most common genotypes (D) and mutations (E) during the first 3 months.
+Genotypes in (D) refer to sequences the {\bf all and only} the mutations indicated, while mutations in (E) count all sequences with a specific mutation, regardless of mutations elsewhere.
     In the case shown (20C), the founder genotype initially dominates and no specific mutation ever reaches high frequency.
@@
 To circumvent many of the above-mentioned problems and still use the majority of the available data, I use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
-I first use Nextclade \citep{aksamentov_nextclade_2021} to assign all sequences to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots).
+I first use Nextclade \citep{aksamentov_nextclade_2021} to assign 12 million sequences available in GISAID \citep{shu_gisaid_2017} to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots), see Fig.~\ref{fig:clade_overview}.
 Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
 These clades represent well-defined groups of sequences with little evidence of recombination within them and are analyzed separately.
+In this analysis, I only consider clades that have had significant circulation for at least 6 months.
+
 For each clade, I define a ``founder'' genotype and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
@@
 Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
-The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5$ days and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$ (note that this rate excludes strongly deleterious mutations, see Discussion).
+The probability that offspring genomes are different from their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5$ days and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$ (note that this rate excludes strongly deleterious mutations, see Discussion).
 
@@
 with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
+Note that this model assumes continuous time and only allows increments by one mutation at a time.
+Multiple mutations in one serial interval can still happen through successive mutations within one host.
 At time $t$ after the emergence of the variant, the number of mutations in the population is expected to be Poisson distributed with mean $ut$.
@@
     In panel D, clades are in alphabetical order, which is similar to their (uncertain) order of emergence.
-    The red outlier is `composite' clade 19++ with inflated rates due to adaptive mutations.
+    The red outlier is `composite' clade 19B++ with inflated rates due to adaptive mutations.
     \label{fig:rate_summary} }
@@
 I then calculate how many lineages show minor variation for each mutation.
-To avoid loss of signal through sequencing errors, I drop lineages at which only a single mutation
-To turn this into a measure of mutational tolerance, I normalize the number of lineages with minor variation by the rate at which mutation rate of the ancestral base at this position.
+To avoid loss of signal through sequencing errors, I drop lineages at which only a single mutation at this position is observed.
+To turn this count into a measure of mutational tolerance, I normalize the number of lineages with minor variation by the mutation rate away from the ancestral base at this position within the lineage.
 
@@
 I gratefully acknowledge the work by 1000s of scientists and healthcare professionals that collect the samples, generate, and share the sequence data on which this analysis is built.
-I am also grateful to Jesse Bloom and Louis du Plessis provided valuable feed-back on the manuscript and to Cornelius Roemer and Emma Hodcroft for stimulating discussions on many aspects of SARS-CoV-2 evolution.
+I am also grateful to Jesse Bloom, Louis du Plessis, and Florence D\'ebarre for valuable feed-back on the manuscript and to Cornelius Roemer and Emma Hodcroft for stimulating discussions on many aspects of SARS-CoV-2 evolution.
 

### 2022-09-01 scripts/combine_fits.py
add additional label to plot

@@ -86,7 +86,7 @@ if __name__=="__main__":
     plt.plot(rates["nuc_origin"], np.ones_like(rates['nuc_origin'])*inter_clade_rates["aa"], label='inter-clade amino acid rate', c=f"C{0}", lw=3)
     plt.plot(rates["nuc_origin"], np.ones_like(rates['nuc_origin'])*inter_clade_rates["syn"], label='inter-clade synonymous rate', c=f"C{1}", lw=3)
     plt.ylabel('rate estimate [1/y]')
-    plt.ylim(0,25)
+    plt.ylim(0)
     plt.legend()
 
     if args.output_plot_rates:

### 2022-09-01 scripts/plot_genotype_counts.py

@@ -88,13 +88,13 @@ if __name__=="__main__":
 
     ax = axs[1]
     ax.set_title("mutation frequencies", fontsize=1.2*fs)
-    ax.plot(dates, total, lw=3, c='k', alpha=0.3)
+    ax.plot(dates, total, lw=3, c='k', alpha=0.3, label='total')
     for m in sorted(counts['mutations'].keys(), key=lambda x:int(x[1:-1])):
         ax.plot(dates, counts['mutations'][m], '-o', label=f'{m}')
 
     ax = axs[0]
     ax.set_title("genotype frequencies", fontsize=1.2*fs)
-    ax.plot(dates, total, lw=3, c='k', alpha=0.3)
+    ax.plot(dates, total, lw=3, c='k', alpha=0.3, label='total')
     for m in sorted(counts['genotypes'].keys(), key=lambda x: len(x)):
         ax.plot(dates, counts['genotypes'][m], '-o', label=f'{m}' if m else "founder")
 

### 2022-10-24 manuscript/response_to_reviewers.tex
first round of revisions

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
\usepackage{hyperref}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\definecolor{pink}{rgb}{0.858, 0.188, 0.478}
\definecolor{applegreen}{rgb}{0.55, 0.71, 0.0}
\definecolor{armygreen}{rgb}{0.29, 0.33, 0.13}
\definecolor{brown}{rgb}{0.59, 0.29, 0.0}

\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}

% Add your own color comment command here!
% Uncomment & add your name
\newcommand{\Valentin}[1]{{\color{celestialblue}Robert: #1}}

\newcounter{counter1}[section]
\newcounter{counter2}[section]
\newcounter{counter3}[section]
\newcommand{\refa}[1]{\vskip 5mm \textbf{R1 - \stepcounter{counter1}\arabic{counter1}:} #1}
\newcommand{\refb}[1]{\vskip 5mm \textbf{R2 - \stepcounter{counter2}\arabic{counter2}:} #1}
\newcommand{\refc}[1]{\vskip 5mm \textbf{R3 - \stepcounter{counter3}\arabic{counter3}:} #1}
\newcommand{\editor}[1]{\textbf{Editor:} #1\vskip 5mm}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
\newcommand{\responsedraft}[1]{{\it {\color{pink}\textbf{Response:} #1}}\vskip 5mm}

\begin{document}
\section*{Response to reviewers}

\subsection*{Dear Drs. Stern and Duffy,}

\section*{Editor}

\editor{

Three reviewers have seen the paper, and I completely agree with all three that this is an excellent paper that uses a robust & rapid approach to obtain meaningful insights about SARS-CoV-2 evolution. The paper could be published in its current form but would benefit from very minor corrections suggested by the reviewers.
I also add a very small concern of my own: citing from the paper "...remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations". This is a little ad-hoc, how much would the results change if changing this? How many sequences are removed this way? To emphasize, I don't think any in-depth analysis is necessary here, just a few sentences on why this cutoff is used.
Looking forward to seeing this paper published,
Adi
}

\response{Thanks a lot for handling the paper, the constructive comments, and the rapid turn-around.

You are completely right that these cutoffs are heuristic and I have checked that the results are robust.
TODO.
}

\section*{Reviewer: 1}

\refa{
The author provides a timely analysis of broad interest demonstrating elevated inter vs intra-clade nonsynonymous mutation rates and furthermore that, in contrast to synonymous mutation rates, intra-clade rates decline over time for SARS-CoV-2. I believe the methods applied are reasonable, practical solutions to address problems associated with the very large data set of interest. I have no concerns with the methodology; I believe the findings are supported; I feel that any edits at this stage should be made at the author's discretion; and I believe the manuscript is suitable for publication in its current form. Nonetheless, I have a few comments I hope the author will consider addressing prior to publication.}

\response{Thank you very much.}

\refa{My major comment is related to the author's acknowledgement that, ``Nextstrain clades tend to be defined by long branches leading to a large polytomy. It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.'' As the author describes elsewhere, there are many potential reasons for the appearance of these long branches -- cryptic transfer, prolonged infection within a clinical population, reverse zoonosis. However, as the author states, none of these factors have been conclusively demonstrated for specific VOC emergence much less in general and it is possible that the mutations placed on these long branches emerged within hosts part of acute transmission chains belonging to robustly sequenced populations. Even if this is the case, while the sequencing efforts throughout the pandemic have been unprecedented, they are nonuniform and sequencing efforts (and reporting biases) are influenced by the emergence of mutations of interest. I believe it's likely the relative number of infections in the period immediately ancestral to the last common ancestor is underestimated relative to that of the period immediately following for VOC clades which may impact the estimation of relative mutation rates. This is exacerbated by changing public behavior (less likely to engage in social settings after VOC emergence). This effect may be more pronounced for nonsynonymous events than synonymous events due to the nature of the tree reconstruction.}

\response{This is an interesting thought. However, I am not 100\% sure how this differential ascertainment would affect the accumulation of mutations or inferences of their rate.}

\refa{The trends described here have parallels to those observed in influenza as the authors describe but also in Ebola https://www.sciencedirect.com/science/article/pii/S009286741500690X, https://www.sciencedirect.com/science/article/pii/S1473309919301185 which the author may consider referencing.}

\response{Real-time genomic epidemiology in outbreak settings was pioneered by the authors of these papers. Beyond a general notion that purifying selection supresses the non-synonymous mutations and that mutation spectra are very skewed, I don't see an immediate relevance of these papers. }

\refa{Regarding the relative strength of purifying selection across ORFs, I would selfishly bring the author's attention to my own work on the topic which arrived at similar conclusions: https://www.pnas.org/doi/full/10.1073/pnas.2104241118, see section beginning, ``Beyond this specific context, the presence of any hypervariable sites complicates the computation of the dN/dS ratio''}

\response{Thank you very much for bringing up this paper, which had escaped my attention. I have added a reference to this paper. }

\refa{Regarding the authors comment, ''Another possible explanation could be diminishing returns epistasis.'' I would again selfishly bring the author's attention to my own work demonstrating a constrained epistatic landscape within the RBD such that only a few key residues determine the shape of the interface which I believe supports the authors suggestion that each VOC has a limited mutational repertoire supporting the rapid host adaptation: https://journals.asm.org/doi/full/10.1128/mbio.00135-22}

\response{I have added a citation to this paper in the discussion of the potential role of epistasis. }

\refa{Finally: ``Genotypes in (D) refer to sequences the all and only the mutations indicated,'' I think this should be ... ``[with] all and only''}

\response{Thanks. Fixed!}

\section*{Reviewer: 2}

\refb{
This is a very nice study looking at the rate of evolution in SARS-CoV-2. Whereas other studies have looked at similar questions using increasingly complicated phylogenetic methods, Richard uses simple robust model that allows him to easily analyse huge numbers of sequences. Because of the wealth of SARS-CoV-2 data even a very simple model will illuminate the big patterns. This is an important reminder to all of us: that a model need only be complicated enough to answer the questions at hand. While hard numbers aren't given I assume that millions of genomes went into the analyses presented here. (The Tay and Hill papers cited here could only use hundreds of sequences). One small caveat of the approach is that it isn't suited to all scenarios, but is especially suited to clades with explosive exponential growth (this is also mentioned in the manuscript).

The main conclusions are:
\begin{enumerate}
 \item Synonymous mutations are mostly neutral
 \item The rate of adaptation clearly slowed over the first few months of the pandemic
 \item The process that gave rise to VOCs was decidedly different than normal human-human transmission
 \item The rate of adaptation and the slowdown are highly variable across the genome
\end{enumerate}
Perhaps most interesting to me is that the slowdown in the within-clade evolutionary rate is not seen in Spike, which is most important for infectivity and immune escape. However, Spike still appears to be under quite strong evolutionary constraint, as can be seen from it consistently having one of the slowest within-clade evolutionary rates (Fig. S1). None of these are particularly surprising conclusions, but this is the most elegant way I've seen of reaching them.

Overall I think the results are very interesting, the manuscript is well-written and the figures are good. I only have a few small comments.}

\response{Thank you very much for this accurate summary and generous remarks!}

\refb{1. Can this truly be said to be a tree-free approach? Presumably some phylogenies are calculated in order to assign sequences to Nextstrain clades? However, since this is continuously done and clade assignments are public it can be argued that this initial part of the analysis is "free".}

\response{The reviewer is correct. Sequences are assigned to clades by placing them on a reference tree.
Though this would also be possible using a constellation of mutations. If some clades were the result of recombination (as we soon might experience), the approach would work exactly the same way. So I think it is fair to say that most the results presented to make explicit use of phylogenies, though some implicit usage of phylogenetic concepts (clade and lineage assignment) is present. I chose to not discuss this point at length.  }

\refb{2. Does filtering out all sequences in a clade that don't have the full set of mutations relative to the reference mean you are excluding sequences with reversions to the reference? Do we know how often reversions occur? I don't think they are common, although they have been seen for Omicron (also mentioned here).}

\response{This is an important point (and I tripped over issues related to this when doing the analysis). Importantly, the requirement to have the full set of mutations refers to the clade defining mutations. Omicon clades (22A/BA.4, 22B/BA.5) have reversions, but as a result these mutations are not part of the clade defining mutations. Ignoring sequences that don't have the full set of clade defining mutation will ignore sequences that have additional reversions at clade defining positions. Until summer 2022, this was not much of a concern and most reversions within clades are clear QC issues. For the analysis of fitness costs and purifying selection, all sequences that pass the qualitaty thresholds are included and reversions are counted.
I have made this clearer in the manuscript.}

\refb{3. What is the ordering in Fig. S1? The order in which Nextstrain clades were assigned? I assume this is correlated to the order of clade emergence, but not necessarily exactly the same, nor would I expect the distance from the reference to the clade founder to have the same ordering. Would it make more sense to draw this figure in the same way as Fig. 4?}

\response{I have played with this, but I think the explicit ordering by clade is more informative (too many nearly overlapping markers otherwise). The clades are labeled in their order of designation and thus almost in their order of designation. I have added a statement to the figure caption. }

\refb{4. Could ORF1a and ORF1b be split in Fig. S1? It is mentioned that ORF1b is under the most evolutionary constraint (Fig. 6), but when combined with ORF1a as in Fig. S1 it appears to have the fastest rate of evolution.}

\response{I think the primary issue here is that there rates are per gene, and not per site. The objective here was to show how the total non-synonymous change is broken down by gene. I have now split `ORF1ab' TODO!! }

\refb{5. Should the size of the Nextstrain clades used here be given somewhere, e.g. in Fig. 1 or a supplementary table? It won't change anything, but would make more apparent just how many sequenes went into the analysis.}

\response{Thank you for this suggestion. I have included these numbers in Table 1. TODO}

\refb{6. Page 2, line 36: Instead of saying "separately" wouldn't it be more correct to say these clades are analyzed independently?}

\response{Yes. Thanks. Fixed!}

\refb{7. The last two "I"'s in the discussion should probably be "we"'s.}

\response{Ooops. Thanks. (find-replace problem)}

\section*{Reviewer 3}

\refc{This is a useful and well thought-out investigation of selection on SARS-CoV-2. I have only extremely minor comments on a few points to help clarify a few things.}

\response{Thank you for your constructive comments and generous remarks!}

\refc{In the discussion of what evidence exists already, a citation of Chaguza 2022 (https://www.medrxiv.org/content/10.1101/2022.06.29.22276868v1.full) would be useful to highlight that increased rate of evolution in a chronically infected patient has been documented. Other case studies of chronically infected patients have increased rates of evolution shown as well, but this is the only one to my knowledge that explicitly estimates it and compares it to the background rate of evolution.}

\response{Thank you for this suggestion. This is a very relevant reference that is now cited when discussion chronic infections and their putative role in variant emergence.}

\refc{I would be interested to have a definition (or a citation) of what constitutes a nextstrain clade -- I assume it is phylogenetic, but just a quick summation of what the rules are to define it would be helpful, especially since it is different to the more commonly used pango lineage system.}

\response{Yes, Nextstrain clades are phylogenetically defined and assigned using Nextclade. I have added a reference to a blog post on nextstrain.org that outlines the clade definitions.}

\refc{Paragraph starting ``In other cases, notable subclades…'' page 3 line 55 in the second column -- I don't follow the logic here of increased divergence ``generating an offset to diversity''. What is diversity if not within-clade divergence?}

\response{Yes, this was poorly worded. What I mean to say is that when there are many lineages, time since emergence can estimated accurately from the divergence of many independent evolutionary trajectories. When there is a single or a few large subclades, estimates of emergence times are more noisy. I have made an effort to reword this. }

\refc{What are the advantages of this method over using BEAST to estimate rates?}

\response{The primary advantages are (i) being able to use more data, and (ii) being more robust to problematic or misdated sequences. }

\refc{Figure 2 legend: ``Genotypes in D refer to sequences the all and only the mutations indicated'' -- are there some words missing here?}

\response{Yes. Thanks. Another reviewer picked this up as well and this is now fixed.}

\refc{I would find it helpful to have the evolutionary rates in substitutions/site/year as well as the rate per year in figure 3 (in the overall rate annotation)}

\response{Thank you for this suggestion, I have added this for panels A and B. The number of synonymous sites is tricky to define (2-fold, 4-fold, etc). }

\refc{Line 26 first column page 8 -- 1 x 10-3 would be clearer}

\response{$1\times 10^{-3}$ vs $10^{-3}$? I am not sure this helps.}

\refc{Hill et al 2022 is now published in Virus Evolution so this reference can be updated.}

\response{Updated!}

\end{document}

### 2022-10-24 manuscript/variant_rates.tex
first round of revisions

@@
 The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are common in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
+During such chronic infections, extensive intra-host diversity can develop through accelerated evolution
+\citep{chaguza_accelerated_2022}.
 Onward transmission from such chronic infections has also been documented \citep{gonzalez-reiche_intrahost_2022}.
@@
     Bottom: Panels D and E show the prevalence of the most common genotypes (D) and mutations (E) during the first 3 months.
-Genotypes in (D) refer to sequences the {\bf all and only} the mutations indicated, while mutations in (E) count all sequences with a specific mutation, regardless of mutations elsewhere.
+Genotypes in (D) refer to sequences with {\bf all and only} the mutations indicated, while mutations in (E) count all sequences with a specific mutation, regardless of mutations elsewhere.
     In the case shown (20C), the founder genotype initially dominates and no specific mutation ever reaches high frequency.
@@
 To circumvent many of the above-mentioned problems and still use the majority of the available data, I use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
-I first use Nextclade \citep{aksamentov_nextclade_2021} to assign 12 million sequences available in GISAID \citep{shu_gisaid_2017} to one of the Nextstrain clades \citep{hadfield_nextstrain_2018}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots), see Fig.~\ref{fig:clade_overview}.
+I first use Nextclade \citep{aksamentov_nextclade_2021} to assign 12 million sequences available in GISAID \citep{shu_gisaid_2017} to one of the Nextstrain clades \citep{hadfield_nextstrain_2018,roemer_sars-cov-2_2022}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots), see Fig.~\ref{fig:clade_overview}.
 Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
-These clades represent well-defined groups of sequences with little evidence of recombination within them and are analyzed separately.
+These clades represent well-defined groups of sequences with little evidence of recombination within them and are analyzed independently.
 In this analysis, I only consider clades that have had significant circulation for at least 6 months.
 
-For each clade, I define a ``founder'' genotype and exclude any sequence that does not have the full set of mutations relative to the reference sequence Wuhan/Hu-1.
+For each clade, I define a ``founder'' genotype and exclude any sequence that does not have the full set of clade defining mutations relative to the reference sequence Wuhan/Hu-1.
 This founder sequence is manually curated for each clade considered.
-This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence.
+This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence, but will ignore a few sequences with true reversion mutations.
 For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
@@
 In other cases, notable subclades did become dominant during the early stochastic dynamic.
-This can result in increased within-clade divergence, effectively generating an offset to diversity.
-Extrapolation of the linear fit to zero divergence is thus not necessarily a good estimator of the emergence time of a variant.
+In this case, extrapolation of the linear fit to zero divergence is not necessarily a good estimator of the emergence time of a variant: if branches leading to big subclades carry anomalously many or few mutations, the divergence time will be over- or underestimated.
 For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder-like variants are a minority even in early data.
@@
 Only \emph{E} shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35 (see Fig.~\ref{fig:fitness_landscape}).
+Systematic difference in purifying selection on the amino acid sequence of different reading frames have also been reported by \citet{rochman_ongoing_2021}.
 
@@
 Such diminishing returns epistasis has been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
-Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021}.
+Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021,rochman_epistasis_2022}.
 Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments by \citet{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate but had selective origin.
@@
 This process is likely to continue even in the more and more diverse immunological landscape of the human population in which the viral population spread and adapts.
-But it is unclear whether SARS-CoV-2 continues to evolve in a saltatory fashion with repeated emergence of highly mutated variants, or whether I will see a transition to a more gradual adaptive process.
-I currently observe multiple lineages with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), which could indicated a shift to a more influenza-like evolution.
+But it is unclear whether SARS-CoV-2 continues to evolve in a saltatory fashion with repeated emergence of highly mutated variants, or whether we will see a transition to a more gradual adaptive process.
+We currently observe multiple lineages with several amino acid changes and moderate transmission advantages emerging in a more stepwise fashion within 21L, 22A, 22B (BA.2, BA.4, BA.5), which could indicated a shift to a more influenza-like evolution.
 

### 2022-10-24 scripts/root_to_tip.py

@@ -158,6 +158,8 @@ if __name__=="__main__":
     parser.add_argument('--clade-gts', type=str, required=True, help="input data")
     parser.add_argument('--clade', type=str, required=True, help="input data")
     parser.add_argument('--sub-clades', type=str, required=True, help="input data")
+    parser.add_argument('--qc-cutoff-scale', type=float, default=2, help="number of std dev of rtt filter")
+    parser.add_argument('--qc-cutoff-offset', type=float, default=3, help="number extra mutation")
     parser.add_argument('--min-date', type=float, help="input data")
     parser.add_argument('--max-date', type=float, help="input data")
     parser.add_argument('--max-group', type=int, help="input data")
@@ -178,7 +180,7 @@ if __name__=="__main__":
  #   iqd = scoreatpercentile(filtered_data.residuals, 75) -  scoreatpercentile(filtered_data.residuals, 25)
  #   filtered_data["outlier"] = filtered_data.residuals.apply(lambda x: np.abs(x)>5*iqd)
 
-    tolerance = lambda t: 3 + 2*np.sqrt(np.maximum(0,(regression.intercept + regression.slope*t)))
+    tolerance = lambda t: args.qc_cutoff_offset + args.qc_cutoff_scale*np.sqrt(np.maximum(0,(regression.intercept + regression.slope*t)))
     filtered_data["outlier"] = filtered_data.apply(lambda x: np.abs(x.residuals)>tolerance(x.numdate), axis=1)
     ind = filtered_data.outlier==False
     # regression_clean = linregress(filtered_data.numdate[ind], filtered_data.divergence[ind])
@@ -238,7 +240,10 @@ if __name__=="__main__":
     rate_data = {'clade':args.clade, 'nuc':regression_clean, 'aa':regression_clean_aa,
                  'syn':regression_clean_syn, 'spike':regression_clean_spike,
                  'orf1':regression_clean_ORF1,'enm':regression_clean_ENM,
-                 "top_aaSubs": top_aaSubs,  "top_nucSubs": top_nucSubs}
+                 "top_aaSubs": top_aaSubs,  "top_nucSubs": top_nucSubs,
+                 "outliers_removed": np.sum(filtered_data.outlier),
+                 "qc_filter_fail": len(d) - len(filtered_data),
+                 "total_sequences": len(filtered_data)}
     if args.output_json:
         with open(args.output_json, 'w') as fh:
             json.dump(rate_data, fh)

### 2022-10-25 scripts/combine_fits.py
add per site rates to figure

@@ -22,6 +22,9 @@ if __name__=="__main__":
     parser.add_argument('--output-plot-rates-genes', type=str, help="plot file")
     args = parser.parse_args()
 
+    L = 29903
+    Laa = 9716  # all CDS other than ORF9)
+
     rates = pd.read_csv(args.rate_table, sep='\t', index_col='clade')
     inter_clade_rates = {}
     fig, axs = plt.subplots(2,2, figsize=(12,12))
@@ -54,7 +57,12 @@ if __name__=="__main__":
         ax.fill_between(x, y+std_dev, np.maximum(0, y-std_dev), fc='k', lw=3, alpha=0.1, ec=None)
         ax.set_xlim(x.min(), x.max())
         ax.set_ylim(0)
-        ax.text( 0.4, 0.06,f"overall rate: {reg.slope:1.1f}/year", fontsize=fs, transform=ax.transAxes)
+        if mut_type=='nuc':
+            ax.text( 0.5, 0.05,f"overall rate:\n{reg.slope:1.1f}/year\n{reg.slope/L:1.1e}/year/site",
+                    fontsize=fs, transform=ax.transAxes)
+        else:
+            ax.text( 0.5, 0.05,f"overall rate:\n{reg.slope:1.1f}/year\n{reg.slope/Laa:1.1e}/year/codon",
+                   fontsize=fs, transform=ax.transAxes)
         if mut_type == 'aa': ax.legend(ncol=2)
         ax.set_xticks(xticks, [str(x) for x in xticks], fontsize=fs)
 

### 2022-10-25 scripts/clone_growth.py
log dropped sequences

@@ -28,7 +28,7 @@ if __name__=="__main__":
     clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
-    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.min_date + 0.3, completeness=0, swap_root=args.clade_gts=='19B+')
+    filtered_data, _ = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.min_date + 0.3, completeness=0, swap_root=args.clade_gts=='19B+')
     #filtered_data=filtered_data.loc[filtered_data.country!='China']
 
     intra_subs_dis = filtered_data["divergence"].value_counts().sort_index()

### 2022-10-25 scripts/count_aa_mutations.py

@@ -1,0 +1,247 @@
+import argparse,json
+from unittest.mock import NonCallableMagicMock
+import pandas as pd
+import numpy as np
+import matplotlib as mpl
+from Bio import SeqIO,Seq
+mpl.rcParams['axes.formatter.useoffset'] = False
+from collections import defaultdict
+from scipy.stats import scoreatpercentile
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+columns = ['seqName', 'Nextclade_pango', "privateNucMutations.unlabeledSubstitutions", 'qc.overallStatus', 'privateNucMutations.reversionSubstitutions']
+
+def get_sequence(mods, root_seq):
+    seq = root_seq.copy()
+    for mut in mods['nuc']:
+        a,pos,d = mut[0], int(mut[1:-1])-1, mut[-1]
+        if a!=seq[pos]:
+            print(seq[pos], mut)
+        seq[pos]=d
+    return seq
+
+def translate(codon):
+    try:
+        return Seq.translate("".join(codon))
+    except:
+        return 'X'
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+    parser.add_argument('--metadata', type=str, required=True, help="input data")
+    parser.add_argument('--reference', type=str, required=True, help="input data")
+    parser.add_argument('--pango-gts', type=str, required=True, help="input data")
+    parser.add_argument('--rare-cutoff', type=int, default=2, help="minimal number of occurrences to count mutations")
+    parser.add_argument('--output-fitness', type=str, required=True, help="fitness table")
+    parser.add_argument('--output-events', type=str, required=True, help="events table")
+    parser.add_argument('--output-mutations', type=str, required=True, help="fitness table")
+    args = parser.parse_args()
+
+    ## load reference
+    ref = SeqIO.read(args.reference, 'genbank')
+    ref_array = np.array(ref.seq)
+    base_content = {x:ref.seq.count(x) for x in 'ACGT'}
+
+    ## make a map of codon positions, ignore orf9b (overlaps N)
+    codon_pos = np.zeros(len(ref))
+    map_to_gene = {}
+    gene_position = {}
+    gene_length = {}
+    for feat in ref.features:
+        if feat.type=='CDS':
+            gene_name = feat.qualifiers['gene'][0]
+            gene_position[gene_name] = feat.location
[…]

### 2022-10-25 scripts/explore.py

@@ -6,6 +6,6 @@ from datetime import datetime
 clade_gt = get_clade_gts("data/clade_gts.json", "20I")
 
 alpha = pd.read_csv('subsets/20I.tsv', sep='\t')
-alpha_filtered = filter_and_transform(alpha, clade_gt, min_date=numeric_date(datetime(2020,8,1)))
+alpha_filtered, _ = filter_and_transform(alpha, clade_gt, min_date=numeric_date(datetime(2020,8,1)))
 
 print(alpha_filtered.loc[alpha_filtered.date<"2020-10-01",["strain", "date", "intra_substitutions"]])

### 2022-10-25 scripts/get_genotype_counts.py

@@ -26,7 +26,7 @@ if __name__=="__main__":
     clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.concat([pd.read_csv(x, sep='\t').fillna('') for x in args.metadata])
-    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
+    filtered_data, _ = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
                                          query = args.query, max_group=args.max_group, QC_threshold=80 if args.clade=='21H' else 30,
                                          completeness=0, swap_root=args.clade.startswith('19B+'))
     filtered_data["day"] = filtered_data.datetime.apply(lambda x:x.toordinal())

### 2022-10-25 scripts/mutations_by_clade.py

@@ -1,0 +1,123 @@
+import argparse,json
+from importlib.metadata import metadata
+import pandas as pd
+import numpy as np
+import matplotlib as mpl
+from Bio import SeqIO,Seq
+mpl.rcParams['axes.formatter.useoffset'] = False
+from collections import defaultdict
+from scipy.stats import scoreatpercentile
+import seaborn as sns
+import pandas as pd
+
+import matplotlib.pyplot as plt
+import seaborn as sns
+
+columns = ['seqName', 'clade', "privateNucMutations.unlabeledSubstitutions", 'qc.overallStatus', 'privateNucMutations.reversionSubstitutions']
+
+def get_sequence(mods, root_seq):
+    seq = root_seq.copy()
+    for mut in mods['nuc']:
+        a,pos,d = mut[0], int(mut[1:-1])-1, mut[-1]
+        if a!=seq[pos]:
+            print(seq[pos], mut)
+        seq[pos]=d
+    return seq
+
+def translate(codon):
+    try:
+        return Seq.translate("".join(codon))
+    except:
+        return 'X'
+
+if __name__=="__main__":
+    parser = argparse.ArgumentParser(
+        description="remove time info",
+        formatter_class=argparse.ArgumentDefaultsHelpFormatter
+    )
+    parser.add_argument('--metadata', type=str, required=True, help="input data")
+    parser.add_argument('--reference', type=str, required=True, help="input data")
+    args = parser.parse_args()
+
+    ## load reference
+    ref = SeqIO.read(args.reference, 'genbank')
+    ref_array = np.array(ref.seq)
+    base_content = {x:ref.seq.count(x) for x in 'ACGT'}
+
+    ## make a map of codon positions, ignore orf9b (overlaps N)
+    codon_pos = np.zeros(len(ref))
+    map_to_gene = {}
+    gene_position = {}
+    gene_length = {}
+    for feat in ref.features:
+        if feat.type=='CDS':
+            gene_name = feat.qualifiers['gene'][0]
+            gene_position[gene_name] = feat.location
+            gene_length[gene_name] = (feat.location.end - feat.location.start)//3
+            if gene_name!='ORF9b':
+                for gpos, pos in enumerate(feat.location):
+                    codon_pos[pos] = (gpos%3)+1
+                    map_to_gene[pos]= gene_name
+
+    # load and filter metadata
[…]

### 2022-10-25 scripts/root_to_tip.py

@@ -24,17 +24,28 @@ def week_since2020_to_numdate(d):
 def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False, max_group=None, QC_threshold=30):
     # filter for incomplete data
     d = d.loc[d.date.apply(lambda x:len(x)==10 and 'X' not in x)]
+    pre = len(d)
     d = d.loc[d.QC_overall_score<QC_threshold]
+    dropped_seqs['QC'] = pre - len(d)
+
     if query:
+        pre = len(d)
         d = d.query(query)
+        dropped_seqs['query'] = pre - len(d)
+
     d['datetime'] = d.date.apply(lambda x: datetime.strptime(x, '%Y-%m-%d'))
     d['numdate'] = d.datetime.apply(lambda x: numeric_date(x))
     d['CW'] = d.datetime.apply(date_to_week_since2020)
     # filter date range
+    dropped_seqs = {}
     if min_date:
+        pre = len(d)
         d = d.loc[d.numdate>min_date]
+        dropped_seqs['min_date'] = pre - len(d)
     if max_date:
+        pre = len(d)
         d = d.loc[d.numdate<max_date]
+        dropped_seqs['max_date'] = pre - len(d)
 
     # look for clade defining substitutions
     d["clade_substitutions"] = d.substitutions.apply(lambda x:     [y for y in x.split(',') if y in clade_gt['nuc']] if x else [])
@@ -73,12 +84,14 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
 
     # filter
     if completeness is not None:
+        pre = len(d)
         d = d.loc[d.missing_subs<=completeness]
+        dropped_seqs['completeness'] = pre - len(d)
 
     if max_group:
         return d.groupby(['CW', 'country']).sample(max_group, replace=True).drop_duplicates(subset='strain')
 
-    return d
+    return d, dropped_seqs
 
 def weighted_regression(x,y,w):
     '''
@@ -171,7 +184,7 @@ if __name__=="__main__":
     clade_gt = get_clade_gts(args.clade_gts, args.sub_clades)
 
     d = pd.read_csv(args.metadata, sep='\t').fillna('')
-    filtered_data = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
+    filtered_data, dropped_seqs = filter_and_transform(d, clade_gt, min_date=args.min_date, max_date=args.max_date,
[…]

### 2022-10-25 scripts/root_to_tip.py
add qc log output

@@ -255,7 +255,8 @@ if __name__=="__main__":
                  'orf1':regression_clean_ORF1,'enm':regression_clean_ENM,
                  "top_aaSubs": top_aaSubs,  "top_nucSubs": top_nucSubs,
                  "outliers_removed": np.sum(filtered_data.outlier),
-                 "qc_filter_fail": len(d) - len(filtered_data),
+                 "qc_filter_fail": dropped_seqs["QC"],
+                 "incomplete": dropped_seqs["completeness"],
                  "total_sequences": len(filtered_data)}
     if args.output_json:
         with open(args.output_json, 'w') as fh:

### 2022-10-25 scripts/root_to_tip.py
add qc log output

@@ -254,7 +254,7 @@ if __name__=="__main__":
                  'syn':regression_clean_syn, 'spike':regression_clean_spike,
                  'orf1':regression_clean_ORF1,'enm':regression_clean_ENM,
                  "top_aaSubs": top_aaSubs,  "top_nucSubs": top_nucSubs,
-                 "outliers_removed": np.sum(filtered_data.outlier),
+                 "outliers_removed": int(np.sum(filtered_data.outlier)),
                  "qc_filter_fail": dropped_seqs["QC"],
                  "incomplete": dropped_seqs["completeness"],
                  "total_sequences": len(filtered_data)}

### 2022-10-30 manuscript/response_to_reviewers.tex
work on revision

@@
 You are completely right that these cutoffs are heuristic and I have checked that the results are robust.
-TODO.
+
+The number of outliers removed varies from clade to clade as the clades, but never exceeds 1\% with the standard parameters.
+Changing these parameters (using 4 instead of 2 sigma of the Poisson distribution, 6 instead or 3 mutations offset which reduces outlier removal by 90\%) changes the rate estimates by a less than one percent (3\% at most for clade 20H).
+The relative insensitivity to these outliers is due to the fact that we don't build a tree and a few outliers don't distort the overall picture.
+Similarly, making the QC filter less stringent (using 80 instead of 30) doesn't change rate estimate by more than 2\% on average. The biggest effect of the QC filter is filtering for completeness.
+I have added a few sentences to highlight this.
+
 }
@@
 
-\response{This is an interesting thought. However, I am not 100\% sure how this differential ascertainment would affect the accumulation of mutations or inferences of their rate.}
+\response{This is an interesting thought. However, I am not 100\% sure I understand how this differential ascertainment would affect the accumulation of mutations or inferences of their rate.}
 
@@
 
-\response{Thank you very much for this accurate summary and generous remarks!}
+\response{Thank you very much for this accurate summary and the generous remarks!}
 
@@
 \response{The reviewer is correct. Sequences are assigned to clades by placing them on a reference tree.
-Though this would also be possible using a constellation of mutations.
+Though this would also be possible using a constellation of mutations or some other sort of clustering.
 If some clades were the result of recombination (as we soon might experience), the approach would work exactly the same way.
-So I think it is fair to say that most the results presented to make explicit use of phylogenies, though some implicit usage of phylogenetic concepts (clade and lineage assignment) is present.
+So I think it is fair to say that most the results presented do not make explicit use of phylogenies, though some implicit usage of phylogenetic concepts (clade and lineage assignment) is present.
 I chose to not discuss this point at length.
@@
 
-\response{This is an important point (and I tripped over issues related to this when doing the analysis).
-Importantly, the requirement to have the full set of mutations refers to the clade defining mutations.
-Omicon clades (22A/BA.4, 22B/BA.5) have reversions, but as a result these mutations are not part of the clade defining mutations.
+\response{This is an important point (and I tripped over issues related to this when doing the analysis of fitness costs).
+Importantly, the requirement for a sequence to have the complete set of mutations refers to the clade defining mutations.
+Omicon clades (22A/BA.4, 22B/BA.5) have reversions and the reverted positions are not part of the clade defining mutations (their genotype is ancestral).
 Ignoring sequences that don't have the full set of clade defining mutation will ignore sequences that have additional reversions at clade defining positions.
@@
 What I mean to say is that when there are many lineages, time since emergence can estimated accurately from the divergence of many independent evolutionary trajectories.
-When there is a single or a few large subclades, estimates of emergence times are more noisy.
+When there is a single subclade or a few large subclades, estimates of emergence times are more noisy.
 I have made an effort to reword this.
@@
 
-\response{Thank you for this suggestion, I have added this for panels A and B. The number of synonymous sites is tricky to define (2-fold, 4-fold, etc). }
+\response{Thank you for this suggestion, I have added this for panels A and B.
+If give amino acid and synonynmous rates 'per codon' to avoid subtleties counting synonymous sites (2-fold, 4-fold, etc).
+}
 
@@
 
-\response{$1\times 10^{-3}$ vs $10^{-3}$? I am not sure this helps.}
+\response{$1\times 10^{-3}$ vs $10^{-3}$? I am not sure adding ``$1\times$" makes this clearer.}
 

### 2022-10-30 manuscript/variant_rates.tex
work on revision

@@
 This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence, but will ignore a few sequences with true reversion mutations.
+Results are insensitive to the stringency of this filtering.
 For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
@@
 To exclude these outliers, I perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations.
+These outliers are never more than 0.5\% of all sequences that pass the first round of filtering and the results are insensitive to the filtering criteria.
 
@@
 Only \emph{E} shows slightly less variation at 3rd positions than other genes with a notable dip in the middle of the gene around codon 35 (see Fig.~\ref{fig:fitness_landscape}).
-Systematic difference in purifying selection on the amino acid sequence of different reading frames have also been reported by \citet{rochman_ongoing_2021}.
+Systematic difference in purifying selection on the amino acid sequence of different open reading frames have also been reported by \citet{rochman_ongoing_2021}.
 
@@
 Such diminishing returns epistasis has been observed in experimental evolution with yeast \citep{kryazhimskiy_global_2014}.
-Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021,rochman_epistasis_2022}.
+Global epistasis of this nature can exist along side epistasis of specific amino acid changes as for example suggested for the \texttt{S:N501Y} mutation \citep{martin_emergence_2021,rochman_epistasis_2022,martin_selection_2022}.
 Differences in diversification of SARS-CoV-2 with and without the S:D614G mutation where also observed in evolution experiments by \citet{amicone_mutation_2022}, who concluded that these differences were not due to a change in base line mutation rate but had selective origin.

### 2022-10-30 scripts/root_to_tip.py
split orf1ab in rate estimation

@@ -58,7 +58,8 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
     # define "with-in clade substitutions"
     d["intra_aaSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y] if x else [])
     d["intra_SpikeSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0]=='S'] if x else [])
-    d["intra_ORF1Substitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:4]=='ORF1'] if x else [])
+    d["intra_ORF1bSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:5]=='ORF1a'] if x else [])
+    d["intra_ORF1aSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:5]=='ORF1b'] if x else [])
     d["intra_ENMSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0] in ['E','N','M']] if x else [])
 
     if swap_root:
@@ -77,7 +78,8 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
     d["divergence"] =   d.intra_substitutions.apply(lambda x:   len(x))
     d["aaDivergence"] = d.intra_aaSubstitutions.apply(lambda x: len(x))
     d["spikeDivergence"] = d.intra_SpikeSubstitutions.apply(lambda x: len(x))
-    d["orf1Divergence"] = d.intra_ORF1Substitutions.apply(lambda x: len(x))
+    d["orf1aDivergence"] = d.intra_ORF1aSubstitutions.apply(lambda x: len(x))
+    d["orf1bDivergence"] = d.intra_ORF1bSubstitutions.apply(lambda x: len(x))
     d["enmDivergence"] = d.intra_ENMSubstitutions.apply(lambda x: len(x))
     d["synDivergence"] = d["divergence"] - d["aaDivergence"]
 
@@ -202,7 +204,8 @@ if __name__=="__main__":
[…]

### 2022-10-30 scripts/combine_fits.py
split orf1ab in plots

@@ -105,9 +105,10 @@ if __name__=="__main__":
     plt.figure()
     plt.plot(rates["aa_rate"], 'o-', label='Overall amino acid rate')
     plt.plot(rates["spike_rate"], 's-', label='spike protein')
-    plt.plot(rates["orf1ab_rate"], 'd-', label='ORF1ab')
+    plt.plot(rates["orf1a_rate"], 'd-', label='ORF1a')
+    plt.plot(rates["orf1b_rate"], 'd-', label='ORF1b')
     plt.plot(rates["enm_rate"], 'd-', label='E,M,N')
-    plt.plot(rates["aa_rate"] - rates["spike_rate"] - rates["orf1ab_rate"]- rates["enm_rate"],
+    plt.plot(rates["aa_rate"] - rates["spike_rate"] - rates["orf1a_rate"] - rates["orf1b_rate"]- rates["enm_rate"],
              'v-', label='other ORFs')
     plt.ylabel('rate estimate [subs/y]')
     plt.legend()

### 2022-10-30 scripts/root_to_tip.py
fix orf1ab mix up

@@ -58,8 +58,8 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
     # define "with-in clade substitutions"
     d["intra_aaSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and 'ORF9' not in y] if x else [])
     d["intra_SpikeSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0]=='S'] if x else [])
-    d["intra_ORF1bSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:5]=='ORF1a'] if x else [])
-    d["intra_ORF1aSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:5]=='ORF1b'] if x else [])
+    d["intra_ORF1aSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:5]=='ORF1a'] if x else [])
+    d["intra_ORF1bSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[:5]=='ORF1b'] if x else [])
     d["intra_ENMSubstitutions"] = d.aaSubstitutions.apply(lambda x: [y for y in x.split(',') if y not in clade_gt['aa'] and y[0] in ['E','N','M']] if x else [])
 
     if swap_root:

### 2022-10-30 manuscript/response_to_reviewers.tex
final edits

@@
 
-Three reviewers have seen the paper, and I completely agree with all three that this is an excellent paper that uses a robust & rapid approach to obtain meaningful insights about SARS-CoV-2 evolution.
+Three reviewers have seen the paper, and I completely agree with all three that this is an excellent paper that uses a robust \& rapid approach to obtain meaningful insights about SARS-CoV-2 evolution.
 The paper could be published in its current form but would benefit from very minor corrections suggested by the reviewers.
@@
 I have added a few sentences to highlight this.
+
+Please find my response to the reviewer comments below.
+
+best,
+Richard
 
@@
 The objective here was to show how the total non-synonymous change is broken down by gene.
-I have now split `ORF1ab' TODO!!
+I have now split `ORF1ab'.
+The updated plot shows that most of the observed evolution is indeed in ORF1a.
 }
@@
 
-\response{Thank you for this suggestion. I have included these numbers in Table 1. TODO}
+\response{Thank you for this suggestion. I have included these numbers in Table 1.}
 
@@
 What I mean to say is that when there are many lineages, time since emergence can estimated accurately from the divergence of many independent evolutionary trajectories.
-When there is a single subclade or a few large subclades, estimates of emergence times are more noisy.
+When there is a single subclade or a few large subclades, estimates of emergence times are more noisy due to shared ancestry.
 I have made an effort to reword this.
@@
 
-\response{The primary advantages are (i) being able to use more data, and (ii) being more robust to problematic or misdated sequences. }
+\response{The primary advantages are (i) being able to use more data, (ii) being more robust to problematic or misdated sequences, and (iii) fewer cryptic dependencies on model and prior choices.
+Disadvantages compared to phylogenetic methods is not taking into account shared ancestry between sequences.
+But since this method is only applied to variation within clades that have rapidly expanded, this is a minor issue.
+}
 

### 2022-10-30 manuscript/variant_rates.tex
final edits

@@
 The rate of synonymous mutations is found to be around 6 changes per year.
-Synonymous rates within variants vary little from variant to variant and are compatible with the overall rate.
+Synonymous rates within variants vary little from variant to variant and are compatible with the overall rate of 7 changes per year (or $7.5\times 10^{-4}$ per year and codon).
 In contrast, the rate at which variants accumulate amino acid changes (non-synonymous mutation) was initially around 12-16 changes per year, but in 2021 and 2022 dropped to 6-9 changes per year.
-The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 25 amino acid changes per year.
-This 2-fold higher rate indicates that the evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains and likely dominated by adaptive evolution.
-I further quantify the spectrum of mutations and purifying selection in different SARS-CoV-2 proteins.
+The overall rate of non-synonymous evolution, that is across variants, is estimated to be about 26 amino acid changes per year (or $2.7\times 10^{-3}$ per year and codon).
+This strong acceleration of the overall rate compared to within clade evolution indicates that the evolutionary process that gave rise to the different variants is qualitatively different from that in typical transmission chains and likely dominated by adaptive evolution.
+I further quantify the spectrum of mutations and purifying selection in different SARS-CoV-2 proteins and show that the massive global sampling of SARS-CoV-2 is sufficient to estimate site specific fitness costs across the entire genome.
 Many accessory proteins evolve under limited evolutionary constraint with little short term purifying selection.
-About half of the mutations in other proteins are strongly deleterious and rarely observed, not even at low frequency.
+About half of the mutations in other proteins are strongly deleterious.
 \end{abstract}
@@
 Since its emergence in late 2019 \citep{zhu_novel_2020}, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
-Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
+Many of these variants spread considerably faster and quickly displaced the resident variants at the time either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
 Specific Variants of Concern (VoC) or Interest (VoI) were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
@@
 
-The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within clades or variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who showed that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
+The dichotomous pattern of SARS-CoV-2 evolution with step-wise evolution within variants and atypical bursts of evolution leading to new variants has been investigated by \citet{tay_emergence_2022} and \citet{hill_origins_2022}, who showed that the rate of evolution along branches giving rise to new variants is up to four-fold higher than the background rate.
 Here, I build on these results and investigate the patterns of SARS-CoV-2 diversification within variants and compare these to the global dynamics of evolution and adaptation.
@@
 Furthermore, early variants display more rapid non-synonymous evolution than later variants suggesting more ubiquitous adaptive evolution early on.
-I further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome.
+I further quantify the level of functional constraint of different open reading frames and infer a map of mutational tolerance across the genome from patterns of rare diversity.
 
@@
 These methods, however, cannot handle the volume of SARS-CoV-2 data available and data have to be dramatically down-sampled.
-Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated, recombinant, contaminated, or otherwise chimeric sequence can fatally distort the analysis, in particular if placed early in tree.
+Furthermore, phylogenetic methods impose an hierarchical structure on the data and are thus very sensitive to problematic sequences or metadata: Any mis-dated, recombinant, contaminated, or otherwise chimeric sequence can fatally distort the analysis.
 Problematic sequences can be particularly common when a new variant takes over since sequencing protocols need adjusting \citep{deMaio_issues_2020}.
@@
 To circumvent many of the above-mentioned problems and still use the majority of the available data, I use a combination of automated filtering and simple robust approaches to analyze the evolutionary patterns (see Materials and Methods).
-I first use Nextclade \citep{aksamentov_nextclade_2021} to assign 12 million sequences available in GISAID \citep{shu_gisaid_2017} to one of the Nextstrain clades \citep{hadfield_nextstrain_2018,roemer_sars-cov-2_2022}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots), see Fig.~\ref{fig:clade_overview}.
+I first use Nextclade \citep{aksamentov_nextclade_2021} to assign 12 million sequences available in GISAID (2022-07-25) \citep{shu_gisaid_2017} to one of the Nextstrain clades \citep{hadfield_nextstrain_2018,roemer_sars-cov-2_2022}, which are denoted by a Year-Letter combination (e.g.~19A, 20A, 20B, \ldots), see Fig.~\ref{fig:clade_overview}.
 Sequences belonging to recombinant Pango lineages are excluded \citep{rambaut_dynamic_2020}.
@@
 This filtering removes most incomplete sequences as well as sequences where amplicon drop-out are back-filled with the reference sequence, but will ignore a few sequences with true reversion mutations.
+In addition, sequences with a Nextclade QC score above 30 (80 for 21H because of an unaccounted frameshift) are removed.
 Results are insensitive to the stringency of this filtering.
-For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype to analyze diversification and divergence within the clade.
+For this reduced set of sequences, I determine the mutations they carry on top of the founder genotype of the clade to analyze diversification and divergence within the clade.
 The latter step is done for nucleotide changes as well as for amino acid changes.
@@
 For the majority of sequences, this is approximately true, but some problematic sequences have more mutations than expected.
-To exclude these outliers, I perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations.
-These outliers are never more than 0.5\% of all sequences that pass the first round of filtering and the results are insensitive to the filtering criteria.
+To exclude these outliers, I perform a simple linear regression of the number of ``intra-clade'' mutations against time and remove sequences whose deviation from the linear fit exceeds twice the expected standard deviation by 3 mutations (see Fig.~\ref{fig:within_clade}).
+These outliers are never more than 1\% of all sequences that pass the first round of filtering and the results are insensitive to the filtering criteria.
 
@@
 Evolutionary rate and putative emergence date of the variant are then estimated by weighted linear regression where each time bin is weighted with the fourth root of the number of sequences in the bins.
-The exact functional form of this weighting does not have a big influence on the results, but a sublinear weighting helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early after a new variant emerges.
-Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, corresponding to a large clade that emerged in early 2020 that was common in North America and Scandinavia.
+The exact functional form of this weighting does not have a big influence on the results, but a sublinear weighting helps to counter the large variation in sequencing effort across countries and the unavoidable imbalance due to the fact that few sequences are available early on when an emerging variant is still rare.
+Fig.~\ref{fig:within_clade} shows the increasing intra-clade divergence for clade 20C, a large clade that emerged in early 2020 that was common in North America and Scandinavia.
 Both synonymous and non-synonymous within-clade average divergence increase linearly over time allowing for a robust estimate of the rate.
 
-Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is not suitable to estimate evolutionary rates in general.
+Due to shared ancestry, divergences of sequences are not independent data points and a regression against time is general not a suitable method to estimate evolutionary rates.
 In particular, confidence intervals are difficult to obtain.
@@
 A simple model for diversity within a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
-Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
+Offspring of genomes with $i$ mutations will carry $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
 The probability that offspring genomes are different from their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5$ days and a rate of $\mu = 15/year$ evaluates to $u\approx 0.2$ (note that this rate excludes strongly deleterious mutations, see Discussion).
@@
 In this case, extrapolation of the linear fit to zero divergence is not necessarily a good estimator of the emergence time of a variant: if branches leading to big subclades carry anomalously many or few mutations, the divergence time will be over- or underestimated.
-For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21 (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder-like variants are a minority even in early data.
+For several clades 19B, 20H (Beta), 21D (Eta), 21G (Lambda), 21H (Mu), 21I (Delta), 21J (Delta), 21L (Omicron, BA.2), and 22B (Omicron, BA.5), founder-like variants are a minority even in early data.
 
@@
     The standard deviation expected based on Poisson statistics is indicated as shaded area.
-    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate ofinter-clade rates (gray lines).
+    Panel D summarizes the individual rate estimates (dots and squares) and compares them to the estimate of inter-clade rates (gray lines).
     In panel D, clades are in alphabetical order, which is similar to their (uncertain) order of emergence.
-    The red outlier is `composite' clade 19B++ with inflated rates due to adaptive mutations.
+    The red outlier is `composite' clade 19B++ (containing 19B, 19A, 20A, 20B, 20C) with inflated rates due to adaptive mutations on the branch leading to clade 20A.
     \label{fig:rate_summary} }
@@
 These rates are summarized in Fig.~\ref{fig:rate_summary} and Tab.~\ref{tab:rates}.
-In addition, I estimated the rate at which the variants themselves accumulated amino acid and synonymous changes by regressing the number of changes of the variants (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the variants.
+In addition to these within-clade rates, I estimated the rate at which the clades themselves accumulated amino acid and synonymous changes by regressing the number of differences of the clade's founder sequence (relative to the putative root in clade 19B \citep{caraballo-ortiz_tophap_2022}) against the estimated time of origin of the clade.
 These regressions are shown in thick gray lines in Fig.~\ref{fig:rate_summary}.
 
-Rates of synonymous change are very consistent across variants (about 5-8 changes per genome per year) and also agree with the overall rate of synonymous changes of 7.3 changes per genome per year.
+Rates of synonymous change are very consistent across clades (about 5-8 changes per genome per year) and also agree with the overall rate of synonymous changes of 7.3 changes per genome per year.
 The rates of non-synonymous changes are much more variable (Fig.~\ref{fig:rate_summary}D).
@@
 It could thus be that the estimated inter-clade rate exceeds the intra-clade rate purely because of this conditioning.
-This effect might be particularly important early on in the pandemic where branches with as few as two mutations were used to define new clades.
+This effect might be particularly important early on in the pandemic when diversity was low and when branches with as few as two mutations were used to define new clades.
 I therefore also included composite clades 19B+, 19B++ and 20A+ containing sequences from 19A and 19B (rooted on 19B), 19A, 19B, 20A, 20B, 20C, 20D (rooted on 19B), and 20A, 20B, 20C, 20D (rooted on 20A).
@@
 \begin{table*}
-\begin{tabular}{l|rrrrrr}
+\begin{tabular}{l|rrrrrrr}
     \hline
-    {\bf clade} &  overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
+    {\bf clade} & \# of seq. & overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
     \hline
-    19B &     19.46 &    12.37 &      7.09 &        0 &       0 &        0 \\
-   19B+ &     18.56 &    11.92 &      6.64 &        0 &       0 &        0 \\
-  19B++ &     27.91 &    16.41 &     11.50 &        0 &       0 &        0 \\
-  19A &     19.73 &    12.79 &      6.95 &        2 &       1 &        1 \\
-  20A &     15.08 &     7.94 &      7.15 &        6 &       3 &        3 \\
-    20B &     14.29 &     7.69 &      6.59 &        9 &       5 &        4 \\
-    20C &     16.73 &     9.92 &      6.81 &        8 &       5 &        3 \\
-   20A+ &     16.30 &     9.40 &      6.90 &        6 &       3 &        3 \\
-    20E &     14.04 &     5.65 &      8.39 &       13 &       5 &        8 \\
-    20H (Beta) &     15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
-    20I (Alpha) &     11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
-    20J (Gamma) &     12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
-    21D (Eta) &     11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
-    21G (Lambda) &     12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
-    21H (Mu) &     18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
-    21I (Delta) &     16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
-    21J (Delta) &     14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
-    21K (Omicron) &     13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
-    21L (Omicron) &      9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
-    22A (Omicron) &      7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
-    22B (Omicron) &     11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
+    19B &           8187 & 19.46 &    12.37 &      7.09 &        0 &       0 &        0 \\
+   19B+ &           22745 & 18.56 &    11.92 &      6.64 &        0 &       0 &        0 \\
+  19B++ &           126823 & 27.91 &    16.41 &     11.50 &        0 &       0 &        0 \\
+  19A &             14382 & 19.73 &    12.79 &      6.95 &        2 &       1 &        1 \\
+  20A &             44629 & 15.08 &     7.94 &      7.15 &        6 &       3 &        3 \\
+    20B &           42077 & 14.29 &     7.69 &      6.59 &        9 &       5 &        4 \\
+    20C &           47865 & 16.73 &     9.92 &      6.81 &        8 &       5 &        3 \\
+   20A+ &           136863 & 16.30 &     9.40 &      6.90 &        6 &       3 &        3 \\
+    20E &           82240 & 14.04 &     5.65 &      8.39 &       13 &       5 &        8 \\
+    20H (Beta) &    5077 & 15.48 &     9.51 &      5.96 &       21 &      16 &        5 \\
+    20I (Alpha) &   277226 & 11.56 &     5.49 &      6.07 &       30 &      19 &       11 \\
+    20J (Gamma) &   24970 & 12.97 &     6.29 &      6.68 &       35 &      22 &       13 \\
+    21D (Eta) &     3653 & 11.36 &     6.29 &      5.07 &       27 &      13 &       14 \\
+    21G (Lambda) &  4567 & 12.04 &     6.51 &      5.53 &       28 &      19 &        9 \\
+    21H (Mu) &      2897 & 18.25 &    10.05 &      8.20 &       31 &      21 &       10 \\
+    21I (Delta) &   86585 & 16.55 &     8.93 &      7.62 &       30 &      24 &        6 \\
+    21J (Delta) &   733825 & 14.30 &     7.17 &      7.14 &       34 &      28 &        6 \\
+    21K (Omicron) & 956696 & 13.48 &     6.45 &      7.03 &       56 &      46 &       10 \\
+    21L (Omicron) & 690177 & 9.24 &     4.39 &      4.85 &       68 &      50 &       18 \\
+    22A (Omicron) & 33451 & 7.71 &     2.81 &      4.90 &       71 &      52 &       19 \\
+    22B (Omicron) & 124239 & 11.90 &     5.03 &      6.87 &       66 &      50 &       16 \\
            \hline
@@
 \caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.}
-    The last three columns give the distances of the clade founder sequence from putative MRCA of SARS-CoV-2.
+The column `\# of seq.' shows the number of sequences that entered the analysis after filtering and restriction to the first 6 months after the emergence of the variant.
+    The last three columns give the distances of the clade founder sequence from putative MRCA of SARS-CoV-2 (19B).
 \label{tab:rates}}
@@
 By analyzing the rate of mutations that spread only on short times scales within fine-grained Pango lineages, I estimated the level of constraint on different parts of the SARS-CoV-genome.
-The great majority of 3rd positions in codons at which  most mutations are synonymous do not show strong signatures of conservations.
+The great majority of 3rd positions in codons -- at which most mutations are synonymous -- do not show strong signatures of conservation.
 The major open reading frames \emph{ORF1ab, S, N, E} and \emph{M} show clear signatures of purifying selection with around 50\% of sites being more constrained than the most constrained 10\% of 3rd positions.
-Purifying selection analyzed there is operating over the time rare lineage circulate, which is typically a few weeks.
+Purifying selection analyzed here is operating over the time during which rare lineage circulate, which is typically a few weeks.
 On shorter time scales, more strongly deleterious mutations will be observed.
-The degree of purifying selection is consistent with a similar analysis in HIV-1 that also show small fitness costs for most synonymous mutations, while half of the non-synonymous mutations are so deleterious that they are not even observed on short timescales \citep{zanini_vivo_2017}.
+On longer time scales, weaker deleterious effects can be quantified.
+The degree of purifying selection is consistent with a similar analysis in HIV-1 that also inferred small fitness costs for most synonymous mutations, while half of the non-synonymous mutations are so deleterious that they are not even observed on short timescales \citep{zanini_vivo_2017}.
 
-The SARS-CoV-2 \emph{ORF3a, ORF6, ORF7a, ORF7b} and \emph{ORF8} show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
+The SARS-CoV-2 genes \emph{ORF3a, ORF6, ORF7a, ORF7b} and \emph{ORF8} show little global signal of constraint and 1st and 2nd positions are as variable as 3rd positions (see Fig.~\ref{fig:fitness_costs}).
 This lack of strong purifying selection in these ORFs does not necessarily imply that they don't matter or are not possible loci of adaptation -- but the exact amino acid sequence does not seem essential for their function.
@@
 These factors make extrapolation -- whether it is reconstruction of past events or scenarios of future evolution -- uncertain.
-Standard phylogenetic or phylodynamic methods will underestimate that uncertainty.
+Standard phylogenetic or phylodynamic methods will underestimate this uncertainty.
 

### 2022-10-30 scripts/root_to_tip.py
put qc stats in more sensible order

@@ -23,11 +23,8 @@ def week_since2020_to_numdate(d):
 
 def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False, max_group=None, QC_threshold=30):
     dropped_seqs = {}
-    # filter for incomplete data
+    # filter for incomplete dates
     d = d.loc[d.date.apply(lambda x:len(x)==10 and 'X' not in x)]
-    pre = len(d)
-    d = d.loc[d.QC_overall_score<QC_threshold]
-    dropped_seqs['QC'] = pre - len(d)
 
     if query:
         pre = len(d)
@@ -46,6 +43,10 @@ def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, 
         pre = len(d)
         d = d.loc[d.numdate<max_date]
         dropped_seqs['max_date'] = pre - len(d)
+
+    pre = len(d)
+    d = d.loc[d.QC_overall_score<QC_threshold]
+    dropped_seqs['QC'] = pre - len(d)
 
     # look for clade defining substitutions
     d["clade_substitutions"] = d.substitutions.apply(lambda x:     [y for y in x.split(',') if y in clade_gt['nuc']] if x else [])

### 2022-11-09 scripts/plot_rtt_counts.py
add script to combine plots

@@ -1,0 +1,343 @@
+import pandas as pd
+import argparse,json
+from treetime.utils import numeric_date, datestring_from_numeric
+from datetime import datetime
+import numpy as np
+from scipy.stats import linregress, scoreatpercentile
+import matplotlib.pyplot as plt
+import matplotlib as mpl
+import seaborn as sns
+from collections import defaultdict
+mpl.rcParams['axes.formatter.useoffset'] = False
+
+def poisson(lam, order):
+    p = np.exp(-lam)
+    try:
+        for i in range(0,int(order)):
+            p *= lam/(i+1)
+    except:
+        p_cum = np.copy(p)
+        for i in range(0,int(order[:-1])-1):
+            p *= lam/(i+1)
+            p_cum += p
+        p = 1 - p_cum
+
+    return np.maximum(0.0, np.minimum(1.0,p))
+
+def fit_poisson(n,k_vecs,t, rate=None):
+    from scipy.optimize import minimize
+    eps = 1e-16
+    def binom(x,n,k_vecs,t, rate):
+        res = 0
+        mu = rate or x[1]**2
+        tu = np.maximum(0,mu*(t-x[0]))
+        for order, k in k_vecs.items():
+            p = poisson(tu, order)
+            res -= np.sum(((n-k)*np.log(1-p+eps) + k*np.log(p+eps)))
+
+        return res
+
+    if rate is None:
+        sol = minimize(binom, (-10,0.25), args=(n,k_vecs,t,rate), method="Powell")
+        res = {'offset':sol['x'][0], 'rate':sol['x'][1]**2}
+    else:
+        sol = minimize(binom, (-10,), args=(n,k_vecs,t,rate), method="Powell")
+        res = {'offset':sol['x'][0], 'rate':rate}
+
+    return res
+
+fs=14
+date_reference = datetime(2020,1,1).toordinal()
+def date_to_week_since2020(d):
+    return (d.toordinal() - date_reference)//7
+
+def week_since2020_to_date(d):
+    return datetime.fromordinal(int(d*7 + date_reference))
+
+def week_since2020_to_numdate(d):
+    return numeric_date(week_since2020_to_date(d))
+
+def filter_and_transform(d, clade_gt, min_date=None, max_date=None, query=None, completeness=None, swap_root=False, max_group=None, QC_threshold=30):
+    dropped_seqs = {}
+    # filter for incomplete dates
[…]
